### 2018-05-23 paper/main.tex
Add initial literature review

\documentclass[a4paper]{article}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{verbatim}
\usepackage{amsfonts}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{listings}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{mathtools}
\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}

\usepackage[round]{natbib}

\makeatletter
\def\BState{\State\hskip-\ALG@thistlm}
\makeatother

\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\sep}{sep}
\lstset{frame=tb,
  language=Python,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{red},
  stringstyle=\color{red},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

\newcommand{\norm}[1]{\lVert #1 \rVert}
\newcommand{\fig}[1]{\includegraphics[width=0.48\textwidth]{#1}}
\newcommand{\figb}[1]{\includegraphics[width=0.95\textwidth]{#1}}
\newcommand{\figz}[1]{\includegraphics[width=0.23\textwidth]{#1.png}}
\newcommand{\m}[1]{\mathbf{#1}}
\newcommand{\EE}{\mathbb{E}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathcal{T}}

\newcommand{\highlight}[1]{{\leavevmode\color{blue} #1}}
\newcommand{\warning}[1]{{\leavevmode\color{red}Warning: #1}}

\usepackage{fancyhdr}
\pagestyle{fancy}

\title{Curriculum and Transfer Learning of Hierarchical Locomotion Policies}
\author{TBD}

\begin{document}

\maketitle

Definitions:
\begin{itemize}
  \item HLC: high-level controller
  \item LLC: low-level contoller
  \item HRL: hierarchical reinforcement learning
\end{itemize}

\section{Literature review}

\subsection*{Locomotion on varied terrain}

\begin{itemize}
  \item \cite{heess2017emergence} trains policies that can, to some extent, handle bumpy ground (see \href{https://www.youtube.com/watch?v=hx_bgoTF7bs&feature=youtu.be&t=1m28s}{this video}).
    However, the agents here do not actually observe the bumps.
    They assume flat ground and depend on policy robustness to handle unexpected terrain variation.
    They do not report quantitative results.
    This approach is likely limited in the kinds of bumps that it can handle.
    In the video, the humanoid trips.

  \item \cite{peng2017deeploco}
    learns locomotion in an environment that includes bumpy ground, but the path they follow is perfectly flat.
    They also use motion capture to train the LLC, and the steps are all of uniform length (0.5 seconds), which we hope to avoid.

  \item \cite{peng2018deepmimic} includes some tasks that involve running across varied terrain.
    They also use a form of curriculum learning, where the policies are first trained on flat ground (and without a vision component), before training then continues on the full problem.
    Their policies are somewhat hierarchical, in the sense that they output a target pose, which is then used as input for a PD controller to generate torques.
    Qualitatively, the policies on these tasks seem to mostly ignore the obstacles, relying on being able to recover from the perturbations generated by collisions.
    In contrast, we would like to avoid the kind of stumbling seen in these videos by introducing another level of hierarchy that generates an explicit footstep plan.

  \item \cite{peng2016terrain}: I haven't read this yet, but it also deals directly with locomotion across irregular terrain (for two-dimensional environments).

\end{itemize}

\subsection*{Hierarchical reinforcement learning}

\begin{itemize}
  \item \cite{heess2016learning} discusses using hierarchical transfer learning to improve exploration in new environments, but this is useful mostly for sparse reward environments.
    In contrast, we are not interested in addressing sparse rewards, but rather in (qualitative) performance on the difficult locomotion task.
    That is, we should feel free to make the rewards as dense and as shaped as we want.
    The reward signal is a tool, not something given to us a priori.

  \item \cite{hausman2018learning} is an interesting theoretical paper showing how to get diverse meta-controls without imposing much structure on the interface between the HLC and the LLC.
    However, in our case, we have a prior idea about a good hierarchical breakdown of the tasks, which we can use to learn a policy more quickly than we otherwise would. \highlight{I had some trouble with this paper; I should probably reread it and update this summary.}

  \item \cite{daume18ilrl} consider using hierarchical policies to improve the query efficiency of interactive imitation learning.
    In contrast, our project does not make use of an expert.
    Even if we do end up using motion capture, such an expert is not hierarchical, and it does not specify actual actions, just target poses.
    However, perhaps for some of the tasks in our curriculum, we will use a hand-coded expert for the HLC in order to pretrain the LLC.

  \item \cite{vezhnevets2017feudal}: \highlight{I have not read this yet, but I plan to.}

  \item \cite{frans2018meta} uses HRL for transfer learning, but they do not apply their results to curriculum learning.
    Hence, the tasks they accomplish remain fairly simple (simple maze traversal, etc.).

  \item \cite{kulkarni2016hierarchical} uses hierarchical RL to encourage exploration in sparse-reward environments.
    They also only consider discrete action spaces; perhaps we don't need to cite this paper at all.
\end{itemize}

\subsection*{Curriculum and transfer learning for locomotion}

\begin{itemize}
  \item The aforementioned paper \cite{heess2017emergence} trains their policies using curriculum learning: the environment gets progressively more difficult as time goes on.
    This achieves great results, but they do not quite succeed on bumpy ground.
    In this project, we hope to explore hierarchical policies as a potential solution to this difficult task.

  \item As mentioned before, \cite{peng2018deepmimic} uses some curriculum learning to accomplish the ``varied terrain'' tasks.

  \item As mentioned before, \cite{frans2018meta} has the explicit goal of using HRL for transfer learning.

  \item Relay Nets \citep{kumar2017relay} are a form of policy switching using the value function to choose between an ensemble of policies trained on different starting distributions, which uses a form of curriculum learning to train policies sequentially.
    In contrast, we wish to train a single policy while varying the environment itself.

\end{itemize}

\subsection*{Locomotion policy architectures}

\begin{itemize}
  \item We might want to use something simple, like a linear controller as proposed by Ben Recht \cite{mania2018simple} to see if that works well.
    It seems to do a good job on OpenAI benchmarks, at least.
  \item Otherwise, we'll probably end up using TRPO or PPO or something (get citation later\dots)
  \item Other things to consider: prioritized experience replay, hindsight experience replay, universal value function approximators.
\end{itemize}

\section{Project summary}

\begin{enumerate}

\item \textbf{One-sentence summary:}
Learn a difficult locomotion task with curriculum learning, hierarchical policies, and possibly model-based planning.

\item \textbf{Task:}
Train a simulated robotic agent to accomplish locomotion in a difficult environment, such as walking (or preferably running) across bumpy ground.

\item \textbf{Introduction:} Robotic locomotion on uneven terrain is an important problem. \highlight{Visak, I would be interested in your high-level reasons for being interested in this problem. I'm not sure I can motivate the problem as well as you can.} Previous approaches have not solved this problem. For instance, in \href{https://www.youtube.com/watch?v=hx_bgoTF7bs&feature=youtu.be&t=1m28s}{this video} the agent stumbles on rubble, and DeepLoco only walks on flat terrain. \highlight{More examples would help, to make sure that this problem is interesting.} We hope to achieve efficient locomotion across bumpy ground using a combination of three techniques.
  \begin{enumerate}
    \item Curriculum learning: Train the agent first on simpler environments, like flat ground and gradual slopes, and gradually increase the difficulty.
    \item Hierarchical controllers: We use a hierarchical architecture where the HLC plans foot placement and the LLC executes the individual steps. We hypothesize that this architecture will more easily handle the discontinuities caused by contact dynamics. We also hypothesize that, with this architecture, it will be easier to transfer knowledge between tasks in the curriculum.
    \item Model-based planning: To help with sample efficiency, the HLC has access to a physics model for the environment. \highlight{I'm not sure how to accomplish this. Physics models for legged robots are probably hard, especially when the ground is not flat. Maybe some kind of \emph{learned} model would work.}
  \end{enumerate}

\item \textbf{Controller architecture:}
  Like DeepLoco, the HLC receives detailed environmental observations, and outputs the target locations for the next couple of foot placements.
  The LLC takes this input to choose the low-level actions that will achieve those foot placements.
  Unlike DeepLoco, the terrain is much more complicated, so it is important that the HLC foot placements are actually feasible.
  Furthermore, we will train the LLC without using motion captures, and we will permit the time per step to vary, rather than requiring each step to take exactly 0.5 seconds.

  The parametrization of the HLC and LLC will probably end up being some kind of neural network, although it might be worth trying simpler policies first (e.g. linear).

  To train these controllers, we could first train the LLC on simpler terrain, where we can specify target foot placement using a simple calculation.
  When training the HLC, to ensure that it chooses feasible foot placements, we could include some additional loss term in the training objective based on whether the LLC actually achieved the foot placement that the HLC suggested.

\highlight{This section could probably use more detail; suggestions are welcome.
For example, do we need to modify this architecture to better handle contact dynamics?
Is there some way we can incorporate a model to help the HLC make better plans?}

\item \textbf{Why this might be a good idea:}
I have not seen previous papers succeed at this task, so if we could show some nice videos of a quadruped or biped running along bumpy ground, it would catch people's attention.
Intuitively, when running along bumpy hiking trails, I am consciously thinking about where to place my feet, then relying on unconsciously being able to place my foot there while I think about the next step.
This suggests a kind of hierarchical policy.

\item \textbf{Why this might not be a good idea:}
  \highlight{Why would *hierarchical* RL be a particularly good way to do transfer learning for this series of tasks?}
  We will need to implement several other baselines to compare against.
  \highlight{What should those baselines be? (What other approaches have people used for locomotion on bumpy ground?)
Is the ``bumpy ground'' task too difficult? (Byron thinks so.)
Will the simulated robot need toes?}
I haven't worked with locomotion before, so I don't have a good sense for this.

\end{enumerate}

\bibliographystyle{plainnat}
\bibliography{bib}

\end{document}

### 2018-05-23 paper/main.tex
Add DeepTerrain summary

@@
 \begin{itemize}
+  \item \cite{peng2016terrain}: Perhaps the most closely related prior work, this paper trains planar characters to run across varied terrain.
+In the discussion section of this paper, they mention three specific directions for extension that essentially exactly define our current project: move to three dimensions, consider more difficult terrains (smaller, more frequent obstacles) that require footstep planning, and use curriculum learning.
+
   \item \cite{heess2017emergence} trains policies that can, to some extent, handle bumpy ground (see \href{https://www.youtube.com/watch?v=hx_bgoTF7bs&feature=youtu.be&t=1m28s}{this video}).
@@
     In contrast, we would like to avoid the kind of stumbling seen in these videos by introducing another level of hierarchy that generates an explicit footstep plan.
-
-  \item \cite{peng2016terrain}: I haven't read this yet, but it also deals directly with locomotion across irregular terrain (for two-dimensional environments).
 

### 2018-05-24 paper/main.tex
Add more related work
Add more references / paper summaries

@@
 
+  \item \cite{peng2017learning} uses bumpy ground to test robustness of learned policies, but again the bumps are not actually perceived and cannot be planned around.
+The bump height thus must remain small relative to the agent.
+\highlight{We should compare this quantitatively, later.} Also, this is 2D.
 \end{itemize}
@@
 \begin{itemize}
+  \item \cite{karpathy2012curriculum} \highlight{TODO: read this}
+
   \item The aforementioned paper \cite{heess2017emergence} trains their policies using curriculum learning: the environment gets progressively more difficult as time goes on.
@@
 \begin{itemize}
+  \item At the lowest level, we will probably be using some kind of stable PD controller, in order to make the reinforcement learning problem as easy as possible.
+I need to read these papers: \cite{tan2011stable}, \cite{shrivastava2013stable}
+  \item \cite{peng2015terrain} introduces a very interesting way of parametrizing the action space in terms of reference points for PD controllers, as well as ``virtual forces''.
+We will want to do something similar.
   \item We might want to use something simple, like a linear controller as proposed by Ben Recht \cite{mania2018simple} to see if that works well.
@@
 \end{itemize}
+
+\subsection*{Miscellaneous}
+
+Other things I should read / potential baselines to compare against.
+\cite{de2010feature} \cite{mordatch2010robust} \cite{mordatch2012discovery}
 

### 2018-05-29 paper/main.tex
Complete literature review (mostly)

@@
 \highlight{We should compare this quantitatively, later.} Also, this is 2D.
+  \item \cite{manchester2011stable} This paper stands out in that they derive a provably stable controller that they test on a physical robot. \highlight{I haven't read it in detail yet, though.}
 \end{itemize}
@@
   \item \cite{hausman2018learning} is an interesting theoretical paper showing how to get diverse meta-controls without imposing much structure on the interface between the HLC and the LLC.
-    However, in our case, we have a prior idea about a good hierarchical breakdown of the tasks, which we can use to learn a policy more quickly than we otherwise would.
+    However, in our case, we have a prior idea about a good hierarchical breakdown of the tasks (namely, a footstep plan).
 \highlight{I had some trouble with this paper; I should probably reread it and update this summary.}
@@
 
-  \item \cite{vezhnevets2017feudal}: \highlight{I have not read this yet, but I plan to.}
+  \item \cite{vezhnevets2017feudal} \emph{learns} diverse low-level policies using a latent state space and directional goals.
+However, we would like a more explicit hierarchical interface in the form of a footstep plan, and our low-level action space is continuous rather than discrete.
 
@@
     They also only consider discrete action spaces; perhaps we don't need to cite this paper at all.
+  \item \cite{nachum2018data}: This paper tries to do off-policy hierarchical RL by changing the actions from the transitions stored in the high-level replay buffer.
+This seems like a hacky way to do things, and their experiments are clearly gaming the system to make their method look good.
+I don't think we should cite this.
 \end{itemize}
@@
 \begin{itemize}
-  \item \cite{karpathy2012curriculum} \highlight{TODO: read this}
+  \item \cite{karpathy2012curriculum} use a two-level hand-designed curriculum to achieve impressive acrobatic results for a low-dimenisonal hopper.
+However, their method involves random search and nonparametric nearest-neighbor policies, which are unlikely to generalize to higher-dimensional morphologies.
 
@@
   \item At the lowest level, we will probably be using some kind of stable PD controller, in order to make the reinforcement learning problem as easy as possible.
-I need to read these papers: \cite{tan2011stable}, \cite{shrivastava2013stable}
+\highlight{I need to read these papers:} \cite{tan2011stable}, \cite{shrivastava2013stable}
   \item \cite{peng2015terrain} introduces a very interesting way of parametrizing the action space in terms of reference points for PD controllers, as well as ``virtual forces''.
 We will want to do something similar.
-  \item We might want to use something simple, like a linear controller as proposed by Ben Recht \cite{mania2018simple} to see if that works well.
+  \item We might want to use something simple, like a linear controller as proposed by \cite{mania2018simple} to see if that works well.
     It seems to do a good job on OpenAI benchmarks, at least.
@@
 
-\subsection*{Miscellaneous}
+\subsection*{Model-based locomotion}
 
-Other things I should read / potential baselines to compare against.
-\cite{de2010feature} \cite{mordatch2010robust} \cite{mordatch2012discovery}
+\begin{itemize}
+  \item \cite{de2010feature} introduce a method for specifying locomotion behaviors with high-level features.
+From these features, low-level torques can be derived automatically by solving a series of quadratic programs based on physical models.
+This method is somewhat slow (50\% real time), and we might be able to avoid using hand-engineered models by employing deep networks.
+
+  \item \cite{mordatch2010robust} use \cite{de2010feature} to execute a footstep plan which is generated using a lower-dimensional simplified physics model (a spring-loaded inverted pendulum).
+We might be able to use such an optimization- / model-based method as an expert, if we want to employ some kind of imitation learning.
+This could be similar to the approach of AlphaGo Zero.
+
+  \item \cite{mordatch2012discovery} handles contacts in a clever way, using piecewise-constant, smooth contact variables and a continuation scheme.
+This allows complex motions to emerge from a quadratic optimization procedure.
+However, this method is very slow (2\% real time) and it generates only static motion (no jumping or slipping).
+Again, maybe we could use it as an expert.
+
+\end{itemize}
 

### 2018-05-31 paper/main.tex
Write introduction, aspirational experiments introduction, and some ideas for policy architecture

@@
 
-\newcommand{\highlight}[1]{{\leavevmode\color{blue} #1}}
-\newcommand{\warning}[1]{{\leavevmode\color{red}Warning: #1}}
+\newcommand{\nhatch}[1]{{\leavevmode\color{blue} Nathan: #1}}
 
@@
   \item HRL: hierarchical reinforcement learning
+\end{itemize}
+
+\section{Introduction}
+
+Locomotion is hard because the changing contacts make the dynamics difficult to model.
+Recent deep reinforcement learning approaches try to learn this model implicitly via the policy network, which has trouble learning such complicated models, possibly due to the limitations of current optimization methods (like policy gradient).
+Such approaches fail in environments that require predictive planning at the level of individual contacts, such as running across irregular terrain, where small miscalculations have disproportionally large effects.
+
+We aim to simplify the learning problem by imposing prior structure on the policy: we split it into a two-level hierarchy.
+The higher level breaks the plan into smaller chunks, which are easier to achieve.
+The lower level must then execute these smaller chunks.
+Each chunk is a single footstep.
+This choice simplifies the dynamics that the lower level needs to model, because each step involves no (or very few) changes of contact.
+It also simplifies the upper level learning problem, since the controller does not need to handle details of torque, other than knowing which step plans can actually be achieved.
+
+\section{Experiments}
+
+Our experiments are designed to test two hypotheses.
+
+Our primary hypothesis is that this hierarchical architecture can navigate a variety of complicated locomotion environments, learning faster and achieving better results than other reinforcement learning methods.
+These locomotion environments include flat ground, inclines, stairs, random inclines, stepping stones, hurdles, and combinations of the above.
+
+Our two metrics for ``learning faster'' are (1) number of simulation frames of experience collected, and (2) total computation time.
+We consider both necessary, because the relative expense of more experience versus more computation may vary across application domains.
+
+We collect both quantitative and qualitative metrics to measure ``achieving better results''.
+Quantitatively, we compare the average rate of forward progress.
+We also consider the robustness of the model to force perturbations and various measures of the ``difficulty'' of the terrain (such as stair height).
+Qualitatively, we document motion quality by means of supplementary videos, which we encourage the reader to consult.
+
+Our secondary hypothesis is that the learned policy can be run in real-time while still achieving results comparable to more expensive, model-based optimization methods.
+
+This hypothesis is difficult to test, because model-based optimization methods are difficult to implement for direct comparison.
+Thus, our metric is simply the percent real time necessary to compute the control for the next time step, which we compare to numbers reported in optimization-based papers.
+
+Additionally, we conduct ablative analysis of the importance of various components of our approach.
+\nhatch{In particular, we want to consider various different HLC / LLC interfaces that we could have used.
+For example: should we include a target foot height that the LLC should reach?
+(I think this would help when going over hurdles.)
+For another example (assuming we use some kind of PD control at the lowest level): how much does performance degrade if we don't use PD control?
+(That is, if the LLC outputs raw torques.)}
+
+\section{Policy architecture}
+
+\nhatch{Ideally, maybe we should design this so that walking and running can both be achieved in the same episode.}
+
+\nhatch{Should there be some ensemble of low-level policy networks that the HLC can choose between?
+For example, we could take the policy with the highest estimated value given the target step location.
+Having some discrete choices like this might be easier than using solely continuous parameters.}
+
+\textbf{High-level observations.}
+We assume that perception is solved.
+The high-level controller receives proprioceptive information and environmental information.
+The environmental information consists of a list of potential surfaces to step on, specified in terms of relative displacement from the agent and surface normal.
+\nhatch{Maybe, for polygonal surfaces, we could also report the surface ``size'' (e.g. largest-radius circle that fits inside the polygon).}
+
+\textbf{High-level actions.}
+High-level actions are chosen on each heel-down event (once per footstep).
+At these events, the HLC outputs a target for the next footstep that it expects the LLC can achieve.
+This target consists of several parameters: \nhatch{TODO think about this some more}
+\nhatch{It might be possible to incorporate some optimization / search / planning into the HLC, because its actions happen infrequently.
+But this would require some objective to be optimized, which might involve implementing a model like a spring-loaded inverted pendulum (SLIP), or inventing some heuristic objective that might characterize good plans.
+We might want to implement a model like SLIP anyway, in order to use it as an expert to train a neural network for the HLC (similar to AlphaGo Zero and ExIt).
+In any case, only the first step of the plan would be revealed to the LLC.}
+
+\textbf{Low-level observations.}
+The low-level controller receives proprioceptive information and the target location for the next footstep.
+Unlike the HLC, it does not observe the external environment.
+
+\textbf{Low-level actions.}
+On the one hand, we could have low-level actions being the raw torques.
+On the other hand, this might make the learning problem harder, so we could specify some more abstract actions and use PD controllers or other analytical formulas to derive the actual torques.
+For instance, the output could be a target pose or end-effector position which is chosen more infrequently than the simulation framerate.
+Then we could solve an inverse kinematics or inverse dynamics problem to calculate a target pose, which is then achieved using PD controllers.
+\nhatch{We probably don't need all of this machinery.
+For the initial proof of concept, it might make sense to see how we can do using just raw torques.}
+\nhatch{TODO think about this some more.}
+
+\textbf{PD control.}
+\nhatch{We might have to rethink this abstraction layer once the ground includes inclines.
+For instance, the PD controller might now have to handle different target ankle angles.}
+
+\section{Timeline of work}
+
+The goal is to submit this to ICRA on September 15.
+
+\begin{itemize}
+  \item Early June (2 weeks): Proof-of-concept: train a 2D walker low-level controller to follow an irregular step plan on flat ground, including some low-level PD control if necessary.
+  \item Late June (2 weeks): Train low-level controller on stepping stones.
+    Train high-level controller on flat ground, then on stepping stones.
+    Compare performance on stepping stones to baseline deep RL approach (TRPO).
+    If we can't get this to work (better than TRPO), we should consider abandoning the project.
+  \item July (4 weeks): Implement and train on inclines, stairs, and hurdles.
+    Perform ablative analyses.
+  \item Early August (2 weeks): I (Nathan) will be on vacation.
+  \item Late August and early September (4 weeks): Last-minute experiments.
+    Write and revise the paper.
+  \item September 15: ICRA deadline.
 \end{itemize}
@@
     The bump height thus must remain small relative to the agent.
-\highlight{We should compare this quantitatively, later.} Also, this is 2D.
-  \item \cite{manchester2011stable} This paper stands out in that they derive a provably stable controller that they test on a physical robot. \highlight{I haven't read it in detail yet, though.}
+    \nhatch{We should compare this quantitatively, later.} Also, this is 2D.
+  \item \cite{manchester2011stable} This paper stands out in that they derive a provably stable controller that they test on a physical robot.
+    \nhatch{I haven't read it in detail yet, though.}
 \end{itemize}
@@
     However, in our case, we have a prior idea about a good hierarchical breakdown of the tasks (namely, a footstep plan).
-\highlight{I had some trouble with this paper; I should probably reread it and update this summary.}
+    \nhatch{I had some trouble with this paper; I should probably reread it and update this summary.}
 
@@
   \item At the lowest level, we will probably be using some kind of stable PD controller, in order to make the reinforcement learning problem as easy as possible.
-\highlight{I need to read these papers:} \cite{tan2011stable}, \cite{shrivastava2013stable}
+    \nhatch{I need to read these papers:} \cite{tan2011stable}, \cite{shrivastava2013stable}
   \item \cite{peng2015terrain} introduces a very interesting way of parametrizing the action space in terms of reference points for PD controllers, as well as ``virtual forces''.
@@
 \item \textbf{Introduction:} Robotic locomotion on uneven terrain is an important problem.
-\highlight{Visak, I would be interested in your high-level reasons for being interested in this problem.
+  \nhatch{Visak, I would be interested in your high-level reasons for being interested in this problem.
   I'm not sure I can motivate the problem as well as you can.} Previous approaches have not solved this problem.
   For instance, in \href{https://www.youtube.com/watch?v=hx_bgoTF7bs&feature=youtu.be&t=1m28s}{this video} the agent stumbles on rubble, and DeepLoco only walks on flat terrain.
-\highlight{More examples would help, to make sure that this problem is interesting.} We hope to achieve efficient locomotion across bumpy ground using a combination of three techniques.
+  \nhatch{More examples would help, to make sure that this problem is interesting.} We hope to achieve efficient locomotion across bumpy ground using a combination of three techniques.
   \begin{enumerate}
@@
     \item Model-based planning: To help with sample efficiency, the HLC has access to a physics model for the environment.
-\highlight{I'm not sure how to accomplish this.
+      \nhatch{I'm not sure how to accomplish this.
       Physics models for legged robots are probably hard, especially when the ground is not flat.
@@
 
-\highlight{This section could probably use more detail; suggestions are welcome.
+\nhatch{This section could probably use more detail; suggestions are welcome.
 For example, do we need to modify this architecture to better handle contact dynamics?
@@
 \item \textbf{Why this might not be a good idea:}
-  \highlight{Why would *hierarchical* RL be a particularly good way to do transfer learning for this series of tasks?}
+  \nhatch{Why would *hierarchical* RL be a particularly good way to do transfer learning for this series of tasks?}
   We will need to implement several other baselines to compare against.
-  \highlight{What should those baselines be? (What other approaches have people used for locomotion on bumpy ground?)
+  \nhatch{What should those baselines be? (What other approaches have people used for locomotion on bumpy ground?)
 Is the ``bumpy ground'' task too difficult? (Byron thinks so.)

### 2018-05-31 source/pd_stand.py
Implement a rough PD controller for DartHopper

@@ -1,0 +1,62 @@
+import gym
+import time
+import numpy as np
+import sys
+from IPython import embed
+
+class PDStand:
+    def __init__(self):
+        self.env = gym.make('DartHopper-v1')
+        self.n_act = self.env.action_space.shape[0]
+
+    def flop(self):
+        self.env.reset()
+        while True:
+            control = np.zeros(self.n_act)
+            observation, reward, done, info = self.env.step(control)
+            self.env.render()
+            if done:
+                return
+
+    def simulate(self, controller, steps):
+        for _ in range(steps):
+            self.obs, _, _, _ = self.env.step(controller)
+            self.env.render()
+        return self.obs
+
+    def stand(self):
+        obs = self.env.reset()
+        pose = obs[2:5] # hip, knee, and ankle angles (only DOFs we control)
+        target_pose = np.array([-0.05,-0.08,0.05])*np.pi # slightly bent joints for stability?
+                                                         # and to avoid hitting joint limits?
+        prev_pose = pose
+        prev_d = np.zeros(self.n_act)
+        for i in range(1000):
+            if i % 200 == 100:
+                q = self.env.env.robot_skeleton.q.copy()
+                dq = self.env.env.robot_skeleton.dq.copy()
+                # Edit these as desired, to introduce perturbations
+                dq[0] += i/200
+                dq[1] += i/200
+                self.env.env.set_state(q, dq)
+            control, prev_d = self.pd_control(pose, target_pose, prev_pose, prev_d)
+            obs, _, done, _ = self.env.step(control)
+            prev_pose, pose = pose, obs[2:5]
+            self.env.render()
+            if done:
+                pass
+                #return
+
+    def pd_control(self, p, target_p, prev_p, prev_d):
+        # Tuned these constants via guess and check.
+        kp = np.array([1,1,1])
+        kd = np.sqrt(2*kp) + [0,0,3]
+        d_damper = 0.8
+        diff = p - prev_p
+        d = d_damper*prev_d + (1-d_damper)*diff
+        control = -kp * (p - target_p) - kd * d
+        return control, d
+
+if __name__ == "__main__":
+    #PDStand().flop()
+    PDStand().stand()

### 2018-06-04 paper/main.tex
Add initial 2d walker script
2dwalker.py contains code to run an two-step episode for a planar
walker. It also has code for a PD controller that can hold a target
pose.

@@
 We also consider the robustness of the model to force perturbations and various measures of the ``difficulty'' of the terrain (such as stair height).
+\footnote{Another interesting question would be robustness with respect to accuracy of perception.
+If the potential step target locations / surface normals are perturbed, what happens?}
 Qualitatively, we document motion quality by means of supplementary videos, which we encourage the reader to consult.
@@
 On the one hand, we could have low-level actions being the raw torques.
-On the other hand, this might make the learning problem harder, so we could specify some more abstract actions and use PD controllers or other analytical formulas to derive the actual torques.
+On the other hand, this might make the learning problem harder.
+Instead, we could specify ``abstract actions'' and use PD controllers or other analytical formulas to derive the actual torques.
 For instance, the output could be a target pose or end-effector position which is chosen more infrequently than the simulation framerate.
@@
 For instance, the PD controller might now have to handle different target ankle angles.}
+
+\textbf{Miscellaneous ideas.}
+\begin{itemize}
+  \item Could the HLC-LLC interface be in terms of a \emph{force} that the LLC should try to generate on the next foot contact?
+  \item How can we design the LLC to generate good motion quality (e.g. low energy, low angular momentum)?
+\end{itemize}
 

### 2018-06-04 source/2dwalker.py

@@ -1,0 +1,117 @@
+import pydart2 as pydart
+from IPython import embed
+from enum import Enum
+import numpy as np
+
+EPISODE_TIME_LIMIT = 5.0 # seconds
+
+class Controller:
+    def __init__(self, skel):
+        self.skel = skel
+        self.target = None
+        self.inactive = True
+        brickdof = 3 # We're in 2D
+        self.Kp = np.array([0.0] * brickdof + [400.0] * (self.skel.ndofs - brickdof))
+        self.Kd = np.array([0.0] * brickdof + [40.0] * (self.skel.ndofs - brickdof))
+
+    def compute(self):
+        if self.inactive:
+            return np.zeros_like(self.Kp)
+        return -self.Kp * (self.skel.q - self.target) - self.Kd * self.skel.dq
+
+class State(Enum):
+    INIT = 0
+    RIGHT_PLANT = 1
+    RIGHT_SWING = 2
+    LEFT_PLANT = 3
+    LEFT_SWING = 4
+
+class Episode:
+    def __init__(self, world, low_level_controller):
+        self.world = world
+        self.target = 0.5
+        self.score = 0.0
+        walker = world.skeletons[1]
+        self.r_foot = walker.bodynodes[5]
+        self.l_foot = walker.bodynodes[8]
+
+        controller = Controller(walker)
+        target = walker.positions()
+        target[3] = 1
+        walker.set_positions(target)
+        controller.target = target
+        controller.inactive = False
+
+        walker.set_controller(controller)
+        self.state = State.INIT
+        #pydart.gui.viewer.launch(world)
+
+    def find_contact(self, bodynode):
+        # TODO: It is possible for a body node to have multiple contacts.
+        # Which one should be returned?
+        for c in self.world.collision_result.contacts:
+            if c.bodynode2 == bodynode:
+                return c
+
+    def calc_score(self, contact):
+        # p[0] is the X position of the contact
+        return np.exp(-(contact.p[0] - self.target)**2)
+
+    def print_contact(self, contact):
+        print(
+            "{:.3f}, {:.3f}, {}: {} made contact at {:.3f} (target: {:.3f})".format(
+            self.world.time(), self.score, self.state,
+            contact.bodynode2, contact.p[0], self.target))
+
+    def run(self):
+        while True:
+            if self.world.time() > EPISODE_TIME_LIMIT:
+                print("Time limit reached")
+                return self.score
+
+            l_foot_contact = self.find_contact(self.l_foot)
[…]

### 2018-06-04 source/2dwalker.py
Add more comments

@@ -6,10 +6,11 @@ import numpy as np
 EPISODE_TIME_LIMIT = 5.0 # seconds
 
 class Controller:
-    def __init__(self, skel):
+    def __init__(self, skel, low_level_controller):
         self.skel = skel
         self.target = None
-        self.inactive = True
+        self.inactive = False
+        self.low_level_controller = low_level_controller
         brickdof = 3 # We're in 2D
         self.Kp = np.array([0.0] * brickdof + [400.0] * (self.skel.ndofs - brickdof))
         self.Kd = np.array([0.0] * brickdof + [40.0] * (self.skel.ndofs - brickdof))
@@ -35,17 +36,21 @@ class Episode:
         self.r_foot = walker.bodynodes[5]
         self.l_foot = walker.bodynodes[8]
 
-        controller = Controller(walker)
-        target = walker.positions()
-        target[3] = 1
-        walker.set_positions(target)
-        controller.target = target
-        controller.inactive = False
+        controller = Controller(walker, low_level_controller)
+        pos = walker.positions()
+        pos[3] = 1
+        # TODO: introduce some randomness to the starting conditions
+        walker.set_positions(pos)
+        # For now, just set a static PD controller target pose
+        # TODO: Choose this target pose from some other policy
+        controller.target = pos
 
         walker.set_controller(controller)
         self.state = State.INIT
-        #pydart.gui.viewer.launch(world)
+        pydart.gui.viewer.launch(world)
 
+    # We locate heeldown events for a given foot based on the first contact
+    # for that foot after it has been in the air for at least one frame.
     def find_contact(self, bodynode):
         # TODO: It is possible for a body node to have multiple contacts.
         # Which one should be returned?
@@ -53,6 +58,8 @@ class Episode:
             if c.bodynode2 == bodynode:
                 return c
 
+    # On each heeldown event, we get points based on how close we were to
+    # the target step location. Maximum 1 point per step.
     def calc_score(self, contact):
         # p[0] is the X position of the contact
         return np.exp(-(contact.p[0] - self.target)**2)
@@ -63,6 +70,8 @@ class Episode:
             self.world.time(), self.score, self.state,
             contact.bodynode2, contact.p[0], self.target))
 
+    # The episode terminates after one right footstep and one left footstep,
+    # or if the time limit is reached.
     def run(self):
         while True:
             if self.world.time() > EPISODE_TIME_LIMIT:

### 2018-06-04 source/2dwalker.py
Reformat as a reinforcement learning environment
Restructures 2dwalker.py to conform to the OpenAI gym interface.
Hooks this up to some code to learn a linear controller by random
search.

The learning code doesn't work yet and is really buggy, sorry.

@@ -2,18 +2,25 @@ import pydart2 as pydart
 from IPython import embed
 from enum import Enum
 import numpy as np
+import random_search
 
+# For rendering
+from gym.envs.dart.static_window import *
+from pydart2.gui.trackball import Trackball
+
+SIMULATION_RATE = 1.0 / 2000.0 # seconds
 EPISODE_TIME_LIMIT = 5.0 # seconds
+LLC_QUERY_PERIOD = 1.0 / 30.0 # seconds
+STEPS_PER_QUERY = int(LLC_QUERY_PERIOD / SIMULATION_RATE)
+BRICK_DOF = 3 # We're in 2D
 
 class Controller:
-    def __init__(self, skel, low_level_controller):
+    def __init__(self, skel):
         self.skel = skel
         self.target = None
         self.inactive = False
-        self.low_level_controller = low_level_controller
-        brickdof = 3 # We're in 2D
-        self.Kp = np.array([0.0] * brickdof + [400.0] * (self.skel.ndofs - brickdof))
-        self.Kd = np.array([0.0] * brickdof + [40.0] * (self.skel.ndofs - brickdof))
+        self.Kp = np.array([0.0] * BRICK_DOF + [400.0] * (self.skel.ndofs - BRICK_DOF))
+        self.Kd = np.array([0.0] * BRICK_DOF + [40.0] * (self.skel.ndofs - BRICK_DOF))
 
     def compute(self):
         if self.inactive:
@@ -27,27 +34,45 @@ class State(Enum):
     LEFT_PLANT = 3
     LEFT_SWING = 4
 
-class Episode:
-    def __init__(self, world, low_level_controller):
+class TwoStepEnv:
+    def __init__(self, world):
         self.world = world
-        self.target = 0.5
-        self.score = 0.0
         walker = world.skeletons[1]
+        self.agent = walker
         self.r_foot = walker.bodynodes[5]
         self.l_foot = walker.bodynodes[8]
 
-        controller = Controller(walker, low_level_controller)
-        pos = walker.positions()
-        pos[3] = 1
-        # TODO: introduce some randomness to the starting conditions
-        walker.set_positions(pos)
-        # For now, just set a static PD controller target pose
-        # TODO: Choose this target pose from some other policy
-        controller.target = pos
+        # We just want this to be something that has a "shape" method
+        # TODO incorporate adding target step locations into the observations
+        self.observation_space = np.zeros(18)
+        self.action_space = np.zeros(6)
 
[…]

### 2018-06-04 source/random_search.py

@@ -1,0 +1,95 @@
+import numpy as np
+from IPython import embed
+
+SHIFT = 0.0
+
+class Whitener:
+    def __init__(self, env):
+        self.N = 0
+        self.env = env
+        obs_dim = env.observation_space.shape[0]
+        self.mean = np.zeros(obs_dim)
+        self.var = np.ones(obs_dim)
+
+    def update_stats(self, obs):
+        mean = (self.N * self.mean + obs) / (self.N+1)
+        var = (self.N * self.var + self.N * (self.mean - mean)**2 + (obs - mean)**2) / (self.N+1)
+        self.N = self.N+1
+        self.mean = mean
+        self.var = var
+
+    def whiten_observation(self, obs, stats):
+        if stats:
+            self.update_stats(obs)
+        var_to_use = 1 if self.N < 2 else self.var
+        return (obs - self.mean) / var_to_use**(0.5)
+
+    def run_trajectory(self, controller, seed, visual=True, stats=True, shift=SHIFT):
+        self.env.seed(seed)
+        observation = self.env.reset()
+        obs = self.whiten_observation(observation, stats)
+        rewards = []
+        while True:
+            observation, reward, done, info = self.env.step(controller(obs))
+            obs = self.whiten_observation(observation, stats)
+            if visual:
+                self.env.render()
+            rewards.append(reward - shift) # Subtract the "alive_bonus"
+            if done:
+                return sum(rewards)
+
+    def rt(self, controller_matrix, seed=None, visual=False, stats=True, shift=SHIFT):
+        c = lambda obs: np.dot(controller_matrix, obs)
+        return self.run_trajectory(c, seed, visual, stats, shift)
+
+class RandomSearch:
+    def __init__(self, env, n_dirs, step_size, eps):
+        obs_dim = env.observation_space.shape[0]
+        action_dim = env.action_space.shape[0]
+        self.w_policy = np.zeros((action_dim, obs_dim))
+
+        self.whitener = Whitener(env)
+        self.n_dirs = n_dirs
+        self.eps = eps
+        self.step_size = step_size
+        self.episodes = 0
+
+    def sample_perturbation(self):
+        delta = np.random.randn(np.prod(self.w_policy.shape))
+        delta /= np.linalg.norm(delta)
+        return delta.reshape(self.w_policy.shape)
+
+    def estimate_grad(self, policy):
+        grad = np.zeros_like(policy)
+        rets = []
[…]

### 2018-06-12 paper/main.tex
Add Little Dog DARPA challenge to literature review

@@
 
+  \item \cite{zucker2010optimization} is a project from the DARPA Learning Locomotion challenge that uses optimization to generate hierarchical motion plans and recover from errors in real time.
+    It has been tested on a physical robot with \href{https://www.youtube.com/watch?v=KpqFebvRLeQ}{impressive results}.
+    We will need to compare our work carefully with this project in the final paper, and we will need to improve upon their results one way or another.
+    The biggest difference might be in robot morphology / stability: executing step plans on bipeds is perhaps more difficult than executing step plans on quadrupeds given the small base of support during each step.
+    Other avenues for improvement might be eliminating the need for 45 seconds of initial planning, traveling faster (running?), or using some kind of interesting new technique (e.g.
+reinforcement learning or deep learning).
+    It is also worth considering whether our algorithm could work on a physical robot, though for the initial work we will unfortunately be working purely in simulation.
+
   \item \cite{heess2017emergence} trains policies that can, to some extent, handle bumpy ground (see \href{https://www.youtube.com/watch?v=hx_bgoTF7bs&feature=youtu.be&t=1m28s}{this video}).
@@
     They do not report quantitative results.
-    This approach is likely limited in the kinds of bumps that it can handle.
+    This approach is likely limited in the kinds of terrain that it can handle.
     In the video, the humanoid trips.
@@
     \nhatch{We should compare this quantitatively, later.} Also, this is 2D.
+
   \item \cite{manchester2011stable} This paper stands out in that they derive a provably stable controller that they test on a physical robot.
     \nhatch{I haven't read it in detail yet, though.}
+
 \end{itemize}

### 2018-06-12 source/stepper.py

@@ -126,5 +126,5 @@ register(
         id='Stepper-v0',
         entry_point='stepper:DartStepperEnv',
         max_episode_steps=150,
-        reward_threshold=1.99
+        reward_threshold=1.99 # I don't actually know what this number does
         )

### 2018-06-28 paper/main.tex
Add ICLR distillation paper to lit review

@@
 \begin{itemize}
+  \item \cite{2018-ICLR-distill} This paper uses a similar curriculum-learning approach to 2D-biped locomotion on varied terrains, starting with flat ground and working up to a mixture of slopes, steps, and gaps.
+We may want to use their distillation approach when integrating multiple skills.
+(Alternatively, we could use what they call ``pure transfer learning'', but have each new task incorporate the relevant challenges of previous tasks.) They use motion capture, and the actions are target joint angles.
+
   \item \cite{karpathy2012curriculum} use a two-level hand-designed curriculum to achieve impressive acrobatic results for a low-dimenisonal hopper.

### 2018-06-28 source/experiment.py
Graph the learning curve over multiple iterations

@@ -1,0 +1,46 @@
+from IPython import embed
+import numpy as np
+import pickle
+import os
+import matplotlib.pyplot as plt
+from inverse_dynamics import LearnInverseDynamics
+
+
+class Experiment:
+    def __init__(self, learner):
+        self.learner = learner
+
+    def repeat_eval(self, n_eval_repeats):
+        avg_losses = []
+        max_errors = []
+        for k in range(n_eval_repeats):
+            print("Starting evaluation", k)
+            avg_loss, max_error = self.learner.evaluate()
+            avg_losses.append(avg_loss)
+            max_errors.append(max_error)
+        return np.mean(avg_losses), np.std(avg_losses), np.mean(max_errors), np.std(max_errors)
+
+    def multi_evaluate(self, n_iters=3, n_eval_repeats=8):
+        results = np.zeros((n_iters+1, 4))
+        results[0] = self.repeat_eval(n_eval_repeats)
+        for i in range(n_iters):
+            self.learner.training_iter()
+            results[i+1] = self.repeat_eval(n_eval_repeats)
+        self.plot_results(results)
+
+    def plot_results(self, results):
+        x = list(range(results.shape[0]))
+        avg_loss_line, = plt.plot(x, results[:,0], color="red")
+        max_error_line, = plt.plot(x, results[:,2], color="blue")
+        plt.fill_between(x, results[:,0]-results[:,1], results[:,0]+results[:,1], color="red", alpha=0.2)
+        plt.fill_between(x, results[:,2]-results[:,3], results[:,2]+results[:,3], color="blue", alpha=0.2)
+        plt.savefig('{}.png'.format("thingy"))
+
+if __name__ == '__main__':
+    from walker import TwoStepEnv
+    from simbicon import Simbicon
+    env = TwoStepEnv(Simbicon)
+    learn = LearnInverseDynamics(env)
+    ex = Experiment(learn)
+    ex.multi_evaluate()
+    embed()

### 2018-06-28 source/inverse_dynamics.py

@@ -15,7 +15,7 @@ TRAIN_FILENAME = 'data/train.pkl'
 
 RIDGE_ALPHA = 10.0
 
-class LearnInvDynamics:
+class LearnInverseDynamics:
     def __init__(self, env):
         self.env = env
         self.initialize_start_states()
@@ -69,7 +69,7 @@ class LearnInvDynamics:
     def generate_target(self, prev_target=None):
         # Choose some random targets that are hopefully representative of what
         # we will see during testing episodes.
-        if prev_target:
+        if prev_target is not None:
             current_v = prev_target[1]
         else:
             current_v = 1.0
@@ -93,7 +93,7 @@ class LearnInvDynamics:
         self.collect_dataset()
         self.train_inverse_dynamics()
         if demo:
-            self.demo_start_state(0, render=1)
+            self.evaluate(render=1.0)
 
     def train_inverse_dynamics(self):
         # TODO find some library to do gradient descent
@@ -111,27 +111,25 @@ class LearnInvDynamics:
         if state is None:
             # The robot crashed or something. Just hack so the script doesn't *also* crash.
             state = start_state
-            step_dist = -1000
+            step_dist = -1 # Should make this bigger, but then the graph gets way out of scale
         loss = (step_dist - target[0])**2 + (state[9] - target[1])**2
         error = np.abs(step_dist - target[0])
-        print("sq. loss: {:.3f}, error: {:.3f}".format(loss, error))
         return state, loss, error
 
-    def demo_start_state(self, i, n_steps=8, render=1.0, record_video=False):
-        state = self.start_states[i]
-        self.env.reset(state, record_video=record_video)
-        total = 0
+    def evaluate(self, n_steps=8, render=None, record_video=False):
+        state = self.env.reset(record_video=record_video)
+        total_loss = 0
         max_error = 0
         t = self.generate_target()
         for _ in range(n_steps):
             state, loss, error = self.run_step(state, render, t)
             if error > max_error:
                 max_error = error
-            total += loss
+            total_loss += loss
             t = self.generate_target(t)
-        print("Avg. loss: {:.3f}".format(total/n_steps))
-        print("Max error: {:.3f}".format(max_error))
+        avg_loss = total_loss/n_steps
         self.env.reset() # This ensures the video recorder is closed properly.
+        return avg_loss, max_error
 
[…]

### 2018-06-28 source/simbicon.py

@@ -93,14 +93,16 @@ class Simbicon(PDController):
     def change_state(self):
         swing = "RIGHT" if self.stance_idx == 6 else "LEFT"
         suffix = " at {:.3f}".format(self.contact_x) if self.direction == DOWN else ""
-        print("{:.3f}: Ended state {} {}{}".format(self.time(), swing, self.direction, suffix))
+        result = None
         if self.direction == DOWN:
             self.swing_idx, self.stance_idx = self.stance_idx, self.swing_idx
             self.direction = UP
+            result = "{:.3f}: Ended state {} {}{}".format(self.time(), swing, self.direction, suffix)
         else:
             # TODO skip this state if the swing foot is already in contact?
             self.direction = DOWN
         self.state_started = self.time()
+        return result
 
     def compute(self):
         # TODO the walking controller can start from rest,

### 2018-06-28 source/walker.py

@@ -18,6 +18,11 @@ REAL_TIME_STEPS_PER_RENDER = 25 # Number of simulation steps to run per frame so
 
 FRICTION_COEFF = 100.0
 
+class StepResult(Enum):
+    ERROR = -1
+    IN_PROGRESS = 0
+    COMPLETE = 1
+
 class TwoStepEnv:
     def __init__(self, controller_class):
         world = load_world()
@@ -87,26 +92,27 @@ class TwoStepEnv:
             if not (b.name in allowed_contact):
                 return True
 
-    # The episode terminates after one right footstep and one left footstep,
-    # or if the time limit is reached.
+    # Executes one world step.
+    # Returns (status code, step distance, human-readable message) tuple.
     def simulation_step(self):
         if self.world.time() > EPISODE_TIME_LIMIT:
-            return "Time limit reached"
+            return StepResult.ERROR, None, "Time limit reached"
         obs = self.current_observation()
         if not np.isfinite(obs).all():
-            return "Numerical explosion"
+            return StepResult.ERROR, None, "Numerical explosion"
         if self.crashed():
-            return "Crashed"
+            return StepResult.ERROR, None, "Crashed"
 
         l_contact = self.find_contacts(self.l_foot)
         r_contact = self.find_contacts(self.r_foot)
         state_complete, step_dist = self.controller.state_complete(l_contact, r_contact)
         if state_complete:
-            self.controller.change_state()
+            status_string = self.controller.change_state()
             if step_dist:
-                return step_dist
+                return StepResult.COMPLETE, step_dist, status_string
 
         self.world.step()
+        return StepResult.IN_PROGRESS, None, None
 
     def current_observation(self):
         obs = self.robot_skeleton.x.copy()
@@ -155,12 +161,14 @@ class TwoStepEnv:
         while True:
             if steps_per_render and self.world.frame % steps_per_render == 0:
                 self._render()
-            step_dist = self.simulation_step()
-            if type(step_dist) == str:
-                self.log("ERROR: " + step_dist)
+            status_code, step_dist, status_string = self.simulation_step()
+            if status_code == StepResult.ERROR:
+                self.log("ERROR: " + status_string)
                 return None, None
-            if step_dist:
+            if status_code == StepResult.COMPLETE:
                 end_state = self.current_observation()
+                if render:
+                    self.log(status_string)
                 return end_state, step_dist
 
     def step(self, action):

### 2018-06-28 source/simbicon.py
Enforce joint limits (and fix skel file)

@@ -96,8 +96,8 @@ class Simbicon(PDController):
         result = None
         if self.direction == DOWN:
             self.swing_idx, self.stance_idx = self.stance_idx, self.swing_idx
-            self.direction = UP
             result = "{:.3f}: Ended state {} {}{}".format(self.time(), swing, self.direction, suffix)
+            self.direction = UP
         else:
             # TODO skip this state if the swing foot is already in contact?
             self.direction = DOWN

### 2018-06-28 source/walker.py

@@ -36,6 +36,8 @@ class TwoStepEnv:
         walker.bodynodes[5].set_friction_coeff(FRICTION_COEFF)
         walker.bodynodes[8].set_friction_coeff(FRICTION_COEFF)
         world.skeletons[0].bodynodes[0].set_friction_coeff(FRICTION_COEFF)
+        for j in walker.joints:
+            j.set_position_limit_enforced()
 
         self.controller = controller_class(walker, self)
         walker.set_controller(self.controller)
@@ -225,9 +227,7 @@ class TwoStepEnv:
             self.put_dot(t, 0, i)
 
 def load_world():
-    DARTENV_SKEL_ROOT = "/home/nathan/research/dart-env/gym/envs/dart/assets/"
-    skel = DARTENV_SKEL_ROOT + "walker2d.skel"
-
+    skel = "walker2d.skel"
     pydart.init()
     world = pydart.World(SIMULATION_RATE, skel)
     # These will mark footstep target locations

### 2018-06-28 source/pd_control.py
Clamp control; reduce gains and friction coeff

@@ -3,8 +3,8 @@ import numpy as np
 
 # TODO there is some foot slip happening. Is this because these gains are too high?
 BRICK_DOF = 3 # We're in 2D
-KP_GAIN = 300.0
-KD_GAIN = 30.0
+KP_GAIN = 200.0
+KD_GAIN = 15.0
 
 # TODO implement Stable PD controllers?
 class PDController:
@@ -15,6 +15,7 @@ class PDController:
         self.reset()
         self.Kp = np.array([0.0] * BRICK_DOF + [KP_GAIN] * (self.skel.ndofs - BRICK_DOF))
         self.Kd = np.array([0.0] * BRICK_DOF + [KD_GAIN] * (self.skel.ndofs - BRICK_DOF))
+        self.control_bounds = 1.5 * np.array([100, 100, 20, 100, 100, 20])
 
     def set_actuated_pose(self, pose):
         self.target_q[BRICK_DOF:] = pose
@@ -22,8 +23,13 @@ class PDController:
     def compute(self):
         if self.inactive:
             return np.zeros_like(self.Kp)
-        # TODO clamp control
-        return -self.Kp * (self.skel.q - self.target_q) - self.Kd * self.skel.dq
+        control = -self.Kp * (self.skel.q - self.target_q) - self.Kd * self.skel.dq
+        for i in range(BRICK_DOF, self.skel.ndofs):
+            if control[i] > self.control_bounds[i-BRICK_DOF]:
+                control[i] = self.control_bounds[i-BRICK_DOF]
+            if control[i] < -self.control_bounds[i-BRICK_DOF]:
+                control[i] = -self.control_bounds[i-BRICK_DOF]
+        return control
 
     def state_complete(self, left, right):
         # Stub to conform to Simbicon interface

### 2018-06-28 source/walker.py

@@ -16,7 +16,7 @@ SIMULATION_RATE = 1.0 / 2000.0 # seconds
 EPISODE_TIME_LIMIT = 5.0 # seconds
 REAL_TIME_STEPS_PER_RENDER = 25 # Number of simulation steps to run per frame so it looks like real time. Just a rough estimate.
 
-FRICTION_COEFF = 100.0
+FRICTION_COEFF = 5.0
 
 class StepResult(Enum):
     ERROR = -1

### 2018-06-28 source/inverse_dynamics.py
Tiny bugfix

@@ -30,7 +30,7 @@ class LearnInverseDynamics:
 
     def initialize_start_states(self):
         if not os.path.exists(START_STATES_FILENAME):
-            states = env.collect_starting_states()
+            states = self.env.collect_starting_states()
             with open(START_STATES_FILENAME, 'wb') as f:
                 pickle.dump(states, f)
         with open(START_STATES_FILENAME, 'rb') as f:

### 2018-06-28 source/experiment.py
Clean up training curve plot

@@ -3,44 +3,70 @@ import numpy as np
 import pickle
 import os
 import matplotlib.pyplot as plt
+import seaborn as sns
 from inverse_dynamics import LearnInverseDynamics
+
+N_EVAL_TRAJECTORIES = 8
 
 
 class Experiment:
-    def __init__(self, learner):
+    def __init__(self, name, learner):
         self.learner = learner
+        self.name = name
+        self.iter = 0
+        self.avg_losses = []
+        self.max_errors = []
+        self.n_eval_trajectories = N_EVAL_TRAJECTORIES
+        self.run_evaluations()
 
-    def repeat_eval(self, n_eval_repeats):
+    def checkpoint(self):
+        # TODO add checkpointing (save model params, collected data, training curve statistics)
+        pass
+
+    def run_evaluations(self):
         avg_losses = []
         max_errors = []
-        for k in range(n_eval_repeats):
+        for k in range(self.n_eval_trajectories):
             print("Starting evaluation", k)
             avg_loss, max_error = self.learner.evaluate()
             avg_losses.append(avg_loss)
             max_errors.append(max_error)
-        return np.mean(avg_losses), np.std(avg_losses), np.mean(max_errors), np.std(max_errors)
+        self.avg_losses.append(avg_losses)
+        self.max_errors.append(max_errors)
 
-    def multi_evaluate(self, n_iters=3, n_eval_repeats=8):
-        results = np.zeros((n_iters+1, 4))
-        results[0] = self.repeat_eval(n_eval_repeats)
+    def run_iters(self, n_iters):
         for i in range(n_iters):
             self.learner.training_iter()
-            results[i+1] = self.repeat_eval(n_eval_repeats)
-        self.plot_results(results)
+            self.run_evaluations()
+            self.iter += 1
+        self.plot_results()
 
-    def plot_results(self, results):
-        x = list(range(results.shape[0]))
-        avg_loss_line, = plt.plot(x, results[:,0], color="red")
-        max_error_line, = plt.plot(x, results[:,2], color="blue")
-        plt.fill_between(x, results[:,0]-results[:,1], results[:,0]+results[:,1], color="red", alpha=0.2)
-        plt.fill_between(x, results[:,2]-results[:,3], results[:,2]+results[:,3], color="blue", alpha=0.2)
-        plt.savefig('{}.png'.format("thingy"))
+    def plot_results(self):
[…]

### 2018-06-28 source/inverse_dynamics.py

@@ -116,7 +116,9 @@ class LearnInverseDynamics:
         error = np.abs(step_dist - target[0])
         return state, loss, error
 
-    def evaluate(self, n_steps=8, render=None, record_video=False):
+    def evaluate(self, n_steps=8, render=None, record_video=False, seed=None):
+        if seed:
+            self.env.seed(seed)
         state = self.env.reset(record_video=record_video)
         total_loss = 0
         max_error = 0

### 2018-06-28 source/experiment.py
Don't set target velocities
For now, we just care about getting the step distance right.

@@ -14,7 +14,7 @@ class Experiment:
         self.learner = learner
         self.name = name
         self.iter = 0
-        self.avg_losses = []
+        self.avg_errors = []
         self.max_errors = []
         self.n_eval_trajectories = N_EVAL_TRAJECTORIES
         self.run_evaluations()
@@ -24,37 +24,38 @@ class Experiment:
         pass
 
     def run_evaluations(self):
-        avg_losses = []
+        avg_errors = []
         max_errors = []
         for k in range(self.n_eval_trajectories):
             print("Starting evaluation", k)
-            avg_loss, max_error = self.learner.evaluate()
-            avg_losses.append(avg_loss)
+            avg_error, max_error = self.learner.evaluate()
+            avg_errors.append(avg_error)
             max_errors.append(max_error)
-        self.avg_losses.append(avg_losses)
+        self.avg_errors.append(avg_errors)
         self.max_errors.append(max_errors)
 
     def run_iters(self, n_iters):
         for i in range(n_iters):
+            self.iter += 1
+            print("Starting training iteration ", self.iter)
             self.learner.training_iter()
             self.run_evaluations()
-            self.iter += 1
         self.plot_results()
 
     def plot_results(self):
         n_points = self.iter+1
-        avg_losses = np.array(self.avg_losses)
+        avg_errors = np.array(self.avg_errors)
         max_errors = np.array(self.max_errors)
         x = range(n_points)
-        avg_loss_line, = plt.plot(x, np.mean(avg_losses, 1), color="red")
+        avg_error_line, = plt.plot(x, np.mean(avg_errors, 1), color="red")
         max_error_line, = plt.plot(x, np.mean(max_errors, 1), color="blue")
-        plt.fill_between(x, np.min(avg_losses, 1), np.max(avg_losses, 1), color="red", alpha=0.2)
+        plt.fill_between(x, np.min(avg_errors, 1), np.max(avg_errors, 1), color="red", alpha=0.2)
         plt.fill_between(x, np.min(max_errors, 1), np.max(max_errors, 1), color="blue", alpha=0.2)
 
         plt.title("Training curve")
         plt.xlabel("Number of data collection iterations")
-        plt.ylabel("Average loss / maximum placement error")
-        plt.legend([avg_loss_line, max_error_line], ["Average loss", "Max error"])
+        plt.ylabel("Foot placement error")
[…]

### 2018-06-28 source/inverse_dynamics.py

@@ -20,10 +20,10 @@ class LearnInverseDynamics:
         self.env = env
         self.initialize_start_states()
         self.train_set = []
-        # Use a linear policy for now
-        target_space_shape = 2
+        target_space_shape = 1
         self.n_action = env.action_space.shape[0]
         self.n_dynamic = env.observation_space.shape[0] + target_space_shape
+        # Use a linear policy for now
         self.lm = Ridge(alpha=RIDGE_ALPHA)
         # Initialize the model with all zeros
         self.lm.fit(np.zeros((1, self.n_dynamic)), np.zeros((1,self.n_action)))
@@ -53,7 +53,7 @@ class LearnInverseDynamics:
             self.env.reset(start_state)
             end_state, step_dist = self.env.simulate(action)
             if end_state is not None:
-                achieved_target = (step_dist, end_state[9]) # Include ending velocity
+                achieved_target = [step_dist] # Include ending velocity
                 self.train_set.append((start_state, action, achieved_target))
                 self.start_states.append(end_state)
             else:
@@ -69,13 +69,8 @@ class LearnInverseDynamics:
     def generate_target(self, prev_target=None):
         # Choose some random targets that are hopefully representative of what
         # we will see during testing episodes.
-        if prev_target is not None:
-            current_v = prev_target[1]
-        else:
-            current_v = 1.0
-        target_dist = current_v * 0.5 + (0.5 * np.random.uniform() - 0.25)
-        target_v = current_v + (0.3 * np.random.uniform() - 0.15)
-        target = [target_dist, target_v]
+        target = 0.5 + (0.5 * np.random.uniform() - 0.25)
+        target = [target]
         self.show_target(target)
         return target
 
@@ -89,7 +84,6 @@ class LearnInverseDynamics:
         print(len(self.start_states), len(self.train_set))
 
     def training_iter(self, demo=True):
-        print("Starting new training iteration")
         self.collect_dataset()
         self.train_inverse_dynamics()
         if demo:
@@ -112,26 +106,25 @@ class LearnInverseDynamics:
             # The robot crashed or something. Just hack so the script doesn't *also* crash.
             state = start_state
             step_dist = -1 # Should make this bigger, but then the graph gets way out of scale
-        loss = (step_dist - target[0])**2 + (state[9] - target[1])**2
         error = np.abs(step_dist - target[0])
-        return state, loss, error
+        return state, error
 
[…]

### 2018-07-16 paper/main.tex
Draft updated introduction and related work

@@
 
-\title{Curriculum and Transfer Learning of Hierarchical Locomotion Policies}
+\title{Curriculum and Imitation Learning for Biped Locomotion Over Rough Terrain}
 \author{TBD}
@@
 \end{itemize}
+
+\section{Introduction}
+
+Locomotion is a difficult, underactuated control problem.
+\emph{Biped} locomotion is more difficult because the support polygon is small and the center of mass is relatively high.
+Biped locomotion across \emph{rough terrain} is more difficult still, because the variability in the surface normal makes the dynamics harder to model.
+
+One way to handle these challenges is with detailed models and extensive optimization.
+For example, the WPI-CMU team in the DARPA Robotics Challenge (DRC) controlled ATLAS using a hierarchical optimization approach that carefully controlled the robot's center of mass (CoM) and center of pressure (CoP) \citep{feng2015optimization}.
+Such approaches tend to be reliable but slow.
+They are physically slow because accurate modeling requires relatively static walking, and also because the extensive optimization procedures are computationally slow.
+For the DRC, reliability was more important than speed, but eventually it would be good to have robots that can traverse rough terrain using faster, dynamic gaits.
+\footnote{\nhatch{It might also be worth mentioning that developing accurate models takes a lot of time and expertise, and even then, physics-based models don't account for many important physical phenomena.}}
+
+Recent advances in reinforcement learning (RL) suggest an alternative approach.
+Deep RL is a popular technique to learn highly dynamic motions for humanoids \citep{peng2018deepmimic, heess2017emergence}.
+The resulting policies are both computationally and physically fast.
+
+Building on this work, we approach dynamic biped locomotion over rough terrain using a combination of curriculum learning, imitation learning, and reinforcement learning.
+The controller learns new skills using RL, which are generalized to a fast policy using imitation learning.
+Use of a curriculum allows the controller to build up from flat ground to a challenging combination of stepping stones at various heights, distances, and orientations.
+In contrast to prior work in RL, our controller \emph{plans its footstep locations}, which allows it to handle larger and more frequent obstacles.
+In contrast to prior work in footstep planning, our controller learns a dynamic policy that requires no optimization at execution time.
+
+We demonstrate the real-world potential of this algorithm by running it on a simulated Darwin humanoid robot. \nhatch{Add a paragraph on sim-to-real.}
+
+\nhatch{Here's the order of steps that we would like to take to make sure we can actually achieve that:
+
+1. With the simple 2D walker, be able to navigate the challenging stepping stones environment (steps at various heights, distances, and orientations)
+2. Move to 3D, but still with a simplified robot model.
+3. Switch to the Darwin model.}
+
+\section{Related work}
+
+\subsection{Step planning}
+
+As discussed in the introduction, there are many papers about locomotion on physical robots that use detailed models and a lot of optimization.
+These approaches tend to use a hierarchy of planning steps to reduce the amount of computation.
+First, they generate a plan for footstep locations using search, like A*.
+Based on the footstep plan, they generate a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum (SLIP).
+Then, they use inverse kinematics to plan trajectories for end-effectors.
+Finally, these trajectories are optimized using the full dynamics model to account for details like obstacles.
+Examples of this include several of the DARPA Robotics Challenge teams, as well as the DARPA Learning Locomotion Challenge teams.
+\nhatch{Put some examples from other DRC teams as well.} Further examples of this are the zero-moment point (ZMP) approaches from Honda / ASIMO.
+A similar approach (though not used on a physical robot) is \cite{mordatch2010robust}.
+These approaches tend to be reliable but slow, because static walking is easier to model and optimization takes a lot of computation time.
+They also require a lot of domain knowledge and manual tuning in order to engineer good models and handle situations when the real world deviates from ideal physics.
+
+\nhatch{Add some discussion of why reliability isn't as important for our current project.}
+
+There are a few prior works that have used reinforcement learning to do step planning.
+\cite{peng2017deeploco} uses a hierarchical reinforcement learning algorithm where the upper-level controller chooses footstep locations and the lower-level controller executes those plans.
+However, the path that the agent follows is perfectly flat, and the target step locations are rarely actually achieved.
+On rough terrain, such inaccurate step planning would probably fail.
+\cite{karpathy2012curriculum} uses curriculum learning to collect a set of states and actions together with the step distance that they achieve.
+However, the final policy is based on nearest-neighbor search among this set of collected experience.
+With a higher-dimensional, more complicated model, this approach would probably fail.
+
+\subsection{Rough terrain}
+
+Most of the prior work on DRL for biped locomotion over rough terrain does not explicitly plan individual footstep locations.
+Examples of this are \cite{peng2018deepmimic, heess2017emergence, peng2016terrain}.
+The resulting policies travel forward quickly and are robust to small missteps, but would fail in situations with dense or large obstacles where small missteps can be catastrophic.
+
+\subsection{Policy representations and learning algorithms(?)}
+
+This is the prior work on which our learning algorithm is most directly based:
+
+- Recht, random search instead of fancy RL algorithms
+- SIMBICON, to inform our choice of action parametrization
+
+%Learning algorithm:
+%- use ~20 parameters for each *entire step* (based on SIMBICON)
+%- collect a training set of diverse actions/skills using random search
+%- generalize by learning ``inverse dynamics'' (really just a policy where the observation includes the next step target)
+%- perhaps repeat (curriculum learning)
 

### 2018-07-16 source/inverse_dynamics.py
Massive refactor: Support 2D targets and use absolute coordinates
It was making my head hurt to remember which function parameters were
supposed to be in relative coordinates and which in absolute (and what
the coordinates were supposed to be relative to). This commit refactors
so everything is in absolute coordinates. The method `center_state` in
inverse_dynamics now handles moving things to relative coordinates for
the purposes of training and using the model.

This commit also changes targets to (x,y) instead of just (x), in
anticipation of future work where the targets will not all be in t

@@ -32,11 +32,11 @@ class LearnInverseDynamics:
     def __init__(self, env):
         self.env = env
         self.initialize_start_states()
-        self.train_set = []
+        self.train_states, self.train_targets = [], []
         target_space_shape = 1
         self.n_action = env.action_space.shape[0]
         # The actual target is sort of part of the state, hence the *2.
-        self.n_dynamic = env.observation_space.shape[0] + target_space_shape*2
+        self.n_dynamic = 26
         # Use a linear policy for now
         model = Ridge(alpha=RIDGE_ALPHA, fit_intercept=False)
         #model = make_pipeline(PolynomialFeatures(2), model) # quadratic
@@ -65,57 +65,86 @@ class LearnInverseDynamics:
         start_states = []
         self.env.controller.set_gait_raw(np.zeros(self.env.action_space.shape[0]))
         starter = 0.3
-        self.env.put_grounds([], runway_length=100)
+        self.env.put_grounds([[0,0]], runway_length=100)
         for i in range(n_resets):
             length = min_length + (max_length - min_length) * (i / n_resets)
             self.env.log("Starting trajectory {}".format(i))
             self.env.reset()
             # TODO should we include this first state? It will be very different from the rest.
             #start_states.append(self.env.robot_skeleton.x)
-            target = 0
+            target = np.array([0,0])
             for j in range(size):
-                prev_target, target = target, starter + length * j
-                end_state, _ = self.env.simulate(target, render=0.1)
-                if end_state is not None:
-                    # We need to collect the location of the previous target in order to
-                    # place stepping stones properly when resetting to this state.
-                    prev_platform = prev_target - target
-                    start_states.append((end_state, prev_platform))
+                prev_target, target = target, np.array([starter + length * j, 0])
+                end_state = self.env.simulate(target, render=0.1)
+                # We need to collect the location of the previous target in order to
+                # place stepping stones properly when resetting to this state.
+                start_states.append(end_state)
         self.env.clear_skeletons()
         return start_states
 
     def dump_train_set(self):
         with open(TRAIN_FILENAME, 'wb') as f:
-            pickle.dump((self.start_states, self.train_set), f)
+            pickle.dump((self.start_states, self.train_states, self.train_targets), f)
 
     def load_train_set(self):
         with open(TRAIN_FILENAME, 'rb') as f:
[…]

### 2018-07-16 source/sdf_loader.py

@@ -1,4 +1,5 @@
 import os
+from IPython import embed
 
 # SDF color format is space-separated RGBA
 # Here are some suggested settings
@@ -12,7 +13,7 @@ class SDFLoader:
         self.num_dots = 0
         self.grounds = []
 
-    def put_dot(self, x, y, color=RED):
+    def put_dot(self, target, color=RED):
         # Change the skeleton name so that the console output is not cluttered
         # with warnings about duplicate names.
         os.system("sed -e 's/__NAME__/dot_" + str(self.num_dots) + "/' skel/dot.sdf > skel/_dot.sdf")
@@ -21,12 +22,11 @@ class SDFLoader:
 
         dot = self.world.add_skeleton('./skel/__dot.sdf')
         q = dot.q
-        q[3] = x
-        q[4] = y
+        q[3:5] = target
         dot.q = q
 
     # Length is in meters.
-    def put_ground(self, x, length, index):
+    def put_ground(self, x, y, length, index):
         num_grounds = len(self.grounds)
         if num_grounds <= index:
             # Change the skeleton name so that the console output is not cluttered
@@ -41,6 +41,7 @@ class SDFLoader:
         q = ground.q
         # The x coordinate q[3] gives the *center* of the block.
         q[3] = x + 0.5*length
+        q[4] = y
         ground.q = q
 
 if __name__ == "__main__":

### 2018-07-16 source/simbicon.py

@@ -41,16 +41,21 @@ class Simbicon(PDController):
         self.reset()
         self.ik = InverseKinematics(env)
 
-    def reset(self):
+    def reset(self, state=np.zeros(8)):
         self.state_started = self.time()
         self.swing_idx = 3
         self.stance_idx = 6
-        self.contact_x = 0
-        self.target_x = 0
-        self.stance_heel = 0
+        self.contact = state[0:2]
+        self.stance_heel = state[2:4]
+        self.target = state[4:6]
+        # We track prev_target so that, in principle, we know all of the possible
+        # places where the agent is in contact with the environment. However, the
+        # algorithm does not use this (yet) except when resetting the environment
+        # to a previously visited state.
+        self.prev_target = state[6:8]
         self.direction = UP
 
-    def set_gait_raw(self, target_x, raw_gait=None):
+    def set_gait_raw(self, target, raw_gait=None):
         up = walk[UP].raw_params.copy()
         down = walk[DOWN].raw_params.copy()
         if raw_gait is not None:
@@ -61,8 +66,8 @@ class Simbicon(PDController):
 
         # All of these adjustments are just rough linear estimates from
         # fiddling around manually.
-        self.target_x = target_x
-        step_dist_diff = self.target_x - self.stance_heel - 0.4
+        self.target, self.prev_target = target, self.target
+        step_dist_diff = self.target[0] - self.stance_heel[0] - 0.4
         gait[UP].stance_knee_relative  += -0.05 - step_dist_diff * 0.2
         gait[UP].stance_ankle_relative += 0.2 + step_dist_diff * 0.4
         gait[UP].swing_hip_world       += 0.4 + step_dist_diff * 0.4
@@ -72,6 +77,9 @@ class Simbicon(PDController):
         self.FSM = gait
 
     def state(self):
+        return np.concatenate((self.contact, self.stance_heel, self.target, self.prev_target))
+
+    def FSMstate(self):
         return self.FSM[self.direction]
 
     def time(self):
@@ -85,22 +93,21 @@ class Simbicon(PDController):
             early_strike = (duration >= LIFTOFF_DURATION) and (len(contacts_to_check) > 0)
             if early_strike:
                 print("Early strike!")
-            target_diff = self.state().ik_gain * self.skel.dq[0]
+            target_diff = self.FSMstate().ik_gain * self.skel.dq[0]
             # Start DOWN once heel is close enough to target (or time expires)
[…]

### 2018-07-16 source/walker.py

@@ -75,7 +75,6 @@ class TwoStepEnv:
 
     def reset(self, state=None, record_video=False, random=1.0):
         self.world.reset()
-        self.controller.reset()
         self.set_state(state, random)
         if self.video_recorder:
             self.video_recorder.close()
@@ -92,32 +91,28 @@ class TwoStepEnv:
     # Returns (status code, step distance, human-readable message) tuple.
     def simulation_step(self):
         if self.world.time() > EPISODE_TIME_LIMIT:
-            return StepResult.ERROR, None, "Time limit reached"
+            return StepResult.ERROR, "Time limit reached"
         obs = self.current_observation()
         if not np.isfinite(obs).all():
-            return StepResult.ERROR, None, "Numerical explosion"
+            return StepResult.ERROR, "Numerical explosion"
         if obs[1] < -0.5:
-            return StepResult.ERROR, None, "Crashed"
+            return StepResult.ERROR, "Crashed"
 
         l_contact = self.find_contacts(self.l_foot)
         r_contact = self.find_contacts(self.r_foot)
-        state_complete, step_dist = self.controller.state_complete(l_contact, r_contact)
+        state_complete, step_complete = self.controller.state_complete(l_contact, r_contact)
         if state_complete:
             status_string = self.controller.change_state()
-            if step_dist:
-                return StepResult.COMPLETE, step_dist, status_string
+            if step_complete:
+                return StepResult.COMPLETE, status_string
 
         self.world.step()
-        return StepResult.IN_PROGRESS, None, None
+        return StepResult.IN_PROGRESS, None
 
     def current_observation(self):
         obs = self.robot_skeleton.x.copy()
         obs = self.standardize_stance(obs)
-        # Exact x location doesn't affect dynamics. Center so target_x is at 0.
-        base_x = self.controller.target_x
-        obs[0] -= base_x
-        stance = np.array([self.controller.contact_x, self.controller.stance_heel]) - base_x
-        return np.concatenate((obs, stance))
+        return np.concatenate((obs, self.controller.state()))
 
     def standardize_stance(self, state):
         # Ensure the stance state is contained in state[6:9] and state[15:18].
@@ -144,31 +139,29 @@ class TwoStepEnv:
             dq = self.robot_skeleton.dq
             dq[0] += 0.8 + random * np.random.uniform(low=0.0, high=0.4)
             self.robot_skeleton.dq = dq
+            self.controller.reset()
         else:
             self.robot_skeleton.x = state[:18]
[…]

### 2018-08-03 paper/litreview.tex
Update paper draft
Main addition: a description of our algorithm as currently implemented.
Also split the old literature review into a separate TeX file.

--- 2018-06-28 paper/main.tex
@@
 \maketitle
-
-Definitions:
-\begin{itemize}
-  \item HLC: high-level controller
-  \item LLC: low-level contoller
-  \item HRL: hierarchical reinforcement learning
-\end{itemize}
-
-\section{Introduction}
-
-Locomotion is hard because the changing contacts make the dynamics difficult to model.
-Recent deep reinforcement learning approaches try to learn this model implicitly via the policy network, which has trouble learning such complicated models, possibly due to the limitations of current optimization methods (like policy gradient).
-Such approaches fail in environments that require predictive planning at the level of individual contacts, such as running across irregular terrain, where small miscalculations have disproportionally large effects.
-
-We aim to simplify the learning problem by imposing prior structure on the policy: we split it into a two-level hierarchy.
-The higher level breaks the plan into smaller chunks, which are easier to achieve.
-The lower level must then execute these smaller chunks.
-Each chunk is a single footstep.
-This choice simplifies the dynamics that the lower level needs to model, because each step involves no (or very few) changes of contact.
-It also simplifies the upper level learning problem, since the controller does not need to handle details of torque, other than knowing which step plans can actually be achieved.
-
-\section{Experiments}
-
-Our experiments are designed to test two hypotheses.
-
-Our primary hypothesis is that this hierarchical architecture can navigate a variety of complicated locomotion environments, learning faster and achieving better results than other reinforcement learning methods.
-These locomotion environments include flat ground, inclines, stairs, random inclines, stepping stones, hurdles, and combinations of the above.
-
-Our two metrics for ``learning faster'' are (1) number of simulation frames of experience collected, and (2) total computation time.
-We consider both necessary, because the relative expense of more experience versus more computation may vary across application domains.
-
-We collect both quantitative and qualitative metrics to measure ``achieving better results''.
-Quantitatively, we compare the average rate of forward progress.
-We also consider the robustness of the model to force perturbations and various measures of the ``difficulty'' of the terrain (such as stair height).
-\footnote{Another interesting question would be robustness with respect to accuracy of perception.
-If the potential step target locations / surface normals are perturbed, what happens?}
-Qualitatively, we document motion quality by means of supplementary videos, which we encourage the reader to consult.
-
-Our secondary hypothesis is that the learned policy can be run in real-time while still achieving results comparable to more expensive, model-based optimization methods.
-
-This hypothesis is difficult to test, because model-based optimization methods are difficult to implement for direct comparison.
-Thus, our metric is simply the percent real time necessary to compute the control for the next time step, which we compare to numbers reported in optimization-based papers.
-
-Additionally, we conduct ablative analysis of the importance of various components of our approach.
-\nhatch{In particular, we want to consider various different HLC / LLC interfaces that we could have used.
-For example: should we include a target foot height that the LLC should reach?
-(I think this would help when going over hurdles.)
-For another example (assuming we use some kind of PD control at the lowest level): how much does performance degrade if we don't use PD control?
-(That is, if the LLC outputs raw torques.)}
-
-\section{Policy architecture}
-
-\nhatch{Ideally, maybe we should design this so that walking and running can both be achieved in the same episode.}
-
-\nhatch{Should there be some ensemble of low-level policy networks that the HLC can choose between?
-For example, we could take the policy with the highest estimated value given the target step location.
-Having some discrete choices like this might be easier than using solely continuous parameters.}
-
-\textbf{High-level observations.}
-We assume that perception is solved.
-The high-level controller receives proprioceptive information and environmental information.
-The environmental information consists of a list of potential surfaces to step on, specified in terms of relative displacement from the agent and surface normal.
-\nhatch{Maybe, for polygonal surfaces, we could also report the surface ``size'' (e.g. largest-radius circle that fits inside the polygon).}
-
-\textbf{High-level actions.}
-High-level actions are chosen on each heel-down event (once per footstep).
-At these events, the HLC outputs a target for the next footstep that it expects the LLC can achieve.
-This target consists of several parameters: \nhatch{TODO think about this some more}
-\nhatch{It might be possible to incorporate some optimization / search / planning into the HLC, because its actions happen infrequently.
-But this would require some objective to be optimized, which might involve implementing a model like a spring-loaded inverted pendulum (SLIP), or inventing some heuristic objective that might characterize good plans.
-We might want to implement a model like SLIP anyway, in order to use it as an expert to train a neural network for the HLC (similar to AlphaGo Zero and ExIt).
-In any case, only the first step of the plan would be revealed to the LLC.}
-
-\textbf{Low-level observations.}
-The low-level controller receives proprioceptive information and the target location for the next footstep.
-Unlike the HLC, it does not observe the external environment.
-
-\textbf{Low-level actions.}
-On the one hand, we could have low-level actions being the raw torques.
-On the other hand, this might make the learning problem harder.
-Instead, we could specify ``abstract actions'' and use PD controllers or other analytical formulas to derive the actual torques.
-For instance, the output could be a target pose or end-effector position which is chosen more infrequently than the simulation framerate.
-Then we could solve an inverse kinematics or inverse dynamics problem to calculate a target pose, which is then achieved using PD controllers.
-\nhatch{We probably don't need all of this machinery.
-For the initial proof of concept, it might make sense to see how we can do using just raw torques.}
-\nhatch{TODO think about this some more.}
-
-\textbf{PD control.}
-\nhatch{We might have to rethink this abstraction layer once the ground includes inclines.
-For instance, the PD controller might now have to handle different target ankle angles.}
-
-\textbf{Miscellaneous ideas.}
-\begin{itemize}
-  \item Could the HLC-LLC interface be in terms of a \emph{force} that the LLC should try to generate on the next foot contact?
-  \item How can we design the LLC to generate good motion quality (e.g. low energy, low angular momentum)?
-\end{itemize}
-
-\section{Timeline of work}
-
-The goal is to submit this to ICRA on September 15.
-
-\begin{itemize}
-  \item Early June (2 weeks): Proof-of-concept: train a 2D walker low-level controller to follow an irregular step plan on flat ground, including some low-level PD control if necessary.
-  \item Late June (2 weeks): Train low-level controller on stepping stones.
-    Train high-level controller on flat ground, then on stepping stones.
-    Compare performance on stepping stones to baseline deep RL approach (TRPO).
-    If we can't get this to work (better than TRPO), we should consider abandoning the project.
-  \item July (4 weeks): Implement and train on inclines, stairs, and hurdles.
-    Perform ablative analyses.
-  \item Early August (2 weeks): I (Nathan) will be on vacation.
-  \item Late August and early September (4 weeks): Last-minute experiments.
-    Write and revise the paper.
-  \item September 15: ICRA deadline.
-\end{itemize}
 
@@
 
-\section{Project summary}
-
-\begin{enumerate}
-
-\item \textbf{One-sentence summary:}
-Learn a difficult locomotion task with curriculum learning, hierarchical policies, and possibly model-based planning.
-
-\item \textbf{Task:}
-Train a simulated robotic agent to accomplish locomotion in a difficult environment, such as walking (or preferably running) across bumpy ground.
-
-\item \textbf{Introduction:} Robotic locomotion on uneven terrain is an important problem.
-  \nhatch{Visak, I would be interested in your high-level reasons for being interested in this problem.
-  I'm not sure I can motivate the problem as well as you can.} Previous approaches have not solved this problem.
-  For instance, in \href{https://www.youtube.com/watch?v=hx_bgoTF7bs&feature=youtu.be&t=1m28s}{this video} the agent stumbles on rubble, and DeepLoco only walks on flat terrain.
-  \nhatch{More examples would help, to make sure that this problem is interesting.} We hope to achieve efficient locomotion across bumpy ground using a combination of three techniques.
-  \begin{enumerate}
-    \item Curriculum learning: Train the agent first on simpler environments, like flat ground and gradual slopes, and gradually increase the difficulty.
-    \item Hierarchical controllers: We use a hierarchical architecture where the HLC plans foot placement and the LLC executes the individual steps.
-      We hypothesize that this architecture will more easily handle the discontinuities caused by contact dynamics.
-      We also hypothesize that, with this architecture, it will be easier to transfer knowledge between tasks in the curriculum.
-    \item Model-based planning: To help with sample efficiency, the HLC has access to a physics model for the environment.
-      \nhatch{I'm not sure how to accomplish this.
-      Physics models for legged robots are probably hard, especially when the ground is not flat.
-      Maybe some kind of \emph{learned} model would work.}
-  \end{enumerate}
-
-\item \textbf{Controller architecture:}
-  Like DeepLoco, the HLC receives detailed environmental observations, and outputs the target locations for the next couple of foot placements.
-  The LLC takes this input to choose the low-level actions that will achieve those foot placements.
-  Unlike DeepLoco, the terrain is much more complicated, so it is important that the HLC foot placements are actually feasible.
-  Furthermore, we will train the LLC without using motion captures, and we will permit the time per step to vary, rather than requiring each step to take exactly 0.5 seconds.
-
-  The parametrization of the HLC and LLC will probably end up being some kind of neural network, although it might be worth trying simpler policies first (e.g. linear).
-
-  To train these controllers, we could first train the LLC on simpler terrain, where we can specify target foot placement using a simple calculation.
-  When training the HLC, to ensure that it chooses feasible foot placements, we could include some additional loss term in the training objective based on whether the LLC actually achieved the foot placement that the HLC suggested.
-
-\nhatch{This section could probably use more detail; suggestions are welcome.
-For example, do we need to modify this architecture to better handle contact dynamics?
-Is there some way we can incorporate a model to help the HLC make better plans?}
-
-\item \textbf{Why this might be a good idea:}
-I have not seen previous papers succeed at this task, so if we could show some nice videos of a quadruped or biped running along bumpy ground, it would catch people's attention.
-Intuitively, when running along bumpy hiking trails, I am consciously thinking about where to place my feet, then relying on unconsciously being able to place my foot there while I think about the next step.
-This suggests a kind of hierarchical policy.
-
-\item \textbf{Why this might not be a good idea:}
-  \nhatch{Why would *hierarchical* RL be a particularly good way to do transfer learning for this series of tasks?}
-  We will need to implement several other baselines to compare against.
-  \nhatch{What should those baselines be? (What other approaches have people used for locomotion on bumpy ground?)
-Is the ``bumpy ground'' task too difficult? (Byron thinks so.)
-Will the simulated robot need toes?}
-I haven't worked with locomotion before, so I don't have a good sense for this.
-
-\end{enumerate}
-
 \bibliographystyle{plainnat}

### 2018-08-03 paper/main.tex
Update paper draft
Main addition: a description of our algorithm as currently implemented.
Also split the old literature review into a separate TeX file.

--- 2018-06-12 paper/main.tex
@@
 
-Definitions:
-\begin{itemize}
-  \item HLC: high-level controller
-  \item LLC: low-level contoller
-  \item HRL: hierarchical reinforcement learning
-\end{itemize}
-
 \section{Introduction}
 
-Locomotion is hard because the changing contacts make the dynamics difficult to model.
-Recent deep reinforcement learning approaches try to learn this model implicitly via the policy network, which has trouble learning such complicated models, possibly due to the limitations of current optimization methods (like policy gradient).
-Such approaches fail in environments that require predictive planning at the level of individual contacts, such as running across irregular terrain, where small miscalculations have disproportionally large effects.
+Locomotion is a difficult, underactuated control problem.
+\emph{Biped} locomotion is more difficult because the support polygon is small and the center of mass is relatively high.
+Biped locomotion across \emph{rough terrain} is more difficult still, because the variability in the surface normal makes the dynamics harder to model.
 
-We aim to simplify the learning problem by imposing prior structure on the policy: we split it into a two-level hierarchy.
-The higher level breaks the plan into smaller chunks, which are easier to achieve.
-The lower level must then execute these smaller chunks.
-Each chunk is a single footstep.
-This choice simplifies the dynamics that the lower level needs to model, because each step involves no (or very few) changes of contact.
-It also simplifies the upper level learning problem, since the controller does not need to handle details of torque, other than knowing which step plans can actually be achieved.
+One way to handle these challenges is with detailed models and extensive optimization.
+For example, the WPI-CMU team in the DARPA Robotics Challenge (DRC) controlled ATLAS using a hierarchical optimization approach that carefully controlled the robot's center of mass (CoM) and center of pressure (CoP) \citep{feng2015optimization}.
+Such approaches tend to be reliable but slow.
+They are physically slow because accurate modeling requires relatively static walking, and also because the optimization procedures are computationally slow.
+For the DRC, reliability was more important than speed, but eventually we want to have robots that can traverse rough terrain using faster, dynamic gaits.
+
+Recent advances in reinforcement learning (RL) suggest an alternative approach.
+Deep RL is a popular technique to learn highly dynamic motions for humanoids \citep{peng2018deepmimic, heess2017emergence}.
+The resulting policies are both computationally and physically fast.
+However, their effectiveness on rough terrain is uncertain.
+
+Building on this work, we approach dynamic biped locomotion on rough terrain using a combination of curriculum learning, imitation learning, and reinforcement learning, which we call CIRL.
+\nhatch{Just a working title.}
+The controller learns new skills using RL, which are generalized to a fast policy using imitation learning.
+Use of a curriculum allows the controller to build up from flat ground to a challenging combination of stepping stones at various heights, distances, and orientations.
+In contrast to prior work in RL, our controller \emph{plans its footstep locations}, which allows it to handle larger and more frequent obstacles.
+In contrast to prior work in footstep planning, our controller learns a dynamic policy that requires no optimization at execution time.
+
+We evaluate this algorithm on a simulated Darwin humanoid robot.
+Of course, applying the resulting policy to a real-world system requires overcoming the ``reality gap''.
+Nonetheless, recent advances in sim-to-real techniques \citep{tan2016simulation, tan2018sim} make us optimistic that CIRL will be a valuable technique in robotic biped locomotion.
+
+\section{Related work}
+
+\subsection{Step planning}
+
+Many approaches to locomotion on physical robots use optimization-based planning with detailed models.
+These approaches tend to generate \emph{hierarchical} plans in order to reduce computational requirements.
+First, they choose footstep locations using search, like A*.
+Based on the footstep plan, they generate a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum \citep{mordatch2010robust} or contact wrench cone \citep{dai2016planning}.
+Then they use inverse kinematics to plan trajectories for end-effectors \citep{zucker2010optimization}.
+Finally, these trajectories are optimized using the full dynamics model to account for obstacles and other details \citep{ratliff2009chomp}.
+
+These approaches tend to be reliable but slow \nhatch{how slow? find an example / citation}, because static walking is easier to model and optimization takes a lot of computation time.
+They also require a lot of domain knowledge and manual tuning in order to engineer good models and handle situations when the real world deviates from ideal physics.
+For now, they are the best known approaches to robotic locomotion, given the supreme importance of safety and reliability for large physical systems and high-stakes competitions such as the DARPA Robotics and Learning Locomotion Challenges.
+However, we are interested in exploring alternative possibilities that can exploit the dynamics of the system to achieve faster and more energy-efficient gaits.
+
+\nhatch{I think Honda does a lot with dynamic gaits, but I don't really understand how it works.
+I need to find and read some papers.
+To what extent can they plan step locations and handle rough terrain?}
+
+There are a few prior works that use reinforcement learning to do step planning for simulated systems.
+\cite{peng2017deeploco} use a hierarchical reinforcement learning algorithm where the upper-level controller chooses footstep locations and the lower-level controller tries to execute those plans.
+However, the target step locations are unrealistic and rarely actually achieved.
+The path that the agent follows is perfectly flat; on rough terrain, such inaccurate step planning would probably fail.
+\cite{karpathy2012curriculum} use curriculum learning for a simple planar biped to collect a set of (state, action, resulting step distance) tuples.
+However, the final policy is based on nearest-neighbor search among this set of collected experience.
+It would be challenging to apply this approach to a higher-dimensional, more complicated model.
+
+\subsection{Deep reinforcement learning}
+
+Most of the prior work on deep RL for biped locomotion over rough terrain does not explicitly plan individual footstep locations.
+Examples of this are \cite{peng2018deepmimic, heess2017emergence, peng2016terrain}. The resulting policies travel forward quickly, avoid certain obstacles, and are robust to small missteps.
+However, they have not been proven to work in situations with dense or large obstacles where small missteps can be catastrophic.
+Such environments require accurate predictive planning at the level of individual contacts.
+
+\section{Our approach}
+
+In contrast to previous work, our approach requires no optimization during policy execution, but it is still able to land the swing heel at a prespecified target location.
+For now, we assume that perception is solved\footnote{In the experiments, we evaluate the robustness of our algorithm with respect to errors in the perception system.} and that the step plan (in the form of a list of target step locations) is given by an oracle.
+The problem is to choose motor torques to accomplish the given step plan.
+
+Our algorithm is called CIRL, which stands for ``curriculum, imitation, and reinforcement learning''.
+It consists of four components: a basic parametrized walking controller, a higher-level controller that outputs parameters for the basic controller, a data collection scheme for single-step targets according to a gradually more difficult curriculum, and an optimization algorithm to solve those single-step targets.
+\nhatch{This system is rather complicated, and probably some of these components are unnecessary or unwise to include in the current paper.
+A major focus of work in August will be ablation studies to determine which of these ideas are most important.}
+
+\subsection{Basic walking controller}
+
+In order to simplify the learning problem as much as possible, we engineer and manually tune a walking controller that, with no learning, can walk forward and to some extent control the length of its steps.
+While this controller could in principle be learned using RL techniques, our approach has several advantages.
+First, the basic walking gait looks fairly natural, in contrast to most results from deep RL.
+Second, it has an interpretable set of parameters which provide a convenient interface for another, higher-level controller.
+Third, because the basic controller already produces reasonable results, the higher-level controller can learn a residual rather than an entire policy.
+This allows us to use simpler, more sample-efficient function approximators, and it reduces the computational complexity of the learning algorithm.
+
+Our basic controller is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
+SIMBICON uses a finite state machine (FSM) with two target poses: an UP state that ends after 0.3 seconds, and a DOWN state that ends on swing foot contact.
+During each state, proportional derivative (PD) controllers attempt to reach the corresponding target pose.
+Balance feedback is achieved by adjusting the swing hip angles at each time step based on the distance and velocity of the agent's CoM relative to its stance foot contact location.
+After hand-tuning the balance feedback gains and the joint angles for each state, this simple control scheme can walk forward and maintain its balance.
+
+We make several novel modifications to SIMBICON in order to control its step length:
+\nhatch{Might want to put some of this in the appendix instead. Also, using some equations might make the writing more precise.}
+\begin{enumerate}
+  \item At the beginning of each step, set the swing hip, swing knee, stance knee, and stance ankle angles of the UP state as a linear function of the distance to the next step target.
+    The coefficients of this function were manually tuned such that the resulting step distance approximately matches the target distance.
+
+  \item End the UP state only when both the swing heel and the CoM are within a certain distance of the target step location.
+    This distance is proportional to the CoM speed, where the proportionality constant (gain) is an additional parameter.
+    (If the swing heel hits the ground early, the step immediately ends.)
+    Getting the swing heel close enough to the target is important to ensure that the DOWN state accurately hits the target.
+    Getting the CoM close enough is important to ensure that the heel strike will propel the robot forward rather than backward.
+    (This is especially important for long steps.)
+
+  \item Automatically set the swing leg pose in the DOWN state using inverse kinematics (IK).
+    Again, adjust the target swing heel location for the purposes of IK from the true target location based on the CoM speed and an additional gain parameter.
+
+  \item Control the swing ankle angle in world space, rather than relative to the shin.
+    This does not have much effect on the step distance, but it helps make the dynamics of each impact more predictable, which is important for accurate function approximation.
+    \nhatch{Again, all of this will have to be validated with ablations\dots.}
+\end{enumerate}
+
+%This choice simplifies the dynamics that the lower level needs to model, because each step involves no (or very few) changes of contact.
+
+\subsection{Higher-level controller}
+
+The basic controller can walk and balance on flat ground, and to some extent control its step length.
+However, it cannot handle steps and inclines, and when the step distance is irregular, it cannot hit the targets with high accuracy.
+\nhatch{Get some numbers about how big the steps/inclines have to be before problems happen, how bad the accuracy is, etc.}
+Hence, we train a higher-level controller (HLC) on top of the basic controller.
+It is queried once at the beginning of each step.
+It has the same input (``state'') as the basic controller: joint angles and velocities, stance heel location, and target step location (46 dimensions in 3D, 22 dimensions in 2D).
+\nhatch{The basic controller also observes the robot's center of mass. Not sure where to mention this, or whether it will actually be necessary in the final version.}
+Its output (``action'') is a vector of parameters that modify the behavior of the basic controller during the subsequent step (22 dimensions in 3D, 18 dimensions in 2D; see Table \ref{table:params}).
+These parameters are the same parameters that were hand-tuned during the initial design of the basic controller; the HLC output corresponds to a residual on top of the hand-tuned values.
+We exploit bilateral symmetry by mirroring the joint states if necessary so that the swing foot is always the right foot.
+\nhatch{For 3D we'll have to mirror the targets as well; I don't think I've implemented this yet.}
+
+\begin{table}
+  \caption{Parameters of the basic walking controller for the UP phase.
+  The DOWN phase has the same set of eleven parameters (for 22 parameters total) except that \texttt{ik\_gain} now refers to the adjustment of the swing heel target used for IK.
+  \nhatch{For many of these parameters, it is reasonable to set the same values for both the UP and DOWN phases.
+We might be able to reduce problem dimensionality by removing the mostly-redundant parameters.}}
+  \label{table:params}
+  \hrule
+  \begin{tabular}{ll}
+    \texttt{torso\_world} & Torso angle in world reference frame. \\
+    \texttt{swing\_hip\_world} & Swing hip angle (sagittal plane) in world frame. \\
+    \texttt{swing\_knee\_relative} & Swing knee angle relative to thigh.\\
+    \texttt{swing\_ankle\_relative} & Swing ankle angle relative to thigh. \nhatch{Actually this is in world frame too?}\\
+    \texttt{stance\_knee\_relative} & Stance knee angle relative to thigh.\\
+    \texttt{stance\_ankle\_relative} & Stance ankle angle relative to shin.\\
+    \texttt{position\_balance\_gain} & Gain for sagittal swing hip angle based on position.\\
+    \texttt{position\_balance\_gain\_lat} & Same as above for the coronal angle.\\
+    \texttt{velocity\_balance\_gain} & Gain for sagittal swing hip angle based on velocity.\\
+    \texttt{velocity\_balance\_gain\_lat} & Same as above for the coronal angle.\\
+    \texttt{ik\_gain} & Gain for distance threshold that triggers the end of the UP phase.
+  \end{tabular}
+  \hrule
+\end{table}
+
+We implement the HLC by training a parametrized function approximator on a set of collected experience.
+This dataset is a set of (state, action, actual achieved footstep location) tuples, collected as detailed in Sec. \ref{sec:curriculum} and \ref{sec:optimization}.
+Note however that we want the HLC to output an action.
+Hence, for the regression task, the input is the pair (state, actual achieved footstep location) and the output is the action.
+Note that the input includes two ``target inputs'': the target swing heelstrike location, which was originally used when collecting the sample, and the actual swing heelstrike location, which was where the heel actually hit the ground.
+When using the HLC to prescribe actions given a new state and target, we use the same target for both of the HLC's ``target inputs''.
+\nhatch{Now that I write this out, that seems pretty unclean.
+I should probably make the ``target used when collecting the sample'' part of the \emph{action} instead.
+Unfortunately, that puts us in a situation where the HLC is setting ``targets'' for the basic controller that are not actually the targets that it wants the basic controller to hit.
+But I guess it means we could maybe remove the \texttt{ik\_gain} parameters.}
+
+For this project, we chose a linear hypothesis class for the HLC.
+\nhatch{Soon, I want to experiment with more complicated hypothesis classes.}
+Because our dataset is relatively small, we use ridge regression to avoid overfitting.
+Our dataset also tends to contain outliers, because the basic controller can behave unpredictably in some situations (such as when the agent stubs its toe).
+We handle this by using RANSAC \citep{fischler1981random} to discard outliers at training time.
+
+\subsection{Curriculum} \label{sec:curriculum}
+
+We collect the dataset in an iterative, aggregate way, similar to DAgger \citep{ross2011reduction}.
+First, we collect an initial dataset by running the basic controller for several steps on flat ground with a variety of different target step lengths.
+On subsequent rounds of data collection, we proceed as follows:
+\nhatch{Format this as an algorithm.}
+\begin{enumerate}
+  \item Choose $N$ starting states from the existing dataset.
+  \item For each starting state, choose $M$ target locations for the next step.
+  \item For each target location, find an action to achieve that target using random optimization (see Sec. \ref{sec:optimization}).
+  \item For each of $P$ Gaussian perturbations of the optimal action from step 3, run the simulation and add the sample to the dataset.
+\end{enumerate}
+For our experiments, we used $N=128$, $M=1$, $P=4$.
+The optimization in step 3 is rather expensive, hence the relatively small number of samples $NMP$ collected at each round.
+\nhatch{I'm not sure that the Gaussian perturbations are actually helpful. If they are, it might make sense to perturb the starting state as well, not just the action.}
+
+Note that the starting states for each round are chosen from the set of ending states collected during all previous rounds.
+Hence, the input distribution of the dataset is indeed representative of the distribution of states likely to be seen by the HLC at test time.
+
+Finally, we incorporate an element of curriculum learning into the data collection process.
+At certain manually specified points in the process, we change the distribution of targets used in step 2 and/or change some details of the simulation.
+For instance, at first, the targets are all chosen to lie in the ground plane, and the variance in their distance from one another is relatively small.
+Once the HLC reaches a desired level of accuracy, the variance is increased.
+After that, we change the environment such that the ground is no longer a single flat sheet, but rather a series of stepping stones with gaps in between them.
+Later still, we introduce some vertical variation into the target step locations.\footnote{The stepping stones are shifted up or down to match the target location.}.
+\nhatch{And so forth. I haven't systematically studied this yet to see how difficult we can get.}
+
+We believe \nhatch{and test experimentally} that this curriculum of gradually increasing difficulty is key to good performance.
+In our algorithm, the initial point of the optimization problem in step 3 is the action suggested by the current HLC.
+The problem is nonconvex, so this initialization point matters a lot.
+With better initialization, the saddle point is more likely to be a good local optimum, and the optimization problem might be more quickly solvable.
+Hence, it makes sense to scale the difficulty of the optimization problem roughly proportionally to the current capabilities of the HLC.
+
+\subsection{Single-step optimization} \label{sec:optimization}
+
+We now give more details about the optimization problem in step 3 of Sec. \ref{sec:curriculum}.
+\nhatch{Spoiler: It is not actually solved with reinforcement learning, so I should probably remove ``reinforcement learning'' from the title of the algorithm.}
+The objective of the problem is the Euclidean distance between the target and achieved step locations.
+We define the achieved step location as the location of the swing heel at the moment that the step ends, which is the moment that any part of the swing foot makes contact with the ground.\footnote{The step also ends if the robot crashes; i.e.
+falls off of the edge of a stepping stone, falls over, etc.}
+The optimization variables are the parameters of the basic controller (see Table \ref{table:params}).
+There are no constraints.
+\nhatch{Actually, we constrain most of the parameters to be zero.
+In my experiments so far, only seven of the 22 parameters are allowed to vary.
+It might be possible to gradually allow more parameters to be controlled according to the requirements of the curriculum.}
+
+Because it would be difficult to calculate gradients for this objective, we use the zero-order optimization algorithm described in \cite{mania2018simple}.
+At each iteration, we estimate gradients as a weighted average of $Q = 4$ random directions, with the weight for each direction chosen as the central difference of the objective value along that direction.
+Hence, we perform $2Q$ single-step rollouts for each update of the optimization variables.
+We end the optimization once the accuracy reaches two centimeters, or once a certain maximum number of iterations (in our case, ten) is reached.
 
@@
 
+We perform experiments with a series of three robot models of increasing realism: (1) a simple planar biped, (2) a simple 3D biped, and (3) a model of the Darwin robot.
+All experiments are performed using the DART simulator at a simulation rate of 2000 Hz.
+
+\nhatch{This section is still pretty rough; sorry.}
 Our experiments are designed to test two hypotheses.
+Our primary hypothesis is that this combination of imitation and curriculum learning can learn to navigate a variety of complicated locomotion environments, learning faster and achieving better results than other reinforcement learning methods.
 
-Our primary hypothesis is that this hierarchical architecture can navigate a variety of complicated locomotion environments, learning faster and achieving better results than other reinforcement learning methods.
-These locomotion environments include flat ground, inclines, stairs, random inclines, stepping stones, hurdles, and combinations of the above.
+We collect both quantitative and qualitative metrics to measure ``achieving better results''.
+Quantitatively, we evaluate the robot by presenting it with a series of small stepping stones placed at a random $(x,y,z)$ offsets from each other.
+\nhatch{It would be nice to be able to place them at different rotations/orientations as well.}
+We score the robot based on a combination of the number of successful steps before crashing and the accuracy of foot placement at each step.
+We collect this score for a number of different distributions of the $(x,y,z)$ offsets (of increasing difficulty), and find \nhatch{hopefully} that our model achieves better performance than deep RL methods on most of these environments.
+\nhatch{We will have to think carefully about how to make this a fair comparison, since most deep RL methods don't actually care about foot placement.
+For one thing, the action space for the deep RL methods will also have to be ``parameter settings for the basic walking controller''.}
+We also consider the robustness of the model to inaccurate perception by perturbing the target step locations with respect to the actual location of the stepping stones.
+\nhatch{We should probably also evaluate robustness to external force perturbations etc., but I'm nervous that that won't end well for us.
+We don't really have good feedback mechanisms \emph{during} each footstep.}
+Qualitatively, we document motion quality by means of supplementary videos, which we encourage the reader to consult.
 
@@
 
-We collect both quantitative and qualitative metrics to measure ``achieving better results''.
-Quantitatively, we compare the average rate of forward progress.
-We also consider the robustness of the model to force perturbations and various measures of the ``difficulty'' of the terrain (such as stair height).
-\footnote{Another interesting question would be robustness with respect to accuracy of perception.
-If the potential step target locations / surface normals are perturbed, what happens?}
-Qualitatively, we document motion quality by means of supplementary videos, which we encourage the reader to consult.
-
 Our secondary hypothesis is that the learned policy can be run in real-time while still achieving results comparable to more expensive, model-based optimization methods.
-
 This hypothesis is difficult to test, because model-based optimization methods are difficult to implement for direct comparison.
@@
 
-Additionally, we conduct ablative analysis of the importance of various components of our approach.
-\nhatch{In particular, we want to consider various different HLC / LLC interfaces that we could have used.
-For example: should we include a target foot height that the LLC should reach?
-(I think this would help when going over hurdles.)
-For another example (assuming we use some kind of PD control at the lowest level): how much does performance degrade if we don't use PD control?
-(That is, if the LLC outputs raw torques.)}
+Finally, we conduct ablative analysis of the importance of various components of our approach.
+\nhatch{This is super important given the current complexity of the system. We need to remove some of these elements to make a more focused paper, so we need to identify what matters and what doesn't.
+For reference, a list of the various components that we want to pare down: the many modifications to Simbicon, the complexity of the curriculum (and whether a curriculum is necessary at all), the hypothesis class for the HLC, the data collection algorithm from Sec.
+\ref{sec:curriculum} (e.g.
+do Gaussian perturbations help), the set of parameters that are controllable by the HLC, whether RANSAC is necessary, whether the basic controller observes the CoM or not.}
 
-\section{Policy architecture}
+\section{Discussion}
 
-\nhatch{Ideally, maybe we should design this so that walking and running can both be achieved in the same episode.}
-
-\nhatch{Should there be some ensemble of low-level policy networks that the HLC can choose between?
-For example, we could take the policy with the highest estimated value given the target step location.
-Having some discrete choices like this might be easier than using solely continuous parameters.}
-
-\textbf{High-level observations.}
-We assume that perception is solved.
-The high-level controller receives proprioceptive information and environmental information.
-The environmental information consists of a list of potential surfaces to step on, specified in terms of relative displacement from the agent and surface normal.
-\nhatch{Maybe, for polygonal surfaces, we could also report the surface ``size'' (e.g. largest-radius circle that fits inside the polygon).}
-
-\textbf{High-level actions.}
-High-level actions are chosen on each heel-down event (once per footstep).
-At these events, the HLC outputs a target for the next footstep that it expects the LLC can achieve.
-This target consists of several parameters: \nhatch{TODO think about this some more}
-\nhatch{It might be possible to incorporate some optimization / search / planning into the HLC, because its actions happen infrequently.
-But this would require some objective to be optimized, which might involve implementing a model like a spring-loaded inverted pendulum (SLIP), or inventing some heuristic objective that might characterize good plans.
-We might want to implement a model like SLIP anyway, in order to use it as an expert to train a neural network for the HLC (similar to AlphaGo Zero and ExIt).
-In any case, only the first step of the plan would be revealed to the LLC.}
-
-\textbf{Low-level observations.}
-The low-level controller receives proprioceptive information and the target location for the next footstep.
-Unlike the HLC, it does not observe the external environment.
-
-\textbf{Low-level actions.}
-On the one hand, we could have low-level actions being the raw torques.
-On the other hand, this might make the learning problem harder.
-Instead, we could specify ``abstract actions'' and use PD controllers or other analytical formulas to derive the actual torques.
-For instance, the output could be a target pose or end-effector position which is chosen more infrequently than the simulation framerate.
-Then we could solve an inverse kinematics or inverse dynamics problem to calculate a target pose, which is then achieved using PD controllers.
-\nhatch{We probably don't need all of this machinery.
-For the initial proof of concept, it might make sense to see how we can do using just raw torques.}
-\nhatch{TODO think about this some more.}
-
-\textbf{PD control.}
-\nhatch{We might have to rethink this abstraction layer once the ground includes inclines.
-For instance, the PD controller might now have to handle different target ankle angles.}
-
-\textbf{Miscellaneous ideas.}
-\begin{itemize}
-  \item Could the HLC-LLC interface be in terms of a \emph{force} that the LLC should try to generate on the next foot contact?
-  \item How can we design the LLC to generate good motion quality (e.g. low energy, low angular momentum)?
-\end{itemize}
-
-\section{Timeline of work}
-
-The goal is to submit this to ICRA on September 15.
-
-\begin{itemize}
-  \item Early June (2 weeks): Proof-of-concept: train a 2D walker low-level controller to follow an irregular step plan on flat ground, including some low-level PD control if necessary.
-  \item Late June (2 weeks): Train low-level controller on stepping stones.
-    Train high-level controller on flat ground, then on stepping stones.
-    Compare performance on stepping stones to baseline deep RL approach (TRPO).
-    If we can't get this to work (better than TRPO), we should consider abandoning the project.
-  \item July (4 weeks): Implement and train on inclines, stairs, and hurdles.
-    Perform ablative analyses.
-  \item Early August (2 weeks): I (Nathan) will be on vacation.
-  \item Late August and early September (4 weeks): Last-minute experiments.
-    Write and revise the paper.
-  \item September 15: ICRA deadline.
-\end{itemize}
-
-\section{Literature review}
-
-\subsection*{Locomotion on varied terrain}
-
-\begin{itemize}
-  \item \cite{peng2016terrain}: Perhaps the most closely related prior work, this paper trains planar characters to run across varied terrain.
-    In the discussion section of this paper, they mention three specific directions for extension that essentially exactly define our current project: move to three dimensions, consider more difficult terrains (smaller, more frequent obstacles) that require footstep planning, and use curriculum learning.
-
-  \item \cite{zucker2010optimization} is a project from the DARPA Learning Locomotion challenge that uses optimization to generate hierarchical motion plans and recover from errors in real time.
-    It has been tested on a physical robot with \href{https://www.youtube.com/watch?v=KpqFebvRLeQ}{impressive results}.
-    We will need to compare our work carefully with this project in the final paper, and we will need to improve upon their results one way or another.
-    The biggest difference might be in robot morphology / stability: executing step plans on bipeds is perhaps more difficult than executing step plans on quadrupeds given the small base of support during each step.
-    Other avenues for improvement might be eliminating the need for 45 seconds of initial planning, traveling faster (running?), or using some kind of interesting new technique (e.g.
-reinforcement learning or deep learning).
-    It is also worth considering whether our algorithm could work on a physical robot, though for the initial work we will unfortunately be working purely in simulation.
-
-  \item \cite{heess2017emergence} trains policies that can, to some extent, handle bumpy ground (see \href{https://www.youtube.com/watch?v=hx_bgoTF7bs&feature=youtu.be&t=1m28s}{this video}).
-    However, the agents here do not actually observe the bumps.
-    They assume flat ground and depend on policy robustness to handle unexpected terrain variation.
-    They do not report quantitative results.
-    This approach is likely limited in the kinds of terrain that it can handle.
-    In the video, the humanoid trips.
-
-  \item \cite{peng2017deeploco}
-    learns locomotion in an environment that includes bumpy ground, but the path they follow is perfectly flat.
-    They also use motion capture to train the LLC, and the steps are all of uniform length (0.5 seconds), which we hope to avoid.
-
-  \item \cite{peng2018deepmimic} includes some tasks that involve running across varied terrain.
-    They also use a form of curriculum learning, where the policies are first trained on flat ground (and without a vision component), before training then continues on the full problem.
-    Their policies are somewhat hierarchical, in the sense that they output a target pose, which is then used as input for a PD controller to generate torques.
-    Qualitatively, the policies on these tasks seem to mostly ignore the obstacles, relying on being able to recover from the perturbations generated by collisions.
-    In contrast, we would like to avoid the kind of stumbling seen in these videos by introducing another level of hierarchy that generates an explicit footstep plan.
-
-  \item \cite{peng2017learning} uses bumpy ground to test robustness of learned policies, but again the bumps are not actually perceived and cannot be planned around.
-    The bump height thus must remain small relative to the agent.
-    \nhatch{We should compare this quantitatively, later.} Also, this is 2D.
-
-  \item \cite{manchester2011stable} This paper stands out in that they derive a provably stable controller that they test on a physical robot.
-    \nhatch{I haven't read it in detail yet, though.}
-
-\end{itemize}
-
-\subsection*{Hierarchical reinforcement learning}
-
-\begin{itemize}
-  \item \cite{heess2016learning} discusses using hierarchical transfer learning to improve exploration in new environments, but this is useful mostly for sparse reward environments.
-    In contrast, we are not interested in addressing sparse rewards, but rather in (qualitative) performance on the difficult locomotion task.
-    That is, we should feel free to make the rewards as dense and as shaped as we want.
-    The reward signal is a tool, not something given to us a priori.
-
-  \item \cite{hausman2018learning} is an interesting theoretical paper showing how to get diverse meta-controls without imposing much structure on the interface between the HLC and the LLC.
-    However, in our case, we have a prior idea about a good hierarchical breakdown of the tasks (namely, a footstep plan).
-    \nhatch{I had some trouble with this paper; I should probably reread it and update this summary.}
-
-  \item \cite{daume18ilrl} consider using hierarchical policies to improve the query efficiency of interactive imitation learning.
-    In contrast, our project does not make use of an expert.
-    Even if we do end up using motion capture, such an expert is not hierarchical, and it does not specify actual actions, just target poses.
-    However, perhaps for some of the tasks in our curriculum, we will use a hand-coded expert for the HLC in order to pretrain the LLC.
-
-  \item \cite{vezhnevets2017feudal} \emph{learns} diverse low-level policies using a latent state space and directional goals.
-    However, we would like a more explicit hierarchical interface in the form of a footstep plan, and our low-level action space is continuous rather than discrete.
-
-  \item \cite{frans2018meta} uses HRL for transfer learning, but they do not apply their results to curriculum learning.
-    Hence, the tasks they accomplish remain fairly simple (simple maze traversal, etc.).
-
-  \item \cite{kulkarni2016hierarchical} uses hierarchical RL to encourage exploration in sparse-reward environments.
-    They also only consider discrete action spaces; perhaps we don't need to cite this paper at all.
-  \item \cite{nachum2018data}: This paper tries to do off-policy hierarchical RL by changing the actions from the transitions stored in the high-level replay buffer.
-    This seems like a hacky way to do things, and their experiments are clearly gaming the system to make their method look good.
-    I don't think we should cite this.
-\end{itemize}
-
-\subsection*{Curriculum and transfer learning for locomotion}
-
-\begin{itemize}
-  \item \cite{karpathy2012curriculum} use a two-level hand-designed curriculum to achieve impressive acrobatic results for a low-dimenisonal hopper.
-    However, their method involves random search and nonparametric nearest-neighbor policies, which are unlikely to generalize to higher-dimensional morphologies.
-
-  \item The aforementioned paper \cite{heess2017emergence} trains their policies using curriculum learning: the environment gets progressively more difficult as time goes on.
-    This achieves great results, but they do not quite succeed on bumpy ground.
-    In this project, we hope to explore hierarchical policies as a potential solution to this difficult task.
-
-  \item As mentioned before, \cite{peng2018deepmimic} uses some curriculum learning to accomplish the ``varied terrain'' tasks.
-
-  \item As mentioned before, \cite{frans2018meta} has the explicit goal of using HRL for transfer learning.
-
-  \item Relay Nets \citep{kumar2017relay} are a form of policy switching using the value function to choose between an ensemble of policies trained on different starting distributions, which uses a form of curriculum learning to train policies sequentially.
-    In contrast, we wish to train a single policy while varying the environment itself.
-
-\end{itemize}
-
-\subsection*{Locomotion policy architectures}
-
-\begin{itemize}
-  \item At the lowest level, we will probably be using some kind of stable PD controller, in order to make the reinforcement learning problem as easy as possible.
-    \nhatch{I need to read these papers:} \cite{tan2011stable}, \cite{shrivastava2013stable}
-  \item \cite{peng2015terrain} introduces a very interesting way of parametrizing the action space in terms of reference points for PD controllers, as well as ``virtual forces''.
-    We will want to do something similar.
-  \item We might want to use something simple, like a linear controller as proposed by \cite{mania2018simple} to see if that works well.
-    It seems to do a good job on OpenAI benchmarks, at least.
-  \item Otherwise, we'll probably end up using TRPO or PPO or something (get citation later\dots)
-  \item Other things to consider: prioritized experience replay, hindsight experience replay, universal value function approximators.
-\end{itemize}
-
-\subsection*{Model-based locomotion}
-
-\begin{itemize}
-  \item \cite{de2010feature} introduce a method for specifying locomotion behaviors with high-level features.
-    From these features, low-level torques can be derived automatically by solving a series of quadratic programs based on physical models.
-    This method is somewhat slow (50\% real time), and we might be able to avoid using hand-engineered models by employing deep networks.
-
-  \item \cite{mordatch2010robust} use \cite{de2010feature} to execute a footstep plan which is generated using a lower-dimensional simplified physics model (a spring-loaded inverted pendulum).
-    We might be able to use such an optimization- / model-based method as an expert, if we want to employ some kind of imitation learning.
-    This could be similar to the approach of AlphaGo Zero.
-
-  \item \cite{mordatch2012discovery} handles contacts in a clever way, using piecewise-constant, smooth contact variables and a continuation scheme.
-    This allows complex motions to emerge from a quadratic optimization procedure.
-    However, this method is very slow (2\% real time) and it generates only static motion (no jumping or slipping).
-    Again, maybe we could use it as an expert.
-
-\end{itemize}
-
-\section{Project summary}
-
+We see at least two important directions for future work.
 \begin{enumerate}
-
-\item \textbf{One-sentence summary:}
-Learn a difficult locomotion task with curriculum learning, hierarchical policies, and possibly model-based planning.
-
-\item \textbf{Task:}
-Train a simulated robotic agent to accomplish locomotion in a difficult environment, such as walking (or preferably running) across bumpy ground.
-
-\item \textbf{Introduction:} Robotic locomotion on uneven terrain is an important problem.
-  \nhatch{Visak, I would be interested in your high-level reasons for being interested in this problem.
-  I'm not sure I can motivate the problem as well as you can.} Previous approaches have not solved this problem.
-  For instance, in \href{https://www.youtube.com/watch?v=hx_bgoTF7bs&feature=youtu.be&t=1m28s}{this video} the agent stumbles on rubble, and DeepLoco only walks on flat terrain.
-  \nhatch{More examples would help, to make sure that this problem is interesting.} We hope to achieve efficient locomotion across bumpy ground using a combination of three techniques.
-  \begin{enumerate}
-    \item Curriculum learning: Train the agent first on simpler environments, like flat ground and gradual slopes, and gradually increase the difficulty.
-    \item Hierarchical controllers: We use a hierarchical architecture where the HLC plans foot placement and the LLC executes the individual steps.
-      We hypothesize that this architecture will more easily handle the discontinuities caused by contact dynamics.
-      We also hypothesize that, with this architecture, it will be easier to transfer knowledge between tasks in the curriculum.
-    \item Model-based planning: To help with sample efficiency, the HLC has access to a physics model for the environment.
-      \nhatch{I'm not sure how to accomplish this.
-      Physics models for legged robots are probably hard, especially when the ground is not flat.
-      Maybe some kind of \emph{learned} model would work.}
+  \item \emph{Sim-to-real:}
+    We have successfully applied CIRL to a realistic simulated model, but it is likely that more work is required to transfer the resulting policy to the real world.
+  \item \emph{Motion capture:}
+    A significant portion of the effort for this project went into developing a reasonable baseline walking policy.
+    It might be possible to avoid this by bootstrapping the imitation learning process with motion capture data.
+    This would also make it easier to to extend CIRL to other robotic motion problems, such as manipulation.
 \end{enumerate}
 
-\item \textbf{Controller architecture:}
-  Like DeepLoco, the HLC receives detailed environmental observations, and outputs the target locations for the next couple of foot placements.
-  The LLC takes this input to choose the low-level actions that will achieve those foot placements.
-  Unlike DeepLoco, the terrain is much more complicated, so it is important that the HLC foot placements are actually feasible.
-  Furthermore, we will train the LLC without using motion captures, and we will permit the time per step to vary, rather than requiring each step to take exactly 0.5 seconds.
-
-  The parametrization of the HLC and LLC will probably end up being some kind of neural network, although it might be worth trying simpler policies first (e.g. linear).
-
-  To train these controllers, we could first train the LLC on simpler terrain, where we can specify target foot placement using a simple calculation.
-  When training the HLC, to ensure that it chooses feasible foot placements, we could include some additional loss term in the training objective based on whether the LLC actually achieved the foot placement that the HLC suggested.
-
-\nhatch{This section could probably use more detail; suggestions are welcome.
-For example, do we need to modify this architecture to better handle contact dynamics?
-Is there some way we can incorporate a model to help the HLC make better plans?}
-
-\item \textbf{Why this might be a good idea:}
-I have not seen previous papers succeed at this task, so if we could show some nice videos of a quadruped or biped running along bumpy ground, it would catch people's attention.
-Intuitively, when running along bumpy hiking trails, I am consciously thinking about where to place my feet, then relying on unconsciously being able to place my foot there while I think about the next step.
-This suggests a kind of hierarchical policy.
-
-\item \textbf{Why this might not be a good idea:}
-  \nhatch{Why would *hierarchical* RL be a particularly good way to do transfer learning for this series of tasks?}
-  We will need to implement several other baselines to compare against.
-  \nhatch{What should those baselines be? (What other approaches have people used for locomotion on bumpy ground?)
-Is the ``bumpy ground'' task too difficult? (Byron thinks so.)
-Will the simulated robot need toes?}
-I haven't worked with locomotion before, so I don't have a good sense for this.
-
+\nhatch{Miscellaneous TODO notes for me:
+\begin{enumerate}
+  \item Set damping coefficients for the DART simulation. Visak says this will help prevent numerical explosions during collisions.
+  \item Get Simbicon3D to be able to roughly control step distances, similar to 2D Simbicon. Or maybe this isn't necessary? Maybe we could jump straight to the learning algorithm?
+  \item Conduct ablation analysis to try to reduce the number of ``new ideas'' presented in this paper.
+  \item Think about how best to compare this algorithm to existing work (both RL work and traditional optimization-based step planning).
 \end{enumerate}
+}
 

### 2018-08-03 source/consts_3D.py
Commit some commented code to sort of document what I was working on

@@ -5,10 +5,13 @@ RIGHT_IDX = 9
 LEFT_IDX = 15
 Q_DIM = 21
 THETA_IDX = 3
+
 HIP_OFFSET = 0
+HIP_OFFSET_TWIST = 1
 HIP_OFFSET_LAT = 2
 KNEE_OFFSET = 3
 ANKLE_OFFSET = 4
+ANKLE_OFFSET_LAT = 5
 
 PELVIS_BODYNODE_IDX = 2
 RIGHT_BODYNODE_IDX = 5

### 2018-08-03 source/inverse_dynamics.py

@@ -29,7 +29,7 @@ class LearnInverseDynamics:
         self.n_action = env.action_space.shape[0]
         # The actual target is sort of part of the state, hence the *2.
         self.n_dynamic = 26
-        # Use a linear policy for now
+        # TODO try some model more complicated than linear?
         model = Ridge(alpha=RIDGE_ALPHA, fit_intercept=False)
         #model = make_pipeline(PolynomialFeatures(2), model) # quadratic
         self.model = RANSACRegressor(base_estimator=model, residual_threshold=2.0)
@@ -115,6 +115,7 @@ class LearnInverseDynamics:
         self.train_targets.append(action)
 
     def center_state(self, state, target, achieved_target):
+        # TODO this needs to be updated for 2D -> 3D
         centered_state = np.zeros(self.n_dynamic)
 
         centered_state[ 0:18] = state.pose()
@@ -137,7 +138,6 @@ class LearnInverseDynamics:
         return centered_state
 
     def train_inverse_dynamics(self):
-        # TODO try some model more complicated than linear?
         X = np.array(self.train_states)
         y = np.array(self.train_targets)
         self.X_mean = X.mean(0)

### 2018-08-03 source/pd_control.py

@@ -16,6 +16,10 @@ class PDController:
         BRICK_DOF = c.BRICK_DOF
         self.Kp = np.array([0.0] * BRICK_DOF + [c.KP_GAIN] * (self.skel.ndofs - BRICK_DOF))
         self.Kd = np.array([0.0] * BRICK_DOF + [c.KD_GAIN] * (self.skel.ndofs - BRICK_DOF))
+        #if c.BRICK_DOF == 6:
+        #    # I think these "muscles" need to be stronger.
+        #    self.Kp[c.RIGHT_IDX + c.HIP_OFFSET_LAT] *= 4
+        #    self.Kp[c.LEFT_IDX + c.HIP_OFFSET_LAT] *= 4
         self.control_bounds = control_bounds_2D if BRICK_DOF == 3 else control_bounds_3D
 
     def compute(self):

### 2018-08-03 source/simbicon_3D.py

@@ -20,8 +20,8 @@ class Simbicon3D(Simbicon):
         return state
 
     def base_gait(self):
-        return ([0.14, 0.5, 0.2, -0.2, 0.5, -1.1, 0.6, -0.05, 0,   0.5, 0.2],
-                [0.14, 0.5, 0.2, -0.2, -0.1, -0.05, 0.15, -0.1, 0, 0.5, 0.2])
+        return ([0.14, 0.5, 0.2, -0.2,  0.5, -1.1,   0.6, -0.05,   0, 0.5, 0.2],
+                [0.14, 0.5, 0.2, -0.2, -0.1, -0.05, 0.15, -0.1, 0.0, 0.5, 0.2])
 
     def adjust_up_targets(self):
         pass
@@ -53,14 +53,22 @@ class Simbicon3D(Simbicon):
         angle = base + balance_feedback
         q[self.swing_idx+c.HIP_OFFSET_LAT] = angle
 
+        #actual_twist = self.skel.q[4] + self.skel.q[7]
+        # TODO why does the robot rotate so much on the third step of test()?
+        # Try putting dots on all the contact points -- does the robot lose contact
+        # with the stance foot (so it's a conservation of angular momentum thing)?
+        #q[self.stance_idx+c.HIP_OFFSET_TWIST] = actual_twist/2
+        #q[self.swing_idx+c.HIP_OFFSET_TWIST] = -actual_twist/2
+
         self.update_doppelganger(q)
         return q
 
     def update_doppelganger(self, q):
         c = self.env.consts()
         q = q.copy()
-        q[:c.BRICK_DOF] = self.env.robot_skeleton.q[:c.BRICK_DOF]
+        #q[:c.BRICK_DOF] = self.env.robot_skeleton.q[:c.BRICK_DOF]
         q[2] -= 0.5
+        q[0] = self.env.robot_skeleton.q[0]
         self.env.doppelganger.q = q
         self.env.doppelganger.dq = np.zeros(c.Q_DIM)
 
@@ -80,8 +88,8 @@ def test(env):
     env.seed(seed)
     env.reset(random=0.0)
     env.sdf_loader.put_dot([0,0,0])
-    for i in range(8):
-        t = 0.3 + 0.4*i# + np.random.uniform(low=-0.2, high=0.2)
+    for i in range(16):
+        t = 0.1 + 0.2*i# + np.random.uniform(low=-0.2, high=0.2)
         env.simulate([t,0,0], render=1, put_dots=False)
 
 if __name__ == "__main__":

### 2018-10-18 paper/main.tex
Tiny paper edits

@@
 
-Building on this work, we approach dynamic biped locomotion on rough terrain using a combination of curriculum learning, imitation learning, and reinforcement learning, which we call CIRL.
-\nhatch{Just a working title.}
-The controller learns new skills using RL, which are generalized to a fast policy using imitation learning.
+Building on this work, we approach dynamic biped locomotion on rough terrain using a combination of curriculum learning and imitation learning, which we call ResCu. \nhatch{Just a working title.}
+\nhatch{Should mention that the ``controller'' below is only learning a residual on top of an already-reassnable basic controller.}
+The controller learns a fast policy through imitation of an expert that uses (slow) zero-order optimization.
 Use of a curriculum allows the controller to build up from flat ground to a challenging combination of stepping stones at various heights, distances, and orientations.
-In contrast to prior work in RL, our controller \emph{plans its footstep locations}, which allows it to handle larger and more frequent obstacles.
+In contrast to prior work in learning-based locomotion, our controller \emph{plans its footstep locations}, which allows it to handle larger and more frequent obstacles.
 In contrast to prior work in footstep planning, our controller learns a dynamic policy that requires no optimization at execution time.
@@
 Of course, applying the resulting policy to a real-world system requires overcoming the ``reality gap''.
-Nonetheless, recent advances in sim-to-real techniques \citep{tan2016simulation, tan2018sim} make us optimistic that CIRL will be a valuable technique in robotic biped locomotion.
+Nonetheless, recent advances in sim-to-real techniques \citep{tan2016simulation, tan2018sim} make us optimistic that ResCu will be a valuable technique in robotic biped locomotion.
 
@@
 
-There are a few prior works that use reinforcement learning to do step planning for simulated systems.
+There are a few prior works that use learning-based approaches to do step planning for simulated systems.
 \cite{peng2017deeploco} use a hierarchical reinforcement learning algorithm where the upper-level controller chooses footstep locations and the lower-level controller tries to execute those plans.
@@
 
-Our algorithm is called CIRL, which stands for ``curriculum, imitation, and reinforcement learning''.
+Our algorithm is called ResCu, which stands for ``residual curriculum learning''.
 It consists of four components: a basic parametrized walking controller, a higher-level controller that outputs parameters for the basic controller, a data collection scheme for single-step targets according to a gradually more difficult curriculum, and an optimization algorithm to solve those single-step targets.
@@
 We exploit bilateral symmetry by mirroring the joint states if necessary so that the swing foot is always the right foot.
-\nhatch{For 3D we'll have to mirror the targets as well; I don't think I've implemented this yet.}
 
@@
 We handle this by using RANSAC \citep{fischler1981random} to discard outliers at training time.
+\nhatch{RANSAC might not be necessary if the optimization alg. to collect the dataset is good enough. Maybe that alg could throw out examples where no good parameter setting was found.}
 
@@
 The problem is nonconvex, so this initialization point matters a lot.
-With better initialization, the saddle point is more likely to be a good local optimum, and the optimization problem might be more quickly solvable.
+With better initialization, the stationary point is more likely to be a good local optimum, and the optimization problem might be more quickly solvable.
 Hence, it makes sense to scale the difficulty of the optimization problem roughly proportionally to the current capabilities of the HLC.
@@
 We now give more details about the optimization problem in step 3 of Sec. \ref{sec:curriculum}.
-\nhatch{Spoiler: It is not actually solved with reinforcement learning, so I should probably remove ``reinforcement learning'' from the title of the algorithm.}
 The objective of the problem is the Euclidean distance between the target and achieved step locations.
@@
 At each iteration, we estimate gradients as a weighted average of $Q = 4$ random directions, with the weight for each direction chosen as the central difference of the objective value along that direction.
-Hence, we perform $2Q$ single-step rollouts for each update of the optimization variables.
+Hence, we perform $2Q$ single-footstep rollouts for each update of the optimization variables.
 We end the optimization once the accuracy reaches two centimeters, or once a certain maximum number of iterations (in our case, ten) is reached.

### 2019-01-10 source/inverse_dynamics.py
Various refactors/minor fixes in pursuit of inverse_dynamics
1. Actions are now low-dimensional, rather than sparse. This is mostly
to make it easier to just print the whole list of parameters of the
model.

2. Remove RANSAC and training set perturbations (for now).

3. Only use 8 resets when collecting starting states (down from 16 to
save time).

4. Remove action retargeting. It would be too expensive to add three
action dimensions . . . and anyway it wasn't working and was hard to
understand.

5. Store the swing platform in extract_features, and don't filter to
observable features when stor

@@ -25,14 +25,16 @@ class LearnInverseDynamics:
     def __init__(self, env, exp_name=''):
         self.env = env
         self.exp_name = exp_name
+        self.video_save_dir = None
         self.initialize_start_states()
         self.train_features, self.train_responses = [], []
-        self.n_action = len(self.env.controller.base_gait())
+        self.n_action = len(self.env.controller.action_params())
         self.n_dynamic = sum(self.env.consts().observable_features)
         # TODO try some model more complicated than linear?
         model = Ridge(alpha=RIDGE_ALPHA, fit_intercept=False)
         #model = make_pipeline(PolynomialFeatures(2), model) # quadratic
-        self.model = RANSACRegressor(base_estimator=model, residual_threshold=2.0)
+        #model = RANSACRegressor(base_estimator=model, residual_threshold=2.0)
+        self.model = model
         self.X_mean = np.zeros(self.n_dynamic)
         self.y_mean = np.zeros(self.n_action)
         # Maybe try varying these.
@@ -54,7 +56,7 @@ class LearnInverseDynamics:
         fname = START_STATES_FMT.format(self.exp_name)
         if not os.path.exists(fname):
             if env.is_3D:
-                states = self.collect_starting_states(n_resets=8, min_length=0.1, max_length=0.7)
+                states = self.collect_starting_states(min_length=0.1, max_length=0.7)
             else:
                 states = self.collect_starting_states()
             with open(fname, 'wb') as f:
@@ -62,7 +64,7 @@ class LearnInverseDynamics:
         with open(fname, 'rb') as f:
             self.start_states = pickle.load(f)
 
-    def collect_starting_states(self, size=8, n_resets=16, min_length=0.2, max_length=0.8):
+    def collect_starting_states(self, size=8, n_resets=8, min_length=0.2, max_length=0.8):
         self.env.log("Collecting initial starting states")
         start_states = []
         for i in range(n_resets):
@@ -131,51 +133,49 @@ class LearnInverseDynamics:
         runner = Runner(self.env, start_state, target)
         runner.video_save_dir = self.video_save_dir
         rs = RandomSearch(runner, 4, step_size=0.1, eps=0.05)
-        render = 0.7 if self.video_save_dir else 1.0
         rs.w_policy = self.act(start_state, target) # Initialize with something reasonable
         # TODO put max_iters and tol in the object initialization params instead
[…]

### 2019-01-10 source/random_search.py

@@ -7,7 +7,7 @@ import gym
 class RandomSearch:
     def __init__(self, runner, n_dirs, step_size=0.01, eps=0.05):
         self.runner = runner
-        self.w_policy = self.runner.env.controller.controllable_indices() * 0.0
+        self.w_policy = self.runner.env.controller.action_scale() * 0.0
 
         self.n_dirs = n_dirs
         self.eps = eps
@@ -16,7 +16,6 @@ class RandomSearch:
 
     def sample_perturbation(self):
         delta = np.random.randn(np.prod(self.w_policy.shape))
-        delta *= self.runner.env.controller.controllable_indices()
         delta /= np.linalg.norm(delta)
         return delta.reshape(self.w_policy.shape)
 
@@ -46,11 +45,7 @@ class RandomSearch:
         print("Max iters exceeded")
         return None
 
-    def manual_search(self, params, next_target, next_heading, video_save_dir=None):
-        c = self.runner.env.controller
-        action = c.base_gait() * 0.0
-        inds = c.controllable_indices_list()
-        action[inds] = params * c.controllable_indices_magnitudes()
+    def manual_search(self, action, next_target, next_heading, video_save_dir=None):
         action += self.w_policy
         self.runner.reset(video_save_dir)
         env = self.runner.env

### 2019-01-10 source/simbicon.py

@@ -83,7 +83,7 @@ class Simbicon(PDController):
                 0,0,0]
         return np.array(gait)
 
-    def controllable_indices_list(self):
+    def action_params(self):
         return [
                 IK_GAIN,
                 VELOCITY_BALANCE_GAIN,
@@ -91,13 +91,8 @@ class Simbicon(PDController):
                 DN_IDX + STANCE_KNEE_RELATIVE
                 ]
 
-    def controllable_indices_magnitudes(self):
-        return np.ones(len(self.controllable_indices_list()))
-
-    def controllable_indices(self):
-        inds = self.base_gait() * 0.0
-        inds[self.controllable_indices_list()] = self.controllable_indices_magnitudes()
-        return inds
+    def action_scale(self):
+        return np.ones(len(self.action_params()))
 
     def heading(self):
         return 0.0
@@ -105,7 +100,8 @@ class Simbicon(PDController):
     def set_gait_raw(self, target, target_heading=None, raw_gait=None):
         params = self.base_gait()
         if raw_gait is not None:
-            params += raw_gait
+            inds = self.action_params()
+            params[inds] += raw_gait * self.action_scale()
         self.set_gait(Params(params))
 
         self.target, self.prev_target = target, self.target

### 2019-01-10 source/simbicon_3D.py

@@ -37,10 +37,16 @@ class Simbicon3D(Simbicon):
                 0.5, 0.2, 0.0, 0.0, 0.0, 0.0]
         return np.array(gait)
 
-    def controllable_indices_list(self):
+    def set_gait_raw(self, target, target_heading=None, raw_gait=None):
+        if raw_gait is not None and self.swing_idx == LEFT_IDX:
+            sign_switches = np.array([1.0, -1.0, 1.0, 1.0, -1.0])
+            raw_gait = raw_gait * sign_switches
+        return super().set_gait_raw(target, target_heading, raw_gait)
+
+    def action_params(self):
         return [IK_GAIN, HEADING, UP_IDX+SWING_ANKLE_RELATIVE, STANCE_ANKLE_RELATIVE, STANCE_ANKLE_ROLL]
 
-    def controllable_indices_magnitudes(self):
+    def action_scale(self):
         return np.array([0.2, 0.5, 3.0, 1.0, 1.0])
 
     def rotmatrix(self, theta):

### 2019-01-10 source/simbicon_params.py

@@ -27,4 +27,3 @@ STANCE_HIP_ROLL_EXTRA = 19
 STANCE_ANKLE_ROLL = 20
 SWING_ANKLE_ROLL = 21
 
-FLIP_Z = [TZ, HEADING, STANCE_HIP_ROLL_EXTRA, STANCE_ANKLE_ROLL, SWING_ANKLE_ROLL]

### 2019-01-10 source/step_learner.py

@@ -26,7 +26,7 @@ class Runner:
         self.env.sdf_loader.put_grounds(self.grounds)
 
 def collect_start_state(env, targets, video_save_dir):
-    action = env.controller.base_gait()*0
+    action = env.controller.action_scale()*0
     # Two short steps, then a long step. This causes issues for the basic SIMBICON
     # controller, if the next step is another short step. This is because the
     # velocity-based swing hip correction tends to overcorrect so that the robot
@@ -81,6 +81,6 @@ def test_3D(video_save_dir):
     return rs
 
 if __name__ == '__main__':
-    rs = test_2D()
-    #rs = test_3D(None)
+    #rs = test_2D()
+    rs = test_3D(None)
     embed()

### 2019-01-10 source/stepping_stones_env.py

@@ -47,6 +47,7 @@ class SteppingStonesEnv:
                 c.observable_features_q,
                 c.observable_features_t,
                 c.observable_features_t,
+                c.observable_features_t,
                 ])
         self.is_3D = (c.BRICK_DOF == 6)
 

### 2019-01-29 paper/main.tex
Small clarification of paper text
Some paper edits

@@
 \emph{Biped} locomotion is more difficult because the support polygon is small and the center of mass is relatively high.
-Biped locomotion across \emph{rough terrain} is more difficult still, because the variability in the surface normal makes the dynamics harder to model.
+Biped locomotion across \emph{rough terrain} is more difficult still, because the variability in the contact surface complicates the dynamics.
 
@@
 Deep RL is a popular technique to learn highly dynamic motions for humanoids \citep{peng2018deepmimic, heess2017emergence}.
-The resulting policies are both computationally and physically fast.
-However, their effectiveness on rough terrain is uncertain.
+Using sim-to-real techniques, \citet{tan2018sim} have even applied deep RL techniques to physical robots, albeit for quadruped locomotion rather than biped.
+The policies learned by deep RL are both computationally and physically fast.
+However, due to the extreme sample inefficiency of deep RL algorithms, their effectiveness on more challenging environments---for example, physical bipeds or rough terrain---is uncertain.
 
-Building on this work, we approach dynamic biped locomotion on rough terrain using a combination of curriculum learning and imitation learning, which we call ResCu. \nhatch{Just a working title.}
-\nhatch{Should mention that the ``controller'' below is only learning a residual on top of an already-reassnable basic controller.}
-The controller learns a fast policy through imitation of an expert that uses (slow) zero-order optimization.
-Use of a curriculum allows the controller to build up from flat ground to a challenging combination of stepping stones at various heights, distances, and orientations.
-In contrast to prior work in learning-based locomotion, our controller \emph{plans its footstep locations}, which allows it to handle larger and more frequent obstacles.
-In contrast to prior work in footstep planning, our controller learns a dynamic policy that requires no optimization at execution time.
+One way to address this is to simplify the environment by use of what \citet{iscen2018pmtg} call \emph{policies modulating trajectory generators} (PMTG).
+A trajectory generator (TG) is a policy (usually relatively simple) with several high-level parameters.
+In environments with a strong prior on what a good policy might look like (such as locomotion),
+training a high-level policy (HLP) to \emph{modulate} the parameters of a TG may be easier than learning a policy from scratch.
+The TG allows us to take advantage of prior knowledge, while the learned policy allows us to address cases where prior knowledge is insufficient.
 
-We evaluate this algorithm on a simulated Darwin humanoid robot.
-Of course, applying the resulting policy to a real-world system requires overcoming the ``reality gap''.
-Nonetheless, recent advances in sim-to-real techniques \citep{tan2016simulation, tan2018sim} make us optimistic that ResCu will be a valuable technique in robotic biped locomotion.
+Building on this work, we approach dynamic biped locomotion on rough terrain using an approach similar to PMTG that uses curriculum learning and imitation learning.
+We call this curricular imitative modulation (CIM).
+In contrast to prior work in footstep planning, our algorithm learns a dynamic policy that requires no optimization at execution time.
+In contrast to prior work in learning-based locomotion, our learned policy \emph{controls its footstep locations}, which allows it to handle larger and more frequent obstacles.
+Our main contributions are as follows:
+\begin{itemize}
+  \item A trajectory generator for controllable biped locomotion, based on SIMBICON \citep{yin2007simbicon} but with several important modifications
+  \item An algorithm combining curriculum and imitation learning to learn efficiently a policy to modulate the trajectory generator
+  \item Experiments demonstrating the success of this approach in challenging simulated environments
+\end{itemize}
 
@@
 
-These approaches tend to be reliable but slow \nhatch{how slow? find an example / citation}, because static walking is easier to model and optimization takes a lot of computation time.
+These approaches tend to be reliable but slow (four seconds per step in \citet{feng2015optimization}) because static walking is easier to model and optimization takes a lot of computation time.
 They also require a lot of domain knowledge and manual tuning in order to engineer good models and handle situations when the real world deviates from ideal physics.
@@
 \cite{karpathy2012curriculum} use curriculum learning for a simple planar biped to collect a set of (state, action, resulting step distance) tuples.
-However, the final policy is based on nearest-neighbor search among this set of collected experience.
-It would be challenging to apply this approach to a higher-dimensional, more complicated model.
+The final policy is based on nearest-neighbor search among this set of collected experience.
+However, due to the curse of dimensionality of nearest-neighbor search,
+it would be challenging to apply this approach to a more complicated model.
 
@@
 
+\subsection{Policies modulating trajectory generators}
+
+Perhaps the most closely related prior work is \citet{iscen2018pmtg}.
+Like CIM, PMTGs use a trajectory generator to simplify the environment when learning controllable robot locomotion.
+Also like CIM, the hypothesis class of the HLP is very simple, often linear.
+However, there are three important differences between CIM and PMTG.
+
+First: In CIM, the HLP and the TG can run at different frequencies.
+/nhatch{Should we use ``LLP'' instead of ``TG'' in the main body of the paper? Might make it easier to understand.}
+For example, in our experiments, the TG runs at 2000 Hz, while the HLP runs at approximately 2 Hz.
+This works because the HLP controls \emph{only} the parameters of the TG; it does not try to learn a correction term for the low-level joint torques.
+Assuming that the low-level environment dynamics between calls to the HLP are fairly predictable (or that the TG can moderate any stochasticity that does arise), this further simplifies the environment from the point of view of the HLP.
+
+Second: CIM can take advantage of a well-designed reward signal by learning based on \emph{individual state-action pairs} rather than complete episodes.
+For example, in locomotion, it is possible to design a reward function such that a high-reward action for an individual footstep ensures that the robot remains balanced, which allows it to complete the rest of the episode.
+This allows us to learn good parameters for each footstep one at a time, then use imitation learning to generalize.
+By eliminating the infamous credit assignment problem, this makes the learning task much easier for the HLP.
+In contrast, PMTG does not take advantage of the ability to design the reward function.
+
+Finally, CIM uses a curriculum of environments of increasing difficulty to further ease the learning problem.
+The overall thrust of these differences from PMTG is to simplify the high-level environment dynamics as much as possible.
+This makes CIM suitable for highly challenging environments, such as biped locomotion over rough terrain, where pure RL or random search algorithms are unlikely to succeed.
+In \citet{iscen2018pmtg}, the HLP learns to control the walking speed of a quadruped robot.
+In this paper, our HLP learns to control the precise heel placement of each footstep.\footnote{
+One other important difference is that \citet{iscen2018pmtg} test their algorithm on a physical robot, while all of our experiments are simulated.
+We believe that CIM would be useful to learn policies for physical robots as well, and we intend to investigate this in future work.}
+
 \section{Our approach}
 
-In contrast to previous work, our approach requires no optimization during policy execution, but it is still able to land the swing heel at a prespecified target location.
-For now, we assume that perception is solved\footnote{In the experiments, we evaluate the robustness of our algorithm with respect to errors in the perception system.} and that the step plan (in the form of a list of target step locations) is given by an oracle.
+To solve the problem of locomotion over rough terrain, our policy must be able to place its footsteps with high accuracy.
+Our goal is furthermore to do this with a fast, dynamic gait that requires no optimization during policy execution.
+For now, we assume that perception is solved and that the step plan (in the form of a list of target step locations) is given by an oracle.\footnote{
+In 3D, each target step location also includes a target \emph{heading}: the direction that the robot should face during the step.}
 The problem is to choose motor torques to accomplish the given step plan.
 
-Our algorithm is called ResCu, which stands for ``residual curriculum learning''.
-It consists of four components: a basic parametrized walking controller, a higher-level controller that outputs parameters for the basic controller, a data collection scheme for single-step targets according to a gradually more difficult curriculum, and an optimization algorithm to solve those single-step targets.
-\nhatch{This system is rather complicated, and probably some of these components are unnecessary or unwise to include in the current paper.
-A major focus of work in August will be ablation studies to determine which of these ideas are most important.}
+Our algorithm is called ``curricular imitative modulation'' (CIM).
+It consists of four components: a trajectory generator, a high-level policy, an expert that suggests good actions for the high-level policy, and a curriculum of environments of increasing difficulty.
 
-\subsection{Basic walking controller}
+\subsection{The trajectory generator: FSM-based PD control}
 
-In order to simplify the learning problem as much as possible, we engineer and manually tune a walking controller that, with no learning, can walk forward and to some extent control the length of its steps.
-While this controller could in principle be learned using RL techniques, our approach has several advantages.
+In order to simplify the learning problem as much as possible, we engineer and manually tune a walking controller (a.k.a.
+TG) that, with no learning, can walk forward and to some extent control the length of its steps.
+While this controller could in principle be learned using RL techniques, the manual approach has several advantages.
 First, the basic walking gait looks fairly natural, in contrast to most results from deep RL.
-Second, it has an interpretable set of parameters which provide a convenient interface for another, higher-level controller.
-Third, because the basic controller already produces reasonable results, the higher-level controller can learn a residual rather than an entire policy.
-This allows us to use simpler, more sample-efficient function approximators, and it reduces the computational complexity of the learning algorithm.
+Second, it has an interpretable set of parameters which provide a convenient interface for a high-level policy.
+Third, because the basic controller already produces reasonable results, the HLP can fine-tune those results rather than learning from scratch.
+This allows the HLP to use simpler, more sample-efficient function approximators, and it reduces the computational complexity of the learning algorithm.
 
-Our basic controller is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
-SIMBICON uses a finite state machine (FSM) with two target poses: an UP state that ends after 0.3 seconds, and a DOWN state that ends on swing foot contact.
-During each state, proportional derivative (PD) controllers attempt to reach the corresponding target pose.
-Balance feedback is achieved by adjusting the swing hip angles at each time step based on the distance and velocity of the agent's CoM relative to its stance foot contact location.
+Our TG is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
+SIMBICON uses a finite state machine (FSM) with two states: an UP state that ends after 0.3 seconds, and a DOWN state that ends on swing foot contact.
+During each state, proportional derivative (PD) controllers attempt to reach the corresponding set of target joint angles.
+Balance feedback is achieved by adjusting the target swing hip angles based on the distance and velocity of the agent's CoM relative to its stance foot contact location.
 After hand-tuning the balance feedback gains and the joint angles for each state, this simple control scheme can walk forward and maintain its balance.
 
-We make several novel modifications to SIMBICON in order to control its step length:
+We make several novel modifications to SIMBICON in order to control its footstep placement more precisely:
 \nhatch{Might want to put some of this in the appendix instead. Also, using some equations might make the writing more precise.}
 \begin{enumerate}
-  \item At the beginning of each step, set the swing hip, swing knee, stance knee, and stance ankle angles of the UP state as a linear function of the distance to the next step target.
+  \item Insert a new TOE-OFF state at the beginning of each step.
+    This state transitions to the UP state after 0.2s.
+    The TG itself does not use this state (that is, the target joint angles are the same as the UP state).
+    However, it allows the HLP to use the swing ankle to influence the robot's initial acceleration while still returing to a more neutral ankle angle before heel-strike.
+
+  \item At the beginning of each step, set the target swing hip, swing knee, stance knee, and stance ankle angles of the UP state as a linear function of the distance to the next step target.
     The coefficients of this function were manually tuned such that the resulting step distance approximately matches the target distance.
@@
 
-  \item Control the swing ankle angle in world space, rather than relative to the shin.
+  \item Make sure that the swing heel is the first part of the foot to hit the ground, by control the swing ankle angle in world space rather than relative to the shin.
     This does not have much effect on the step distance, but it helps make the dynamics of each impact more predictable, which is important for accurate function approximation.
     \nhatch{Again, all of this will have to be validated with ablations\dots.}
+
+  \item In 3D, control the robot's heading using the stance hip.
+    We found that setting the target stance hip angles using inverse kinematics was sufficient, without considering the effect of other torques on the stance hip.
 \end{enumerate}
+
+Please see the supplemental video for examples of our TG trying to hit step targets, trying to climb stairs, and walking in a circle.
 
@@
 
-\subsection{Higher-level controller}
+\subsection{High-level policy}
 
-The basic controller can walk and balance on flat ground, and to some extent control its step length.
-However, it cannot handle steps and inclines, and when the step distance is irregular, it cannot hit the targets with high accuracy.
-\nhatch{Get some numbers about how big the steps/inclines have to be before problems happen, how bad the accuracy is, etc.}
-Hence, we train a higher-level controller (HLC) on top of the basic controller.
+The TG can walk and balance on flat ground, and to some extent control its step length.
+However, it cannot hit footstep targets with high accuracy, and it usually falls when climbing stairs.
+\nhatch{See Table somewhere-in-experiments for information about how big the steps/inclines have to be before problems happen, how bad the accuracy is, etc.}
+Hence, we train a high-level policy (HLP) on top of the TG.
 It is queried once at the beginning of each step.
-It has the same input (``state'') as the basic controller: joint angles and velocities, stance heel location, and target step location (46 dimensions in 3D, 22 dimensions in 2D).
-\nhatch{The basic controller also observes the robot's center of mass. Not sure where to mention this, or whether it will actually be necessary in the final version.}
-Its output (``action'') is a vector of parameters that modify the behavior of the basic controller during the subsequent step (22 dimensions in 3D, 18 dimensions in 2D; see Table \ref{table:params}).
-These parameters are the same parameters that were hand-tuned during the initial design of the basic controller; the HLC output corresponds to a residual on top of the hand-tuned values.
-We exploit bilateral symmetry by mirroring the joint states if necessary so that the swing foot is always the right foot.
+It has the same input (``state'') as the basic controller: joint angles and velocities, stance heel location, and target step location (26 dimensions in 2D, 54 dimensions in 3D).
+In addition, it sees the two previous target footstep locations.
+(This is important in ``stepping stones'' environments where the exact location of the previous stepping stone might affect the dynamics.)
+Its output (``action'') is a vector of parameters that modify the behavior of the basic controller during the subsequent step (4 dimensions in both 2D and 3D; see Table \ref{table:params}).
+These parameters are a the same parameters that were hand-tuned during the initial design of the basic controller; the HLP output corresponds to a residual on top of the hand-tuned values.
+We exploit bilateral symmetry by mirroring the state and action if necessary so that the swing foot is always the right foot.
 
 \begin{table}
-  \caption{Parameters of the basic walking controller for the UP phase.
-  The DOWN phase has the same set of eleven parameters (for 22 parameters total) except that \texttt{ik\_gain} now refers to the adjustment of the swing heel target used for IK.
-  \nhatch{For many of these parameters, it is reasonable to set the same values for both the UP and DOWN phases.
-We might be able to reduce problem dimensionality by removing the mostly-redundant parameters.}}
+  \caption{Parameters of the trajectory generator.
+  Note that \texttt{stance\_hip\_roll} is used only in 3D, and \texttt{swing\_knee} is used only in 2D (since in 3D we don't attempt environments with stairs).}
   \label{table:params}
@@
   \begin{tabular}{ll}
-    \texttt{torso\_world} & Torso angle in world reference frame. \\
-    \texttt{swing\_hip\_world} & Swing hip angle (sagittal plane) in world frame. \\
-    \texttt{swing\_knee\_relative} & Swing knee angle relative to thigh.\\
-    \texttt{swing\_ankle\_relative} & Swing ankle angle relative to thigh. \nhatch{Actually this is in world frame too?}\\
-    \texttt{stance\_knee\_relative} & Stance knee angle relative to thigh.\\
-    \texttt{stance\_ankle\_relative} & Stance ankle angle relative to shin.\\
-    \texttt{position\_balance\_gain} & Gain for sagittal swing hip angle based on position.\\
-    \texttt{position\_balance\_gain\_lat} & Same as above for the coronal angle.\\
-    \texttt{velocity\_balance\_gain} & Gain for sagittal swing hip angle based on velocity.\\
-    \texttt{velocity\_balance\_gain\_lat} & Same as above for the coronal angle.\\
-    \texttt{ik\_gain} & Gain for distance threshold that triggers the end of the UP phase.
+    \texttt{down\_gain} & Gain used to trigger the end of the UP state. \\
+    \texttt{toe\_off\_angle} & Angle for swing ankle during TOE-OFF state (saggital plane). \\
+    \texttt{swing\_hip} & Angle for swing hip during TOE-OFF and UP states (saggital plane). \\
+    \texttt{swing\_knee} & Angle for swing knee during TOE-OFF and UP states (saggital plane). \\
+    \texttt{stance\_hip\_roll} & Angle for stance hip during all three states (coronal plane). \\
   \end{tabular}
@@
 
-We implement the HLC by training a parametrized function approximator on a set of collected experience.
+We implement the HLP by training a parametrized function approximator on a set of collected experience.
 This dataset is a set of (state, action, actual achieved footstep location) tuples, collected as detailed in Sec. \ref{sec:curriculum} and \ref{sec:optimization}.
-Note however that we want the HLC to output an action.
+Note however that we want the HLP to output an action.
 Hence, for the regression task, the input is the pair (state, actual achieved footstep location) and the output is the action.
 Note that the input includes two ``target inputs'': the target swing heelstrike location, which was originally used when collecting the sample, and the actual swing heelstrike location, which was where the heel actually hit the ground.
-When using the HLC to prescribe actions given a new state and target, we use the same target for both of the HLC's ``target inputs''.
+When using the HLP to prescribe actions given a new state and target, we use the same target for both of the HLP's ``target inputs''.
 \nhatch{Now that I write this out, that seems pretty unclean.
 I should probably make the ``target used when collecting the sample'' part of the \emph{action} instead.
-Unfortunately, that puts us in a situation where the HLC is setting ``targets'' for the basic controller that are not actually the targets that it wants the basic controller to hit.
+Unfortunately, that puts us in a situation where the HLP is setting ``targets'' for the basic controller that are not actually the targets that it wants the basic controller to hit.
 But I guess it means we could maybe remove the \texttt{ik\_gain} parameters.}
 
-For this project, we chose a linear hypothesis class for the HLC.
+For this project, we chose a linear hypothesis class for the HLP.
 \nhatch{Soon, I want to experiment with more complicated hypothesis classes.}
@@
 Note that the starting states for each round are chosen from the set of ending states collected during all previous rounds.
-Hence, the input distribution of the dataset is indeed representative of the distribution of states likely to be seen by the HLC at test time.
+Hence, the input distribution of the dataset is indeed representative of the distribution of states likely to be seen by the HLP at test time.
 
@@
 For instance, at first, the targets are all chosen to lie in the ground plane, and the variance in their distance from one another is relatively small.
-Once the HLC reaches a desired level of accuracy, the variance is increased.
+Once the HLP reaches a desired level of accuracy, the variance is increased.
 After that, we change the environment such that the ground is no longer a single flat sheet, but rather a series of stepping stones with gaps in between them.
@@
 We believe \nhatch{and test experimentally} that this curriculum of gradually increasing difficulty is key to good performance.
-In our algorithm, the initial point of the optimization problem in step 3 is the action suggested by the current HLC.
+In our algorithm, the initial point of the optimization problem in step 3 is the action suggested by the current HLP.
 The problem is nonconvex, so this initialization point matters a lot.
 With better initialization, the stationary point is more likely to be a good local optimum, and the optimization problem might be more quickly solvable.
-Hence, it makes sense to scale the difficulty of the optimization problem roughly proportionally to the current capabilities of the HLC.
+Hence, it makes sense to scale the difficulty of the optimization problem roughly proportionally to the current capabilities of the HLP.
 
@@
 \nhatch{This is super important given the current complexity of the system. We need to remove some of these elements to make a more focused paper, so we need to identify what matters and what doesn't.
-For reference, a list of the various components that we want to pare down: the many modifications to Simbicon, the complexity of the curriculum (and whether a curriculum is necessary at all), the hypothesis class for the HLC, the data collection algorithm from Sec.
+For reference, a list of the various components that we want to pare down: the many modifications to Simbicon, the complexity of the curriculum (and whether a curriculum is necessary at all), the hypothesis class for the HLP, the data collection algorithm from Sec.
 \ref{sec:curriculum} (e.g.
-do Gaussian perturbations help), the set of parameters that are controllable by the HLC, whether RANSAC is necessary, whether the basic controller observes the CoM or not.}
+do Gaussian perturbations help), the set of parameters that are controllable by the HLP, whether RANSAC is necessary, whether the basic controller observes the CoM or not.}
 

### 2019-01-31 paper/main.tex
Update full paper draft

@@
 
-\newcommand{\nhatch}[1]{{\leavevmode\color{blue} Nathan: #1}}
+\newcommand{\nhatch}[1]{{\leavevmode\color{magenta} Nathan: #1}}
 
@@
 
-\title{Curriculum and Imitation Learning for Biped Locomotion Over Rough Terrain}
+\title{Curriculum and Imitation Learning for Biped Locomotion on Rough Terrain}
 \author{TBD}
@@
 \emph{Biped} locomotion is more difficult because the support polygon is small and the center of mass is relatively high.
-Biped locomotion across \emph{rough terrain} is more difficult still, because the variability in the contact surface complicates the dynamics.
+Biped locomotion on \emph{rough terrain} is more difficult still, because large, frequent obstacles make even small mistakes catastrophic.
 
@@
 The policies learned by deep RL are both computationally and physically fast.
-However, due to the extreme sample inefficiency of deep RL algorithms, their effectiveness on more challenging environments---for example, physical bipeds or rough terrain---is uncertain.
+However, due to the extreme sample inefficiency of deep RL algorithms, their effectiveness in more challenging environments---for example, physical bipeds or rough terrain---is uncertain.
 
@@
 The TG allows us to take advantage of prior knowledge, while the learned policy allows us to address cases where prior knowledge is insufficient.
+As shown in \citet{iscen2018pmtg}, this environmental simplification makes it feasible to learn a policy that can control walking speed for a quadruped robot.
 
-Building on this work, we approach dynamic biped locomotion on rough terrain using an approach similar to PMTG that uses curriculum learning and imitation learning.
-We call this curricular imitative modulation (CIM).
-In contrast to prior work in footstep planning, our algorithm learns a dynamic policy that requires no optimization at execution time.
+In this manuscript, we propose several additional ways to simplify robotic learning tasks by taking advantage of prior knowledge.
+We then show that these simplifications, taken together, make it feasible to use machine learning for (simulated) biped locomotion over rough terrain.
+In contrast to prior work in footstep planning, our approach learns a dynamic policy that requires no optimization at execution time.
 In contrast to prior work in learning-based locomotion, our learned policy \emph{controls its footstep locations}, which allows it to handle larger and more frequent obstacles.
@@
 \begin{itemize}
-  \item A trajectory generator for controllable biped locomotion, based on SIMBICON \citep{yin2007simbicon} but with several important modifications
-  \item An algorithm combining curriculum and imitation learning to learn efficiently a policy to modulate the trajectory generator
-  \item Experiments demonstrating the success of this approach in challenging simulated environments
+  \item a trajectory generator for controllable biped locomotion, based on SIMBICON \citep{yin2007simbicon} but with several modifications to enhance controllability;
+  \item a policy architecture similar to PMTG but with a simpler HLP interface;
+  \item an efficient learning algorithm for this architecture, which takes advantage of reward shaping and curriculum learning; and
+  \item simulated experiments evaluating the necessity and sufficiency of this approach for dynamic biped locomotion over rough terrain.
 \end{itemize}
@@
 These approaches tend to generate \emph{hierarchical} plans in order to reduce computational requirements.
-First, they choose footstep locations using search, like A*.
+First, they choose footstep locations using search, like A*.\nhatch{citation??}
 Based on the footstep plan, they generate a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum \citep{mordatch2010robust} or contact wrench cone \citep{dai2016planning}.
@@
 
-These approaches tend to be reliable but slow (four seconds per step in \citet{feng2015optimization}) because static walking is easier to model and optimization takes a lot of computation time.
-They also require a lot of domain knowledge and manual tuning in order to engineer good models and handle situations when the real world deviates from ideal physics.
+These approaches tend to be reliable but slow; e.g. four seconds per step in \citet{feng2015optimization}.
+This is because static walking is easier to model and optimization takes a lot of computation time.
+They also require significant domain knowledge and manual tuning in order to engineer good models and handle situations when the real world deviates from ideal physics.
 For now, they are the best known approaches to robotic locomotion, given the supreme importance of safety and reliability for large physical systems and high-stakes competitions such as the DARPA Robotics and Learning Locomotion Challenges.
@@
 \cite{peng2017deeploco} use a hierarchical reinforcement learning algorithm where the upper-level controller chooses footstep locations and the lower-level controller tries to execute those plans.
-However, the target step locations are unrealistic and rarely actually achieved.
+However, it produces target step locations that are unrealistic and rarely actually achieved.
 The path that the agent follows is perfectly flat; on rough terrain, such inaccurate step planning would probably fail.
@@
 The final policy is based on nearest-neighbor search among this set of collected experience.
-However, due to the curse of dimensionality of nearest-neighbor search,
-it would be challenging to apply this approach to a more complicated model.
+However, the curse of dimensionality for nearest-neighbor search
+makes it challenging to apply this approach to a more complicated model.
 
-\subsection{Deep reinforcement learning}
+\subsection{Learning-based locomotion}
 
-Most of the prior work on deep RL for biped locomotion over rough terrain does not explicitly plan individual footstep locations.
+Most of the prior work on deep RL for biped locomotion on rough terrain does not explicitly plan individual footstep locations.
 Examples of this are \cite{peng2018deepmimic, heess2017emergence, peng2016terrain}. The resulting policies travel forward quickly, avoid certain obstacles, and are robust to small missteps.
@@
 
-\subsection{Policies modulating trajectory generators}
+Perhaps the most closely related prior work is \citet{iscen2018pmtg}.
+As described in the introduction, PMTGs use a trajectory generator to simplify the environment when learning challenging tasks, such as controllable robot locomotion.
+This allows the hypothesis class of the HLP to be very simple (e.g. linear) and the learning algorithm for the HLP to be very simple as well (e.g. ARS \citep{mania2018simple}).
+However, this has been applied only to control walking speed for a quadruped on flat ground.
+The policy architecture in our paper is similar to a PMTG in that we learn a high-level policy that modulates a TG.
+However, one important difference is that in our paper, the HLP and the TG run at different frequencies.
+We will give a detailed description of these differences in the next section.
 
-Perhaps the most closely related prior work is \citet{iscen2018pmtg}.
-Like CIM, PMTGs use a trajectory generator to simplify the environment when learning controllable robot locomotion.
-Also like CIM, the hypothesis class of the HLP is very simple, often linear.
-However, there are three important differences between CIM and PMTG.
+\section{Our approach} \label{sec:approach}
 
-First: In CIM, the HLP and the TG can run at different frequencies.
-/nhatch{Should we use ``LLP'' instead of ``TG'' in the main body of the paper? Might make it easier to understand.}
-For example, in our experiments, the TG runs at 2000 Hz, while the HLP runs at approximately 2 Hz.
-This works because the HLP controls \emph{only} the parameters of the TG; it does not try to learn a correction term for the low-level joint torques.
-Assuming that the low-level environment dynamics between calls to the HLP are fairly predictable (or that the TG can moderate any stochasticity that does arise), this further simplifies the environment from the point of view of the HLP.
-
-Second: CIM can take advantage of a well-designed reward signal by learning based on \emph{individual state-action pairs} rather than complete episodes.
-For example, in locomotion, it is possible to design a reward function such that a high-reward action for an individual footstep ensures that the robot remains balanced, which allows it to complete the rest of the episode.
-This allows us to learn good parameters for each footstep one at a time, then use imitation learning to generalize.
-By eliminating the infamous credit assignment problem, this makes the learning task much easier for the HLP.
-In contrast, PMTG does not take advantage of the ability to design the reward function.
-
-Finally, CIM uses a curriculum of environments of increasing difficulty to further ease the learning problem.
-The overall thrust of these differences from PMTG is to simplify the high-level environment dynamics as much as possible.
-This makes CIM suitable for highly challenging environments, such as biped locomotion over rough terrain, where pure RL or random search algorithms are unlikely to succeed.
-In \citet{iscen2018pmtg}, the HLP learns to control the walking speed of a quadruped robot.
-In this paper, our HLP learns to control the precise heel placement of each footstep.\footnote{
-One other important difference is that \citet{iscen2018pmtg} test their algorithm on a physical robot, while all of our experiments are simulated.
-We believe that CIM would be useful to learn policies for physical robots as well, and we intend to investigate this in future work.}
-
-\section{Our approach}
-
-To solve the problem of locomotion over rough terrain, our policy must be able to place its footsteps with high accuracy.
-Our goal is furthermore to do this with a fast, dynamic gait that requires no optimization during policy execution.
-For now, we assume that perception is solved and that the step plan (in the form of a list of target step locations) is given by an oracle.\footnote{
-In 3D, each target step location also includes a target \emph{heading}: the direction that the robot should face during the step.}
-The problem is to choose motor torques to accomplish the given step plan.
-
-Our algorithm is called ``curricular imitative modulation'' (CIM).
+Our approach is a combination of several techniques, all of which are designed to simplify the policy learning problem as much as possible.
+It applies specifically to periodic control problems such as locomotion.
+For the sake of convenience, let us refer to this bundle of techniques as ``curricular imitative modulation'' (CIM).
 It consists of four components: a trajectory generator, a high-level policy, an expert that suggests good actions for the high-level policy, and a curriculum of environments of increasing difficulty.
 
-\subsection{The trajectory generator: FSM-based PD control}
+%First, we use prior knowledge to design a good trajectory generator for the problem we are trying to solve.
+%This TG should have a set of interpretable \emph{parameters} which can be set at the beginning of a period and will affect the output of the TG until the end of that period.
+%Second, we use prior knowledge to design a per-period \emph{reward function} such that good performance on one period ensures that good performance on future periods is also possible.
+%Third, we design a \emph{single-step expert}: a (possibly expensive) oracle that, given the state at the beginning of the period, will suggest a good action (i.e. set of parameters) for that period.
+%We use imitation learning (specifically DAgger) to generalize these expert annotations to a fast high-level policy.
+%Fourth, we use curriculum learning. \nhatch{dang this is too complicated!!}
 
-In order to simplify the learning problem as much as possible, we engineer and manually tune a walking controller (a.k.a.
-TG) that, with no learning, can walk forward and to some extent control the length of its steps.
-While this controller could in principle be learned using RL techniques, the manual approach has several advantages.
-First, the basic walking gait looks fairly natural, in contrast to most results from deep RL.
-Second, it has an interpretable set of parameters which provide a convenient interface for a high-level policy.
-Third, because the basic controller already produces reasonable results, the HLP can fine-tune those results rather than learning from scratch.
-This allows the HLP to use simpler, more sample-efficient function approximators, and it reduces the computational complexity of the learning algorithm.
+\subsection{Trajectory generator} \label{sec:tg}
 
-Our TG is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
+Our concept of a trajectory generator is essentially identical to \citet{iscen2018pmtg}, except that the parameters of the TG are chosen only \emph{once per period}.
+Whereas in PMTG the high-level policy continuously modulates the output of the TG, in our approach the HLP runs much less frequently.
+It sets the TG parameters at the beginning of the period, and it does not observe the environment again until the beginning of the next period.
+Hence, our TG abstracts not only the \emph{actions} of the HLP, but also \emph{how frequently} it executes them.
+One consequence of this abstraction is that it does not make sense for the HLP to learn a correctional term for the low-level actions (motor torques) produced by the TG.
+The only output of the HLP is the parameter settings for the TG.
+%Assuming that the low-level environment dynamics between calls to the HLP are fairly predictable (or that the TG can moderate any stochasticity that does arise), this further simplifies the environment from the point of view of the HLP.
+
+In subsequent subsections, it will be helpful to have in mind a concrete example of this kind of TG.
+Our TG for biped locomotion is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
 SIMBICON uses a finite state machine (FSM) with two states: an UP state that ends after 0.3 seconds, and a DOWN state that ends on swing foot contact.
@@
 After hand-tuning the balance feedback gains and the joint angles for each state, this simple control scheme can walk forward and maintain its balance.
-
-We make several novel modifications to SIMBICON in order to control its footstep placement more precisely:
-\nhatch{Might want to put some of this in the appendix instead. Also, using some equations might make the writing more precise.}
+We make several novel modifications to SIMBICON in order to control its footstep placement more precisely:\footnote{
+For more details, see Appendix \ref{app:simbicon}.}
 \begin{enumerate}
@@
     This state transitions to the UP state after 0.2s.
-    The TG itself does not use this state (that is, the target joint angles are the same as the UP state).
-    However, it allows the HLP to use the swing ankle to influence the robot's initial acceleration while still returing to a more neutral ankle angle before heel-strike.
+    It is identical to the UP state except that the target swing ankle angle can be modified by the parameter \texttt{toe\_off\_ankle}.
+    This allows the HLP to use the swing ankle to influence the robot's initial acceleration while still returing to a more neutral ankle angle before heel-strike.
 
-  \item At the beginning of each step, set the target swing hip, swing knee, stance knee, and stance ankle angles of the UP state as a linear function of the distance to the next step target.
-    The coefficients of this function were manually tuned such that the resulting step distance approximately matches the target distance.
-
-  \item End the UP state only when both the swing heel and the CoM are within a certain distance of the target step location.
-    This distance is proportional to the CoM speed, where the proportionality constant (gain) is an additional parameter.
-    (If the swing heel hits the ground early, the step immediately ends.)
+  \item Instead of a 0.3s delay, start the DOWN state when both the swing heel and the CoM are within a certain \emph{distance} of the target step location.\footnote{
+    If the swing heel hits the ground early, the step immediately ends.}
+    This distance is proportional to the CoM speed, where the proportionality constant (gain) is controlled by the parameter \texttt{down\_gain}.
     Getting the swing heel close enough to the target is important to ensure that the DOWN state accurately hits the target.
@@
 
-  \item Automatically set the swing leg pose in the DOWN state using inverse kinematics (IK).
-    Again, adjust the target swing heel location for the purposes of IK from the true target location based on the CoM speed and an additional gain parameter.
-
-  \item Make sure that the swing heel is the first part of the foot to hit the ground, by control the swing ankle angle in world space rather than relative to the shin.
-    This does not have much effect on the step distance, but it helps make the dynamics of each impact more predictable, which is important for accurate function approximation.
-    \nhatch{Again, all of this will have to be validated with ablations\dots.}
-
-  \item In 3D, control the robot's heading using the stance hip.
-    We found that setting the target stance hip angles using inverse kinematics was sufficient, without considering the effect of other torques on the stance hip.
+  \item At the beginning of each step, set the target swing hip, swing knee, stance knee, and stance ankle angles of the UP state as a linear function of the distance to the next step target.
+    The coefficients of this function were manually tuned such that the resulting step distance approximately matches the target distance.
+    The parameters \texttt{swing\_hip} and \texttt{swing\_knee} can be used to fine-tune these automatic adjustments.
 \end{enumerate}
-
-Please see the supplemental video for examples of our TG trying to hit step targets, trying to climb stairs, and walking in a circle.
-
-%This choice simplifies the dynamics that the lower level needs to model, because each step involves no (or very few) changes of contact.
-
-\subsection{High-level policy}
-
-The TG can walk and balance on flat ground, and to some extent control its step length.
-However, it cannot hit footstep targets with high accuracy, and it usually falls when climbing stairs.
-\nhatch{See Table somewhere-in-experiments for information about how big the steps/inclines have to be before problems happen, how bad the accuracy is, etc.}
-Hence, we train a high-level policy (HLP) on top of the TG.
-It is queried once at the beginning of each step.
-It has the same input (``state'') as the basic controller: joint angles and velocities, stance heel location, and target step location (26 dimensions in 2D, 54 dimensions in 3D).
-In addition, it sees the two previous target footstep locations.
-(This is important in ``stepping stones'' environments where the exact location of the previous stepping stone might affect the dynamics.)
-Its output (``action'') is a vector of parameters that modify the behavior of the basic controller during the subsequent step (4 dimensions in both 2D and 3D; see Table \ref{table:params}).
-These parameters are a the same parameters that were hand-tuned during the initial design of the basic controller; the HLP output corresponds to a residual on top of the hand-tuned values.
-We exploit bilateral symmetry by mirroring the state and action if necessary so that the swing foot is always the right foot.
 
@@
   \begin{tabular}{ll}
-    \texttt{down\_gain} & Gain used to trigger the end of the UP state. \\
-    \texttt{toe\_off\_angle} & Angle for swing ankle during TOE-OFF state (saggital plane). \\
-    \texttt{swing\_hip} & Angle for swing hip during TOE-OFF and UP states (saggital plane). \\
-    \texttt{swing\_knee} & Angle for swing knee during TOE-OFF and UP states (saggital plane). \\
+    \texttt{toe\_off\_ankle} & Angle for swing ankle during TOE-OFF state (saggital plane). \\
+    \texttt{down\_gain} & Gain on distance from target used to trigger the end of the UP state. \\
+    \texttt{swing\_hip} & Angle for swing hip during UP state (saggital plane). \\
+    \texttt{swing\_knee} & Angle for swing knee during UP state (saggital plane). \\
     \texttt{stance\_hip\_roll} & Angle for stance hip during all three states (coronal plane). \\
@@
 
-We implement the HLP by training a parametrized function approximator on a set of collected experience.
-This dataset is a set of (state, action, actual achieved footstep location) tuples, collected as detailed in Sec. \ref{sec:curriculum} and \ref{sec:optimization}.
-Note however that we want the HLP to output an action.
-Hence, for the regression task, the input is the pair (state, actual achieved footstep location) and the output is the action.
-Note that the input includes two ``target inputs'': the target swing heelstrike location, which was originally used when collecting the sample, and the actual swing heelstrike location, which was where the heel actually hit the ground.
-When using the HLP to prescribe actions given a new state and target, we use the same target for both of the HLP's ``target inputs''.
-\nhatch{Now that I write this out, that seems pretty unclean.
-I should probably make the ``target used when collecting the sample'' part of the \emph{action} instead.
-Unfortunately, that puts us in a situation where the HLP is setting ``targets'' for the basic controller that are not actually the targets that it wants the basic controller to hit.
-But I guess it means we could maybe remove the \texttt{ik\_gain} parameters.}
+See Table \ref{table:params} for a summary of the parameters of the TG.
+These parameters are \emph{residual} in the sense that setting all of them to zero will produce the default behavior of the TG.
+Examples of this default behavior can be seen in the supplemental video.
+On its own, the TG can walk forward, walk in a circle, and to some extent control the length of its steps.
+However, it cannot hit footstep targets with high accuracy, and it usually falls when climbing stairs.
+Now that we have incorporated as much prior knowledge as possible into the design of the TG, we turn our attention to the learning aspects of our approach.
 
-For this project, we chose a linear hypothesis class for the HLP.
-\nhatch{Soon, I want to experiment with more complicated hypothesis classes.}
-Because our dataset is relatively small, we use ridge regression to avoid overfitting.
-Our dataset also tends to contain outliers, because the basic controller can behave unpredictably in some situations (such as when the agent stubs its toe).
-We handle this by using RANSAC \citep{fischler1981random} to discard outliers at training time.
-\nhatch{RANSAC might not be necessary if the optimization alg. to collect the dataset is good enough. Maybe that alg could throw out examples where no good parameter setting was found.}
+\subsection{High-level policy}
+
+Curricular imitative modulation (CIM) uses a high-level policy, queried once per period, to modulate the behavior of the TG.
+Its input is the robot/environment state plus any control input to indicate what we would like to happen during the subsequent period, and
+its output is a set of TG parameters.
+With a strong TG, the policy class for the HLP can be very simple.
+For example, in the locomotion experiments below, the control input is the location of the next footstep target,
+and we use a linear hypothesis class for the HLP.
+
+Assuming we have invented a reward function that usefully describes our learning objective, the next question is how to train this policy.
+One option would be to use the same algorithms as \citet{iscen2018pmtg}: reinforcement learning algorithms or evolutionary strategies.
+Given the simplicity of the policy class, such algorithms might be reasonable despite their sample complexity.
+Indeed, we compare against this approach in the experiments.
+However, CIM finds another way to simplify the learning problem: create an expert.
+
+\subsection{Expert} \label{sec:expert}
+
+Imitation learning (IL) is a common strategy to improve sample complexity in reinforcement learning problems \citep{osa2018algorithmic}.
+Given an \emph{expert}---a policy that can choose good actions for any given state, but that is too expensive or too slow to use at test time---we can use
+a standard IL algorithm such as DAgger \citep{ross2011reduction} to generalize to a fast policy.
+
+Due to the unusual policy architecture of CIM, it is unlikely that preexisting experts exist,
+so CIM designs its own.
+For general RL problems, this would be difficult because of the infamous \emph{credit assignment problem}:
+sometimes, an action during one period of the episode can leave the robot in a bad state to achieve the desired result for the next period.
+%Such an action is in fact a bad action, even if the robot does achieve the desired result during the earlier period.
+For example, in our locomotion experiments below, it is not enough just to hit the footstep targets; the robot must also maintain its balance.
+
+Fortunately, in locomotion and in many other cases of interest, there is an easy way to mitigate this problem: penalize actions that leave the robot in a bad state.
+If the penalty is well designed, we can ensure that a high-reward action for any given period will leave the robot in a good position to find high-reward actions for all future periods.
+Such \emph{reward shaping} effectively eliminates the credit assignment problem and drastically reduces the search horizon needed to find an expert action.
+In turn, this makes imitation learning a feasible way to simplify the HLP learning problem.\footnote{
+In principle, one could use a shaped reward function without also using imitation learning; we compare against this approach in the experiments.}
+
+For example, in the case of the locomotion experiments below, our expert is designed as follows.
+Given a starting state and a target footstep location, we set up an optimization problem
+over the parameters of the TG.
+We define the achieved step location as the location of the swing heel at the moment that the step ends, which is the moment that any part of the swing foot makes contact with the ground.\footnote{The step also ends if the robot crashes; i.e.
+falls off of the edge of a stepping stone, falls over, etc.}
+\nhatch{TODO: describe this reward more carefully. In particular, it is one minus the penalties described below.}
+The objective is the Euclidean distance between the target and achieved step locations,
+plus a hinge loss penalty if the CoM ends up too far \emph{past} the target (projected along the direction of travel).
+At each iteration, we estimate gradients for this objective as a weighted average of $Q$ random directions according to ARS \citep{mania2018simple}.
+We end the optimization once the accuracy reaches two centimeters, or once a certain maximum number of iterations (in our case, five) is reached.
+The hinge loss penalty protects the robot's balance, ensuring that if the expert finds a successful action, it will actually be a \emph{good} action to use to train the HLP.
 
@@
 
-We collect the dataset in an iterative, aggregate way, similar to DAgger \citep{ross2011reduction}.
-First, we collect an initial dataset by running the basic controller for several steps on flat ground with a variety of different target step lengths.
-On subsequent rounds of data collection, we proceed as follows:
-\nhatch{Format this as an algorithm.}
-\begin{enumerate}
-  \item Choose $N$ starting states from the existing dataset.
-  \item For each starting state, choose $M$ target locations for the next step.
-  \item For each target location, find an action to achieve that target using random optimization (see Sec. \ref{sec:optimization}).
-  \item For each of $P$ Gaussian perturbations of the optimal action from step 3, run the simulation and add the sample to the dataset.
-\end{enumerate}
-For our experiments, we used $N=128$, $M=1$, $P=4$.
-The optimization in step 3 is rather expensive, hence the relatively small number of samples $NMP$ collected at each round.
-\nhatch{I'm not sure that the Gaussian perturbations are actually helpful. If they are, it might make sense to perturb the starting state as well, not just the action.}
+Finally, CIM incorporates an element of curriculum learning into the training process.
+Rather than immediately training the HLP on the most difficult version of the problem,
+it uses prior knowledge to design a series of environments of gradually increasing difficulty.
+One rationale for this approach lies in the design of the CIM expert.
+The initial point of the expert optimization problem is the action suggested by the current HLP.
+The problem is usually nonconvex, so this initialization point matters a lot.
+With better initialization, the solution found by the expert is more likely to be a good local optimum, and the optimization problem might be more quickly solvable.
+Hence, it makes sense to scale the difficulty of the expert optimization problem roughly proportionally to the current capabilities of the HLP.
+This is similar to other slow-expert/fast-learner approaches such as ExIt \citep{anthony2017thinking} and AlphaGo Zero \citep{silver2017mastering}.
 
-Note that the starting states for each round are chosen from the set of ending states collected during all previous rounds.
-Hence, the input distribution of the dataset is indeed representative of the distribution of states likely to be seen by the HLP at test time.
+For instance, in our locomotion experiments, the footstep targets are at first all chosen to lie in the ground plane, and the variance in their distance from one another is relatively small.
+After iteration 3, the variance is increased.
+After iteration 9, we change the environment such that the ground is no longer a single flat sheet, but rather a series of irregularly spaced stairs.
 
-Finally, we incorporate an element of curriculum learning into the data collection process.
-At certain manually specified points in the process, we change the distribution of targets used in step 2 and/or change some details of the simulation.
-For instance, at first, the targets are all chosen to lie in the ground plane, and the variance in their distance from one another is relatively small.
-Once the HLP reaches a desired level of accuracy, the variance is increased.
-After that, we change the environment such that the ground is no longer a single flat sheet, but rather a series of stepping stones with gaps in between them.
-Later still, we introduce some vertical variation into the target step locations.\footnote{The stepping stones are shifted up or down to match the target location.}.
-\nhatch{And so forth. I haven't systematically studied this yet to see how difficult we can get.}
-
-We believe \nhatch{and test experimentally} that this curriculum of gradually increasing difficulty is key to good performance.
-In our algorithm, the initial point of the optimization problem in step 3 is the action suggested by the current HLP.
-The problem is nonconvex, so this initialization point matters a lot.
-With better initialization, the stationary point is more likely to be a good local optimum, and the optimization problem might be more quickly solvable.
-Hence, it makes sense to scale the difficulty of the optimization problem roughly proportionally to the current capabilities of the HLP.
-
-\subsection{Single-step optimization} \label{sec:optimization}
-
-We now give more details about the optimization problem in step 3 of Sec. \ref{sec:curriculum}.
-The objective of the problem is the Euclidean distance between the target and achieved step locations.
-We define the achieved step location as the location of the swing heel at the moment that the step ends, which is the moment that any part of the swing foot makes contact with the ground.\footnote{The step also ends if the robot crashes; i.e.
-falls off of the edge of a stepping stone, falls over, etc.}
-The optimization variables are the parameters of the basic controller (see Table \ref{table:params}).
-There are no constraints.
-\nhatch{Actually, we constrain most of the parameters to be zero.
-In my experiments so far, only seven of the 22 parameters are allowed to vary.
-It might be possible to gradually allow more parameters to be controlled according to the requirements of the curriculum.}
-
-Because it would be difficult to calculate gradients for this objective, we use the zero-order optimization algorithm described in \cite{mania2018simple}.
-At each iteration, we estimate gradients as a weighted average of $Q = 4$ random directions, with the weight for each direction chosen as the central difference of the objective value along that direction.
-Hence, we perform $2Q$ single-footstep rollouts for each update of the optimization variables.
-We end the optimization once the accuracy reaches two centimeters, or once a certain maximum number of iterations (in our case, ten) is reached.
+%The overall thrust of these differences from PMTG is to simplify the high-level environment dynamics as much as possible.
+%This makes CIM suitable for highly challenging environments, such as biped locomotion over rough terrain, where pure RL or random search algorithms are unlikely to succeed.
 
@@
 
-We perform experiments with a series of three robot models of increasing realism: (1) a simple planar biped, (2) a simple 3D biped, and (3) a model of the Darwin robot.
-All experiments are performed using the DART simulator at a simulation rate of 2000 Hz.
+\begin{figure}
+  \includegraphics[width=0.48\textwidth]{2D_ep.png}
+  \includegraphics[width=0.48\textwidth]{3D_ep.png}
+  \caption{Screenshots of example episodes.\label{fig:example episodes}}
+\end{figure}
 
-\nhatch{This section is still pretty rough; sorry.}
-Our experiments are designed to test two hypotheses.
-Our primary hypothesis is that this combination of imitation and curriculum learning can learn to navigate a variety of complicated locomotion environments, learning faster and achieving better results than other reinforcement learning methods.
+We evaluate CIM on the problem of dynamic biped locomotion over rough terrain.
+Using the DART simulator \citep{lee2018dart}, we design two RL problems as illustrated in Fig. \ref{fig:example episodes}.
+In both environments, the goal is to walk across 16 stepping stones while maximizing the reward per step as defined in section \ref{sec:expert}.
+Crashing terminates the episode early;
+thus, the maximum reward per episode is 16.
+The distribution of offsets between stepping stones is uniformly random.
+In 2D, the offset in the X direction (the forward direction) is between 30 and 60 cm, while the Y offset (vertical) is between 5 and 10 cm.
+In 3D, the X offset is 25--45cm, the Y offset is 0, and the Z offset is 30--50cm (alternating sign at each footstep).
 
-We collect both quantitative and qualitative metrics to measure ``achieving better results''.
-Quantitatively, we evaluate the robot by presenting it with a series of small stepping stones placed at a random $(x,y,z)$ offsets from each other.
-\nhatch{It would be nice to be able to place them at different rotations/orientations as well.}
-We score the robot based on a combination of the number of successful steps before crashing and the accuracy of foot placement at each step.
-We collect this score for a number of different distributions of the $(x,y,z)$ offsets (of increasing difficulty), and find \nhatch{hopefully} that our model achieves better performance than deep RL methods on most of these environments.
-\nhatch{We will have to think carefully about how to make this a fair comparison, since most deep RL methods don't actually care about foot placement.
-For one thing, the action space for the deep RL methods will also have to be ``parameter settings for the basic walking controller''.}
-We also consider the robustness of the model to inaccurate perception by perturbing the target step locations with respect to the actual location of the stepping stones.
-\nhatch{We should probably also evaluate robustness to external force perturbations etc., but I'm nervous that that won't end well for us.
-We don't really have good feedback mechanisms \emph{during} each footstep.}
-Qualitatively, we document motion quality by means of supplementary videos, which we encourage the reader to consult.
+Using the TG and learning algorithm described in Sec.
+\ref{sec:approach} above, our HLP is able to achieve reward $15.7\pm0.2$ (one standard deviation) after only 1 kajillion simulated training footsteps.
+\nhatch{Put actual numbers here.}
+Examples of the final policies may be seen in the supplemental video.
+To the best of our knowledge, our approach is the first to achieve such precise foot placement
+using a fast, dynamic gait that requires no optimization during policy execution.
 
-Our two metrics for ``learning faster'' are (1) number of simulation frames of experience collected, and (2) total computation time.
-We consider both necessary, because the relative expense of more experience versus more computation may vary across application domains.
+%Our secondary hypothesis is that the learned policy can be run in real-time while still achieving results comparable to more expensive, model-based optimization methods.
+%This hypothesis is difficult to test, because model-based optimization methods are difficult to implement for direct comparison.
+%Thus, our metric is simply the percent real time necessary to compute the control for the next time step, which we compare to numbers reported in optimization-based papers.
 
-Our secondary hypothesis is that the learned policy can be run in real-time while still achieving results comparable to more expensive, model-based optimization methods.
-This hypothesis is difficult to test, because model-based optimization methods are difficult to implement for direct comparison.
-Thus, our metric is simply the percent real time necessary to compute the control for the next time step, which we compare to numbers reported in optimization-based papers.
+\subsection{Imitation learning ablations}
 
-Finally, we conduct ablative analysis of the importance of various components of our approach.
-\nhatch{This is super important given the current complexity of the system. We need to remove some of these elements to make a more focused paper, so we need to identify what matters and what doesn't.
-For reference, a list of the various components that we want to pare down: the many modifications to Simbicon, the complexity of the curriculum (and whether a curriculum is necessary at all), the hypothesis class for the HLP, the data collection algorithm from Sec.
-\ref{sec:curriculum} (e.g.
-do Gaussian perturbations help), the set of parameters that are controllable by the HLP, whether RANSAC is necessary, whether the basic controller observes the CoM or not.}
+In this section, we evaluate CIM against a baseline that does not use imitation learning.
+Instead, the baseline uses ARS \citep{mania2018simple} over entire episodes.
+That is, at each iteration we update the HLP directly by estimating the gradient as a weighted combination of $Q$ random directions,
+with weights determined by the total reward as defined in section \ref{sec:expert}.
+Note that this baseline is very similar to PMTG \citep{iscen2018pmtg}.
+The only difference is that the policy architecture still uses the temporal abstraction defined in section \ref{sec:tg}.\footnote{
+As we have not implemented a TG for biped locomotion that can be continuously modulated, this seems like the fairest comparison we could make.}
+
+For each approach (CIM and baseline) we run the experiment from scratch with four different random seeds.
+Results are shown in Fig.
+TODO.
+The X axis shows the total number of simulated footsteps taken during training, and the Y axis shows total reward.
+We plot the mean and standard deviation over 32 testing episodes (eight episodes for each of the four experiments).
+
+\subsection{Curriculum learning ablations}
+
+In this section, we evaluate CIM against a baseline that does not use curriculum learning.
+Instead, the baseline begins training in the final, most difficult testing environment.
+Results are shown in Fig. TODO.
 
@@
 
-We see at least two important directions for future work.
-\begin{enumerate}
-  \item \emph{Sim-to-real:}
-    We have successfully applied CIRL to a realistic simulated model, but it is likely that more work is required to transfer the resulting policy to the real world.
-  \item \emph{Motion capture:}
-    A significant portion of the effort for this project went into developing a reasonable baseline walking policy.
-    It might be possible to avoid this by bootstrapping the imitation learning process with motion capture data.
-    This would also make it easier to to extend CIRL to other robotic motion problems, such as manipulation.
-\end{enumerate}
+Curricular imitative modulation is a general approach to integrating prior knowledge with machine learning in challenging, periodic robotic control problems.
+Like PMTG \citep{iscen2018pmtg}, it simplifies the learning problem with the use of a well-designed trajectory generator.
+Moreover, it further simplifies the learning problem by (1) using a temporally abstract high-level policy and (2) training with imitation and curriculum learning.
 
-\nhatch{Miscellaneous TODO notes for me:
-\begin{enumerate}
-  \item Set damping coefficients for the DART simulation. Visak says this will help prevent numerical explosions during collisions.
-  \item Get Simbicon3D to be able to roughly control step distances, similar to 2D Simbicon. Or maybe this isn't necessary? Maybe we could jump straight to the learning algorithm?
-  \item Conduct ablation analysis to try to reduce the number of ``new ideas'' presented in this paper.
-  \item Think about how best to compare this algorithm to existing work (both RL work and traditional optimization-based step planning).
-\end{enumerate}
-}
+In particular, CIM is an intriguing direction for biped locomotion research.
+To the best of our knowledge, our experiments are the first to achieve such precise foot placement
+using a fast, dynamic gait that requires no optimization during policy execution.
+However, one significant limitation of our work so far is that all experiments were performed in simulation.
+In future work, we plan to investigate how to transfer policies trained with CIM to real-world bipedal robots.
+
+Finally, we note that a significant portion of the effort for this project went into engineering the trajectory generator.
+It might be possible to avoid this by \emph{training a trajectory generator using motion capture}.
+This would reduce the need for expert engineering and make it easier to test CIM on other robotic motion problems, such as manipulation.
 
@@
 
+\appendix
+
+\section{Detailed description of trajectory generator for biped locomotion}
+
+\label{app:simbicon}
+
+\nhatch{TODO: list the exact inputs and outputs of the TG, in particular knowing the exact CoM is unnecessary}
+
+In addition to the SIMBICON modifications discussed in Section \ref{sec:tg}, we also made the following changes:
+
+\begin{enumerate}
+  \item Set the target swing leg joint angles in the DOWN state using inverse kinematics (IK).
+    Again, adjust the target swing heel location for the purposes of IK from the true target location based on the CoM speed and the same gain parameter as above.
+
+  \item Control the swing ankle angle in world space to ensure that the swing heel is the first part of the foot to hit the ground.
+    This does not have much effect on the step distance, but it helps make the dynamics of each impact more predictable, which is important for accurate function approximation.
+    \nhatch{Again, all of this will have to be validated with ablations\dots.}
+
+  \item In 3D, control the robot's heading using the stance hip.
+    We found that setting the target stance hip angles using inverse kinematics and then using plain PD control was sufficient, without considering the effect of other torques on the stance hip.
+\end{enumerate}
+
+Note that there are also many additional parameters of SIMBICON that we could have exposed to the high-level policy; e.g. the stance ankle angle of the UP state.
+Of course, we prefer to keep the problem dimension as low as possible, and we found that the short list of parameters given in Table \ref{table:params} was enough for our experiments.
+We chose those parameters using a combination of intuition and trial-and-error.
+
+\section{More details about the high-level policy}
+
+It has the same input (``state'') as the basic controller: joint angles and velocities, stance heel location, and target step location (26 dimensions in 2D, 54 dimensions in 3D).
+In addition, it sees the two previous target footstep locations.
+(This is important in ``stepping stones'' environments where the exact location of the previous stepping stone might affect the dynamics.)
+We exploit bilateral symmetry by mirroring the state and action if necessary so that the swing foot is always the right foot.
+
+Because our dataset is relatively small, we use ridge regression to avoid overfitting.
+
+\section{Detailed experimental setup}
+
+All experiments are performed using the DART simulator at a simulation rate of 2000 Hz.
+For now, we assume that perception is solved and that the step plan (in the form of a list of target step locations) is given by an oracle.\footnote{
+In 3D, each target step location also includes a target \emph{heading}: the direction that the robot should face during the step.}
+The problem is to choose motor torques to accomplish the given step plan.
+
+Source code will be released upon publication.
+
 \end{document}

### 2019-01-31 paper/main.tex
Some small tweaks to paper

@@
 
-\section{Introduction}
+\section{Introduction} \label{sec:intro}
 
@@
 \emph{Biped} locomotion is more difficult because the support polygon is small and the center of mass is relatively high.
-Biped locomotion on \emph{rough terrain} is more difficult still, because large, frequent obstacles make even small mistakes catastrophic.
+Biped locomotion on \emph{rough terrain} is more difficult still, because frequent large obstacles make even small mistakes catastrophic.
 
@@
 In this manuscript, we propose several additional ways to simplify robotic learning tasks by taking advantage of prior knowledge.
-We then show that these simplifications, taken together, make it feasible to use machine learning for (simulated) biped locomotion over rough terrain.
+Our experiments then demonstrate that these simplifications, taken together, make it feasible to use machine learning for simulated biped locomotion over rough terrain.
 In contrast to prior work in footstep planning, our approach learns a dynamic policy that requires no optimization at execution time.
@@
   \item a policy architecture similar to PMTG but with a simpler HLP interface;
-  \item an efficient learning algorithm for this architecture, which takes advantage of reward shaping and curriculum learning; and
+  \item an efficient learning algorithm for this architecture, which takes advantage of imitation and curriculum learning; and
   \item simulated experiments evaluating the necessity and sufficiency of this approach for dynamic biped locomotion over rough terrain.
@@
 
-\nhatch{I think Honda does a lot with dynamic gaits, but I don't really understand how it works.
-I need to find and read some papers.
-To what extent can they plan step locations and handle rough terrain?}
+\nhatch{Honda and Boston Dynamics do a lot with dynamic gaits, but I don't really understand how it works. To what extent can they plan step locations and handle rough terrain?}
 
@@
 However, the curse of dimensionality for nearest-neighbor search
-makes it challenging to apply this approach to a more complicated model.
+makes it infeasible to apply this approach to a more complicated model.
 
@@
 Examples of this are \cite{peng2018deepmimic, heess2017emergence, peng2016terrain}. The resulting policies travel forward quickly, avoid certain obstacles, and are robust to small missteps.
-However, they have not been proven to work in situations with dense or large obstacles where small missteps can be catastrophic.
+However, they have not been shown to work in situations with dense or large obstacles where small missteps can be catastrophic.
 Such environments require accurate predictive planning at the level of individual contacts.
@@
 Perhaps the most closely related prior work is \citet{iscen2018pmtg}.
-As described in the introduction, PMTGs use a trajectory generator to simplify the environment when learning challenging tasks, such as controllable robot locomotion.
-This allows the hypothesis class of the HLP to be very simple (e.g. linear) and the learning algorithm for the HLP to be very simple as well (e.g. ARS \citep{mania2018simple}).
-However, this has been applied only to control walking speed for a quadruped on flat ground.
-The policy architecture in our paper is similar to a PMTG in that we learn a high-level policy that modulates a TG.
+As described in section \ref{sec:intro}, PMTGs use a trajectory generator to simplify the environment when learning challenging tasks, such as controllable robot locomotion.
+This allows the HLP to use a very simple hypothesis class (e.g. linear) and a very simple learning algorithm (e.g. ARS \citep{mania2018simple}).
+However, PMTG has been applied only to control walking speed for a quadruped on flat ground.
+The policy architecture in our paper is similar to PMTG in that we learn a high-level policy that modulates a TG.
 However, one important difference is that in our paper, the HLP and the TG run at different frequencies.
@@
 
-Our approach is a combination of several techniques, all of which are designed to simplify the policy learning problem as much as possible.
-It applies specifically to periodic control problems such as locomotion.
-For the sake of convenience, let us refer to this bundle of techniques as ``curricular imitative modulation'' (CIM).
-It consists of four components: a trajectory generator, a high-level policy, an expert that suggests good actions for the high-level policy, and a curriculum of environments of increasing difficulty.
+Our approach is called ``curricular imitative modulation'' (CIM).
+It applies to any periodic control problem (such as locomotion) and is
+designed to take advantage of prior knowledge to simplify the policy learning problem as much as possible.
+It consists of four components: a trajectory generator, a high-level policy, an expert for the high-level policy, and a curriculum of environments of increasing difficulty.
 
@@
 It sets the TG parameters at the beginning of the period, and it does not observe the environment again until the beginning of the next period.
-Hence, our TG abstracts not only the \emph{actions} of the HLP, but also \emph{how frequently} it executes them.
+Hence, our TG abstracts not only the \emph{actions} of the HLP, but also \emph{how frequently} they are executed.
 One consequence of this abstraction is that it does not make sense for the HLP to learn a correctional term for the low-level actions (motor torques) produced by the TG.
@@
 In subsequent subsections, it will be helpful to have in mind a concrete example of this kind of TG.
-Our TG for biped locomotion is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
+For biped locomotion, our TG is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
 SIMBICON uses a finite state machine (FSM) with two states: an UP state that ends after 0.3 seconds, and a DOWN state that ends on swing foot contact.
@@
     It is identical to the UP state except that the target swing ankle angle can be modified by the parameter \texttt{toe\_off\_ankle}.
-    This allows the HLP to use the swing ankle to influence the robot's initial acceleration while still returing to a more neutral ankle angle before heel-strike.
+    This allows the HLP to use the swing ankle to influence the robot's initial acceleration while still returing to a more neutral angle before heel-strike.
 
-  \item Instead of a 0.3s delay, start the DOWN state when both the swing heel and the CoM are within a certain \emph{distance} of the target step location.\footnote{
-    If the swing heel hits the ground early, the step immediately ends.}
+  \item Instead of a 0.3s delay, start the DOWN state when both the swing heel and the CoM are within a certain \emph{distance} of the target heel-strike location.\footnote{
+    If the swing foot hits the ground early, the step immediately ends.}
     This distance is proportional to the CoM speed, where the proportionality constant (gain) is controlled by the parameter \texttt{down\_gain}.
@@
 However, it cannot hit footstep targets with high accuracy, and it usually falls when climbing stairs.
-Now that we have incorporated as much prior knowledge as possible into the design of the TG, we turn our attention to the learning aspects of our approach.
+Now that we have incorporated significant prior knowledge into the design of the TG, we turn our attention to the learning aspects of our approach.
 
@@
 
-Curricular imitative modulation (CIM) uses a high-level policy, queried once per period, to modulate the behavior of the TG.
-Its input is the robot/environment state plus any control input to indicate what we would like to happen during the subsequent period, and
-its output is a set of TG parameters.
-With a strong TG, the policy class for the HLP can be very simple.
+CIM uses a high-level policy, queried once per period, to modulate the behavior of the TG.
+Its input is the robot and environment state plus any control input to indicate what should happen during the subsequent period.
+Its output is a set of TG parameters.
+With a well-designed TG, the policy class for the HLP can be very simple.
 For example, in the locomotion experiments below, the control input is the location of the next footstep target,
@@
 
-Assuming we have invented a reward function that usefully describes our learning objective, the next question is how to train this policy.
+Assuming we have specified a reward function that usefully describes our learning objective, the next question is how to train this policy.
 One option would be to use the same algorithms as \citet{iscen2018pmtg}: reinforcement learning algorithms or evolutionary strategies.
-Given the simplicity of the policy class, such algorithms might be reasonable despite their sample complexity.
+Given the simplicity of the policy class, such algorithms might be reasonable despite their sample inefficiency.
 Indeed, we compare against this approach in the experiments.
-However, CIM finds another way to simplify the learning problem: create an expert.
+However, for CIM we find another way to simplify the learning problem: create an expert.
 
@@
 For general RL problems, this would be difficult because of the infamous \emph{credit assignment problem}:
-sometimes, an action during one period of the episode can leave the robot in a bad state to achieve the desired result for the next period.
+sometimes, an action during one period of the episode can leave the robot in a bad position to achieve the desired result for the next period.
 %Such an action is in fact a bad action, even if the robot does achieve the desired result during the earlier period.
@@
 Such \emph{reward shaping} effectively eliminates the credit assignment problem and drastically reduces the search horizon needed to find an expert action.
-In turn, this makes imitation learning a feasible way to simplify the HLP learning problem.\footnote{
+This makes imitation learning a feasible way to simplify the HLP learning problem.\footnote{
 In principle, one could use a shaped reward function without also using imitation learning; we compare against this approach in the experiments.}
 
-For example, in the case of the locomotion experiments below, our expert is designed as follows.
-Given a starting state and a target footstep location, we set up an optimization problem
-over the parameters of the TG.
-We define the achieved step location as the location of the swing heel at the moment that the step ends, which is the moment that any part of the swing foot makes contact with the ground.\footnote{The step also ends if the robot crashes; i.e.
+For example, in the case of the locomotion experiments below, our expert is defined by the following optimization problem.
+Given a starting state and a target footstep location, we define the \emph{achieved step location} for an action as the location of the swing heel at the moment that the step ends, which is the moment that any part of the swing foot makes contact with the ground.\footnote{The step also ends if the robot crashes; i.e.
 falls off of the edge of a stepping stone, falls over, etc.}
-\nhatch{TODO: describe this reward more carefully. In particular, it is one minus the penalties described below.}
 The objective is the Euclidean distance between the target and achieved step locations,
 plus a hinge loss penalty if the CoM ends up too far \emph{past} the target (projected along the direction of travel).
+\nhatch{TODO: describe this reward more carefully. In particular, it is one minus the penalties described above.}
 At each iteration, we estimate gradients for this objective as a weighted average of $Q$ random directions according to ARS \citep{mania2018simple}.
-We end the optimization once the accuracy reaches two centimeters, or once a certain maximum number of iterations (in our case, five) is reached.
+We stop iterating once the accuracy reaches two centimeters, or once a certain maximum number of iterations (in our case, five) is reached.
 The hinge loss penalty protects the robot's balance, ensuring that if the expert finds a successful action, it will actually be a \emph{good} action to use to train the HLP.
@@
 Rather than immediately training the HLP on the most difficult version of the problem,
-it uses prior knowledge to design a series of environments of gradually increasing difficulty.
+we use prior knowledge to design a series of environments of gradually increasing difficulty.
 One rationale for this approach lies in the design of the CIM expert.
 The initial point of the expert optimization problem is the action suggested by the current HLP.
-The problem is usually nonconvex, so this initialization point matters a lot.
+Since the problem is usually nonconvex, this initialization point matters a lot.
 With better initialization, the solution found by the expert is more likely to be a good local optimum, and the optimization problem might be more quickly solvable.
@@
 After iteration 9, we change the environment such that the ground is no longer a single flat sheet, but rather a series of irregularly spaced stairs.
+\nhatch{TODO: fix these iteration numbers, and maybe put a more detailed description of the curriculum in an appendix}
 
@@
 We evaluate CIM on the problem of dynamic biped locomotion over rough terrain.
-Using the DART simulator \citep{lee2018dart}, we design two RL problems as illustrated in Fig. \ref{fig:example episodes}.
+Using the DART simulator \citep{lee2018dart}, we implement two RL problems as illustrated in Fig. \ref{fig:example episodes}.
 In both environments, the goal is to walk across 16 stepping stones while maximizing the reward per step as defined in section \ref{sec:expert}.
@@
 
-Using the TG and learning algorithm described in Sec.
-\ref{sec:approach} above, our HLP is able to achieve reward $15.7\pm0.2$ (one standard deviation) after only 1 kajillion simulated training footsteps.
+Using the TG and learning algorithm described in Sec. \ref{sec:approach} above, our HLP is able to achieve reward $A\pm B$ (one standard deviation) after only $C$ simulated training footsteps.
 \nhatch{Put actual numbers here.}
@@
 That is, at each iteration we update the HLP directly by estimating the gradient as a weighted combination of $Q$ random directions,
-with weights determined by the total reward as defined in section \ref{sec:expert}.
+with weights determined by the change in total reward as defined in section \ref{sec:expert}.
 Note that this baseline is very similar to PMTG \citep{iscen2018pmtg}.
-The only difference is that the policy architecture still uses the temporal abstraction defined in section \ref{sec:tg}.\footnote{
+The only difference is that its policy architecture still uses the temporal abstraction explained in section \ref{sec:tg}.\footnote{
 As we have not implemented a TG for biped locomotion that can be continuously modulated, this seems like the fairest comparison we could make.}
@@
 Finally, we note that a significant portion of the effort for this project went into engineering the trajectory generator.
-It might be possible to avoid this by \emph{training a trajectory generator using motion capture}.
+It might be possible to avoid this by training a trajectory generator using motion capture.
 This would reduce the need for expert engineering and make it easier to test CIM on other robotic motion problems, such as manipulation.

### 2019-01-31 paper/rss/main.tex
Move paper to RSS format

--- 2019-01-31 paper/main.tex
@@
-\documentclass[a4paper]{article}
+\documentclass[conference]{IEEEtran}
+\usepackage{times}
 
-%% Language and font encodings
-\usepackage[english]{babel}
-\usepackage[utf8x]{inputenc}
-\usepackage[T1]{fontenc}
+% numbers option provides compact numerical references in the text. 
+\usepackage[numbers]{natbib}
+\usepackage{multicol}
+\usepackage[bookmarks=true]{hyperref}
+\usepackage{xcolor}
+\usepackage{graphicx}
 
-%% Sets page size and margins
-\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
-
-%% Useful packages
-\usepackage{amsmath}
-\usepackage{verbatim}
-\usepackage{amsfonts}
-\usepackage{enumitem}
-\usepackage{graphicx}
-\usepackage{listings}
-\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
-\usepackage{algorithm}
-\usepackage[noend]{algpseudocode}
-\usepackage{mathtools}
-\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
-\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
-
-\usepackage[round]{natbib}
-
-\makeatletter
-\def\BState{\State\hskip-\ALG@thistlm}
-\makeatother
-
-\DeclareMathOperator{\Tr}{Tr}
-\DeclareMathOperator{\sep}{sep}
-\lstset{frame=tb,
-  language=Python,
-  aboveskip=3mm,
-  belowskip=3mm,
-  showstringspaces=false,
-  columns=flexible,
-  basicstyle={\small\ttfamily},
-  numbers=none,
-  numberstyle=\tiny\color{gray},
-  keywordstyle=\color{blue},
-  commentstyle=\color{red},
-  stringstyle=\color{red},
-  breaklines=true,
-  breakatwhitespace=true,
-  tabsize=3
+\pdfinfo{
+   /Author (Homer Simpson)
+   /Title  (Robots: Our new overlords)
+   /CreationDate (D:20101201120000)
+   /Subject (Robots)
+   /Keywords (Robots;Overlords)
 }
-
-\newcommand{\norm}[1]{\lVert #1 \rVert}
-\newcommand{\fig}[1]{\includegraphics[width=0.48\textwidth]{#1}}
-\newcommand{\figb}[1]{\includegraphics[width=0.95\textwidth]{#1}}
-\newcommand{\figz}[1]{\includegraphics[width=0.23\textwidth]{#1.png}}
-\newcommand{\m}[1]{\mathbf{#1}}
-\newcommand{\EE}{\mathbb{E}}
-\newcommand{\RR}{\mathbb{R}}
-\newcommand{\TT}{\mathcal{T}}
-
 \newcommand{\nhatch}[1]{{\leavevmode\color{magenta} Nathan: #1}}
-
-\usepackage{fancyhdr}
-\pagestyle{fancy}
-
-\title{Curriculum and Imitation Learning for Biped Locomotion on Rough Terrain}
-\author{TBD}
 
@@
 
+\title{Curriculum and Imitation Learning for Biped Locomotion on Rough Terrain}
+
+% You will get a Paper-ID when submitting a pdf file to the conference system
+\author{Author Names Omitted for Anonymous Review.}
+
+%\author{\authorblockN{Michael Shell}
+%\authorblockA{School of Electrical and\\Computer Engineering\\
+%Georgia Institute of Technology\\
+%Atlanta, Georgia 30332--0250\\
+%Email: mshell@ece.gatech.edu}
+%\and
+%\authorblockN{Homer Simpson}
+%\authorblockA{Twentieth Century Fox\\
+%Springfield, USA\\
+%Email: homer@thesimpsons.com}
+%\and
+%\authorblockN{James Kirk\\ and Montgomery Scott}
+%\authorblockA{Starfleet Academy\\
+%San Francisco, California 96678-2391\\
+%Telephone: (800) 555--1212\\
+%Fax: (888) 555--1212}}
+
+% avoiding spaces at the end of the author lines is not a problem with
+% conference papers because we don't use \thanks or \IEEEmembership
+
+% for over three affiliations, or if they all won't fit within the width
+% of the page, use this alternative format:
+% 
+%\author{\authorblockN{Michael Shell\authorrefmark{1},
+%Homer Simpson\authorrefmark{2},
+%James Kirk\authorrefmark{3}, 
+%Montgomery Scott\authorrefmark{3} and
+%Eldon Tyrell\authorrefmark{4}}
+%\authorblockA{\authorrefmark{1}School of Electrical and Computer Engineering\\
+%Georgia Institute of Technology,
+%Atlanta, Georgia 30332--0250\\ Email: mshell@ece.gatech.edu}
+%\authorblockA{\authorrefmark{2}Twentieth Century Fox, Springfield, USA\\
+%Email: homer@thesimpsons.com}
+%\authorblockA{\authorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
+%Telephone: (800) 555--1212, Fax: (888) 555--1212}
+%\authorblockA{\authorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
+
 \maketitle
+
+\begin{abstract}
+The abstract goes here.
+\end{abstract}
+
+\IEEEpeerreviewmaketitle
 
@@
 
-\begin{table}
+\begin{table*}
   \caption{Parameters of the trajectory generator.
@@
   \hrule
-\end{table}
+\end{table*}
 
@@
 \begin{figure}
-  \includegraphics[width=0.48\textwidth]{2D_ep.png}
-  \includegraphics[width=0.48\textwidth]{3D_ep.png}
+  \includegraphics[width=0.48\textwidth]{../figures/2D_ep.png}
+  \includegraphics[width=0.48\textwidth]{../figures/3D_ep.png}
   \caption{Screenshots of example episodes.\label{fig:example episodes}}
@@
 
+%\section*{Acknowledgments}
+
 \bibliographystyle{plainnat}
-\bibliography{bib}
+\bibliography{../bib}
 

### 2019-01-31 source/curriculum.py
Increase dist_mean to 0.45; make experiment output quieter

@@ -41,7 +41,7 @@ TRAIN_SETTINGS_2D_PLUS = {**TRAIN_SETTINGS_2D,
 
 SETTINGS_2D_EASY = {
     'use_stepping_stones': False,
-    'dist_mean': 0.42,
+    'dist_mean': 0.45,
     'dist_spread': 0.3,
     'n_steps': 16,
     'tol': 0.02,

### 2019-01-31 source/experiment.py

@@ -69,7 +69,7 @@ class Experiment:
                 self.results[settings_name][k].append(v)
             self.plot_results(settings_name)
         self.save()
-        self.learn.evaluate() # For human consumption
+        #self.learn.evaluate() # For human consumption
 
     def run_iters(self, n_iters, eval_settings, train_settings):
         self.learn.set_train_settings(train_settings)
@@ -85,7 +85,9 @@ class Experiment:
             data = np.array(self.results[settings_name][k])
             x = range(data.shape[0])
             color = self.settings[k]
-            line, = plt.plot(x, np.mean(data, 1), color=color)
+            mean = np.mean(data, 1)
+            line, = plt.plot(x, mean, color=color)
+            #plt.fill_between(x, mean-std, mean+std, color=color, alpha=0.2)
             plt.fill_between(x, np.min(data, 1), np.max(data, 1), color=color, alpha=0.2)
             labels.append(k)
             lines.append(line)
@@ -118,10 +120,9 @@ def ex_3D():
 def ex_2D():
     from stepping_stones_env import SteppingStonesEnv
     env = SteppingStonesEnv()
-    ex = Experiment(env, "2D_cur_with_ankle", ['SETTINGS_2D_EASY', 'SETTINGS_2D_HARD'])
-    ex.revert_to_iteration(29, '2D_quadratic_after_29')
+    ex = Experiment(env, "2D_quadratic_after_29", ['SETTINGS_2D_EASY', 'SETTINGS_2D_HARD'])
     #ex.run_iters(3, cur.SETTINGS_2D_EASY, cur.TRAIN_SETTINGS_2D)
-    ex.run_iters(10, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_PLUS)
+    #ex.run_iters(10, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_PLUS)
     embed()
 
 if __name__ == '__main__':

### 2019-01-31 source/inverse_dynamics.py

@@ -90,7 +90,7 @@ class LearnInverseDynamics:
         experience = []
         s = self.train_settings
         for i in range(s['n_trajectories']):
-            r = self.evaluate(max_intolerable_steps=s['max_intolerable_steps'])
+            r = self.evaluate(max_intolerable_steps=s['max_intolerable_steps'], render=None)
             experience += self.evaluator.experience
             self.total_steps += r['n_steps']
         return experience
@@ -126,7 +126,7 @@ class LearnInverseDynamics:
         runner.reset()
         rs.w_policy = self.act(features) # Initialize with something reasonable
         # TODO put max_iters and tol in the object initialization params instead
-        w_policy = rs.random_search(render=1)
+        w_policy = rs.random_search(render=None)
         self.total_steps += runner.n_runs
         return w_policy
 

### 2019-01-31 source/simbicon.py

@@ -206,8 +206,8 @@ class Simbicon(PDController):
             if self.env.is_3D:
                 self.params[SWING_ANKLE_ROLL] = self.base_gait()[SWING_ANKLE_ROLL]
         early_strike = (duration >= LIFTOFF_DURATION) and (len(contacts) > 0)
-        if early_strike:
-            print("Early strike!")
+        #if early_strike:
+        #    print("Early strike!")
         q, dq = self.env.get_x()
         target_diff = self.params[IK_GAIN] * self.speed(dq)
         heel_close = self.distance_to_go(swing_heel) < target_diff

### 2019-01-31 source/curriculum.py
Permit curriculum to change model class

@@ -28,6 +28,7 @@ TRAIN_SETTINGS_2D = {
             sp.UP_IDX+sp.SWING_KNEE_RELATIVE,
             ]),
     'observable_features': o2d,
+    'model_class': 'linear',
     }
 
 params_2Dplus = TRAIN_SETTINGS_2D['controllable_params'].copy()
@@ -37,6 +38,7 @@ o2d_plus = np.ones_like(o2d) # All features are observable
 TRAIN_SETTINGS_2D_PLUS = {**TRAIN_SETTINGS_2D,
     'controllable_params': params_2Dplus,
     'observable_features': o2d_plus,
+    'model_class': 'quadratic',
     }
 
 SETTINGS_2D_EASY = {
@@ -82,6 +84,7 @@ TRAIN_SETTINGS_3D = {
             sp.UP_IDX+sp.SWING_HIP_WORLD,
             ]),
     'observable_features': o3d,
+    'model_class': 'linear',
     }
 
 TRAIN_SETTINGS_3D_PRECISE = {**TRAIN_SETTINGS_3D,

### 2019-01-31 source/experiment.py

@@ -120,9 +120,10 @@ def ex_3D():
 def ex_2D():
     from stepping_stones_env import SteppingStonesEnv
     env = SteppingStonesEnv()
-    ex = Experiment(env, "2D_quadratic_after_29", ['SETTINGS_2D_EASY', 'SETTINGS_2D_HARD'])
-    #ex.run_iters(3, cur.SETTINGS_2D_EASY, cur.TRAIN_SETTINGS_2D)
-    #ex.run_iters(10, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_PLUS)
+    ex = Experiment(env, "cim_final_1", ['SETTINGS_2D_EASY', 'SETTINGS_2D_HARD'])
+    ex.run_iters(3, cur.SETTINGS_2D_EASY, cur.TRAIN_SETTINGS_2D)
+    ex.run_iters(6, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D)
+    ex.run_iters(9, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_PLUS)
     embed()
 
 if __name__ == '__main__':

### 2019-01-31 source/inverse_dynamics.py

@@ -25,19 +25,19 @@ class LearnInverseDynamics:
         self.name = name
         self.train_features, self.train_responses = [], []
         self.history = []
-        # TODO try some model more complicated than linear?
-        model = Ridge(alpha=RIDGE_ALPHA, fit_intercept=False)
-        #model = LinearRegression(fit_intercept=False)
-        #model = make_pipeline(PolynomialFeatures(2), model) # quadratic
-        #model = RANSACRegressor(base_estimator=model, residual_threshold=2.0)
-        self.model = model
-        self.is_fitted = False # For some dang reason sklearn doesn't track this itself
         self.total_steps = 0
         self.total_failed_annotations = 0
         self.total_train_time = 0.0
 
     def set_train_settings(self, settings):
         self.train_settings = settings
+        model = Ridge(alpha=RIDGE_ALPHA, fit_intercept=False)
+        #model = LinearRegression(fit_intercept=False)
+        #model = RANSACRegressor(base_estimator=model, residual_threshold=2.0)
+        if settings['model_class'] == 'quadratic':
+            model = make_pipeline(PolynomialFeatures(2), model)
+        self.model = model
+        self.train_inverse_dynamics()
 
     def dump_train_set(self):
         fname = TRAIN_FMT.format(self.name)
@@ -65,7 +65,6 @@ class LearnInverseDynamics:
         self.dump_train_set()
         self.evaluator.set_eval_settings(h[4])
         self.set_train_settings(h[5])
-        self.train_inverse_dynamics()
 
     def training_iter(self):
         print("STARTING TRAINING ITERATION", len(self.history))
@@ -98,14 +97,14 @@ class LearnInverseDynamics:
     def expert_annotate(self, experience):
         self.env.clear_skeletons()
         total = len(experience)
-        for i, (state, _, reward) in enumerate(experience):
+        for i, (state, action, reward) in enumerate(experience):
             print("Finding expert label for state {}/{}".format(i, total))
             if reward < 1-self.train_settings['tol']:
-                self.label_state(state)
+                self.label_state(state, action)
         self.env.clear_skeletons()
 
-    def label_state(self, features):
-        mean_action = self.learn_action(features)
+    def label_state(self, features, action):
+        mean_action = self.learn_action(features, action)
         if mean_action is None:
             # Random search couldn't find a good enough action; don't use this for training.
             self.total_failed_annotations += 1
@@ -118,34 +117,32 @@ class LearnInverseDynamics:
         # during random search.
[…]

### 2019-02-01 source/experiment.py
Fix bug: ensure save directory exists

@@ -11,7 +11,6 @@ from inverse_dynamics import LearnInverseDynamics
 import curriculum as cur
 
 DIR_FMT = 'data/{}/'
-RESULTS_FMT = DIR_FMT + 'results.pkl'
 
 N_EVAL_TRAJECTORIES = 8
 KEYS_TO_SAVE = ['total_reward', 'max_error', 'n_steps', 'seed']
@@ -29,16 +28,19 @@ class Experiment:
         self.n_eval_trajectories = N_EVAL_TRAJECTORIES
         self.load(final_eval_settings)
 
-    def save(self):
+    def save_filename(self, filename):
         dirname = DIR_FMT.format(self.name)
         if not os.path.exists(dirname):
             os.mkdir(dirname)
-        fname = RESULTS_FMT.format(self.name)
+        return dirname + filename
+
+    def save(self):
+        fname = self.save_filename('results.pkl')
         with open(fname, 'wb') as f:
             pickle.dump(self.results, f)
 
     def load(self, final_eval_settings):
-        fname = RESULTS_FMT.format(self.name)
+        fname = self.save_filename('results.pkl')
         if os.path.exists(fname):
             with open(fname, 'rb') as f:
                 self.results = pickle.load(f)
@@ -106,7 +108,7 @@ class Experiment:
         sns.set_style('white')
         sns.despine()
 
-        plt.savefig('data/{}/{}.png'.format(self.name, settings_name))
+        plt.savefig(self.save_filename('{}.png'.format(settings_name)))
         plt.clf()
 
 def ex_3D(uq_id):

### 2019-02-01 source/curriculum.py
Add "no curriculum" and "no expert" baseline experiments
I'm still fiddling around to find some fair hyperparameter settings for
these comparisons; it's kind of apples-to-oranges. Other changes:

- Put `reward` in utils.py. This means RandomSearch and Evaluator now
both use the *same* (shaped) reward function.
- Change Experiment plotting code so X axis now gives total number of
simulated footsteps taken (and we don't plot max_error anymore).
- Change internal interface of RandomSearch . . . because the random
search baseline directly calls `estimate_grad` rather than the official
API `random_

@@ -33,10 +33,32 @@ TRAIN_SETTINGS_2D = {
     'model_class': 'linear',
     }
 
+# TODO do this for 3D as well
+cp_orig = TRAIN_SETTINGS_2D['controllable_params']
+col = cp_orig.reshape((-1,1))
+row = np.concatenate((o2d, [True])).reshape((1,-1))
+cp_baseline_noexpert = np.dot(col, row) # Outer product
+
+TRAIN_SETTINGS_BASELINE_NOEXPERT = {
+    'n_dirs': 4,
+    'tol': 0.00, # Not actually used
+    'controllable_params': cp_baseline_noexpert,
+    'rs_eps': 0.1,
+    'rs_step_size': 0.01,
+    }
+
 o2d_plus = np.ones_like(o2d) # All features are observable
 
 TRAIN_SETTINGS_2D_PLUS = {**TRAIN_SETTINGS_2D,
     'model_class': 'quadratic',
+    }
+
+TRAIN_SETTINGS_2D_NOCUR_FIRST = {**TRAIN_SETTINGS_2D_PLUS,
+    'n_trajectories': 80,
+    }
+
+TRAIN_SETTINGS_2D_NOCUR_NEXT = {**TRAIN_SETTINGS_2D_PLUS,
+    'n_trajectories': 4,
     }
 
 SETTINGS_2D_EASY = {

### 2019-02-01 source/evaluator.py

@@ -1,4 +1,5 @@
 import numpy as np
+import utils
 
 class Evaluator:
     def __init__(self, env):
@@ -36,7 +37,7 @@ class Evaluator:
             action = policy(features)
             end_state, terminated = self.env.simulate(target, target_heading=0.0, action=action)
             error = np.linalg.norm(end_state.stance_heel_location() - target)
-            reward = 1-error
+            reward = utils.reward(self.env.controller, end_state)
             total_reward += reward
             # In RL terms, (state,target) is the state.
             experience.append((state.extract_features(target), action, reward))

### 2019-02-01 source/experiment.py

@@ -8,23 +8,24 @@ import matplotlib.pyplot as plt
 import seaborn as sns
 from collections import defaultdict
 from inverse_dynamics import LearnInverseDynamics
+from rs_baseline import RandomSearchBaseline
 import curriculum as cur
 
 DIR_FMT = 'data/{}/'
 
 N_EVAL_TRAJECTORIES = 8
 KEYS_TO_SAVE = ['total_reward', 'max_error', 'n_steps', 'seed']
-KEYS_TO_PLOT = ['total_reward', 'max_error']
+KEYS_TO_PLOT = ['total_reward']
 
 class Experiment:
-    def __init__(self, env, name, final_eval_settings):
+    def __init__(self, env, learner_class, name, final_eval_settings):
         self.settings = {
                 "total_reward": "blue",
                 "max_error": "red",
                 "n_steps": "black",
                 }
         self.name = name
-        self.learn = LearnInverseDynamics(env, self.name)
+        self.learn = learner_class(env, self.name)
         self.n_eval_trajectories = N_EVAL_TRAJECTORIES
         self.load(final_eval_settings)
 
@@ -91,7 +92,10 @@ class Experiment:
         labels = []
         for k in KEYS_TO_PLOT:
             data = np.array(self.results[settings_name][k])
-            x = range(data.shape[0])
+            # Index `1` gives total_steps for both RandomSearchBaseline and LearnInverseDynamics.
+            # TODO make this interface more robust/obvious.
+            x = map(lambda h: h[1], self.learn.history)
+            x = [0] + list(x)
             color = self.settings[k]
             mean = np.mean(data, 1)
             line, = plt.plot(x, mean, color=color)
@@ -101,8 +105,8 @@ class Experiment:
             lines.append(line)
 
         plt.title(settings_name)
-        plt.xlabel("Number of data collection iterations")
-        plt.ylabel("Foot placement error")
+        plt.xlabel("Total number of simulated footsteps taken")
[…]

### 2019-02-01 source/random_search.py

@@ -15,17 +15,17 @@ class RandomSearch:
         self.max_iters = settings.get('max_iters') or 5
         self.controllable_params = settings['controllable_params']
 
-    def sample_perturbation(self):
-        delta = np.random.randn(np.prod(self.w_policy.shape))
+    def sample_perturbation(self, policy):
+        delta = np.random.randn(np.prod(policy.shape)).reshape(policy.shape)
         delta *= self.controllable_params
         delta /= np.linalg.norm(delta)
-        return delta.reshape(self.w_policy.shape)
+        return delta
 
     def estimate_grad(self, policy):
         grad = np.zeros_like(policy)
         rets = []
         for j in range(self.n_dirs):
-            p = self.sample_perturbation()
+            p = self.sample_perturbation(policy)
             self.runner.reset()
             p_ret = self.runner.run(policy + self.step_size*p)
             self.runner.reset()
@@ -56,5 +56,5 @@ class RandomSearch:
         # The idea is, the closer we are to the target, the smaller our next
         # step size should be. (But when we are very far from the target, we should
         # still use a reasonably small step size to avoid divergence.)
-        self.step_size = min(-ret, 0.3)
-        return ret > -self.tol
+        self.step_size = min(-(ret-1), 0.3)
+        return ret > 1-self.tol

### 2019-02-01 source/rs_baseline.py

@@ -1,0 +1,105 @@
+from IPython import embed
+import numpy as np
+import pickle
+import os
+import time
+
+import simbicon_params as sp
+from state import reconstruct_state
+from step_learner import Runner
+from random_search import RandomSearch
+from evaluator import Evaluator
+
+TRAIN_FMT = 'data/{}/history.pkl'
+import curriculum as cur
+
+def coef_to_fn(coef):
+    def act(features):
+        X = np.concatenate((features, [1.0]))
+        return np.dot(coef, X)
+    return act
+
+class EpisodeRunner:
+    def __init__(self, evaluator):
+        self.evaluator = evaluator
+        self.n_steps = 0
+        self.seed = np.random.randint(100000)
+
+    def run(self, coef):
+        policy = coef_to_fn(coef)
+        # TODO use shaped total reward
+        r = self.evaluator.evaluate(policy,
+                seed=self.seed, render=self.render, video_save_dir=self.video_save_dir)
+        ret = r['total_reward'] / 16 - 1 # Max ret of 0, min of -1
+        self.n_steps += r['n_steps']
+        return ret
+
+    def reset(self, video_save_dir=None, render=None):
+        self.video_save_dir = video_save_dir
+        self.render = render
+
+class RandomSearchBaseline:
+    def __init__(self, env, name=''):
+        self.env = env
+        self.evaluator = Evaluator(env)
+        self.name = name
+        self.history = []
+        self.total_steps = 0
+        self.total_train_time = 0.0
+        self.coef_ = np.zeros_like(cur.TRAIN_SETTINGS_BASELINE_NOEXPERT['controllable_params'], dtype=np.float32)
+
+    def set_train_settings(self, settings):
+        self.train_settings = settings
+        # TODO implement switching to a quadratic model....
+        # ...I have no idea how one would do this
+
+    def dump_train_set(self):
+        fname = TRAIN_FMT.format(self.name)
+        with open(fname, 'wb') as f:
+            pickle.dump(self.history, f)
+
+    def load_train_set(self):
+        fname = TRAIN_FMT.format(self.name)
+        with open(fname, 'rb') as f:
+            self.history = pickle.load(f)
+        self.revert_to_iteration(len(self.history), self.name)
+
+    def revert_to_iteration(self, iteration, new_name):
+        # Resets so the next iteration index will be `iteration`
+        self.history = self.history[:iteration]
+        h = self.history[-1]
+        self.coef_ = h[0]
+        self.total_steps = h[1]
[…]

### 2019-02-01 source/step_learner.py

@@ -3,6 +3,7 @@ import numpy as np
 from IPython import embed
 from random_search import RandomSearch
 from curriculum import *
+import utils
 
 class Runner:
     def __init__(self, env, start_state, target, use_stepping_stones=True):
@@ -17,18 +18,9 @@ class Runner:
     def run(self, action):
         self.n_runs += 1
         r, _ = self.env.simulate(self.target, target_heading=0.0, action=action, put_dots=True)
-        score = -np.linalg.norm(r.stance_heel_location() - r.stance_platform())
-        # Having the COM in front of the stance foot is bad
-        com_dist = self.env.controller.distance_to_go(r.pose()[:3])
-        heel_dist = self.env.controller.distance_to_go(r.stance_heel_location())
-        if com_dist - heel_dist < 0.0:
-            # TODO probably better to use stance heel for this penalty, not the target itself
-            score += com_dist - heel_dist
-        if com_dist - heel_dist > 0.2:
-            score -= com_dist - heel_dist - 0.2
         if self.env.video_recorder is not None:
             self.env.pause(0.3)
-        return score
+        return utils.reward(self.env.controller, r)
 
     def reset(self, video_save_dir=None, render=None):
         self.env.reset(self.start_state, video_save_dir=video_save_dir, render=render)

### 2019-02-01 source/utils.py

@@ -17,6 +17,17 @@ def heading_from_vector(d, branch=0.0):
             heading += k*2*np.pi
         return planar_d, heading
 
+def reward(controller, state):
+    score = 1.0
+    score -= np.linalg.norm(state.stance_heel_location() - state.stance_platform())
+    # Balance penalty: Having the COM far away from the stance foot is bad
+    com_dist = controller.distance_to_go(state.pose()[:3])
+    heel_dist = controller.distance_to_go(state.stance_heel_location())
+    if heel_dist - com_dist > 0.0: # COM is past the heel
+        score -= heel_dist - com_dist
+    if com_dist - heel_dist > 0.2: # COM is far behind the heel
+        score -= com_dist - heel_dist - 0.2
+    return score
 
 def rotmatrix(theta):
     # Note we're rotating in the X-Z plane instead of X-Y, so some signs are weird.

### 2019-02-01 source/curriculum.py
Pick some reasonable hyperparameters for the baselines

@@ -40,7 +40,7 @@ row = np.concatenate((o2d, [True])).reshape((1,-1))
 cp_baseline_noexpert = np.dot(col, row) # Outer product
 
 TRAIN_SETTINGS_BASELINE_NOEXPERT = {
-    'n_dirs': 4,
+    'n_dirs': 8,
     'tol': 0.00, # Not actually used
     'controllable_params': cp_baseline_noexpert,
     'rs_eps': 0.1,
@@ -54,11 +54,11 @@ TRAIN_SETTINGS_2D_PLUS = {**TRAIN_SETTINGS_2D,
     }
 
 TRAIN_SETTINGS_2D_NOCUR_FIRST = {**TRAIN_SETTINGS_2D_PLUS,
-    'n_trajectories': 80,
+    'n_trajectories': 32,
     }
 
 TRAIN_SETTINGS_2D_NOCUR_NEXT = {**TRAIN_SETTINGS_2D_PLUS,
-    'n_trajectories': 4,
+    'n_trajectories': 16,
     }
 
 SETTINGS_2D_EASY = {

### 2019-02-01 source/experiment.py

@@ -137,12 +137,19 @@ def ex_2D_cim(uq_id):
     ex.run_iters(12, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_PLUS)
     embed()
 
+def ex_2D_cim_easy(uq_id):
+    from stepping_stones_env import SteppingStonesEnv
+    env = SteppingStonesEnv()
+    ex = Experiment(env, LearnInverseDynamics, "cim_final_easy_"+uq_id, ['SETTINGS_2D_EASY'])
+    ex.run_iters(4, cur.SETTINGS_2D_EASY, cur.TRAIN_SETTINGS_2D)
+    embed()
+
 def ex_2D_nocur(uq_id):
     from stepping_stones_env import SteppingStonesEnv
     env = SteppingStonesEnv()
     ex = Experiment(env, LearnInverseDynamics, "nocur_final_"+uq_id, ['SETTINGS_2D_HARD'])
     ex.run_iters(1, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_NOCUR_FIRST)
-    ex.run_iters(10, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_NOCUR_NEXT)
+    ex.run_iters(8, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_NOCUR_NEXT)
     embed()
 
 def ex_2D_rs(uq_id):
@@ -154,4 +161,4 @@ def ex_2D_rs(uq_id):
 
 if __name__ == '__main__':
     UQ_ID = sys.argv[1]
-    ex_2D_rs(UQ_ID)
+    ex_2D_cim_easy(UQ_ID)

### 2019-02-01 source/rs_baseline.py

@@ -89,7 +89,7 @@ class RandomSearchBaseline:
             self.total_steps,
             self.total_train_time,
             self.evaluator.eval_settings,
-            self.train_settings))
+            self.train_settings.copy()))
         print("Total steps:", self.total_steps)
         self.dump_train_set()
 
@@ -100,6 +100,8 @@ class RandomSearchBaseline:
         grad = rs.estimate_grad(self.coef_)
         self.coef_ += self.train_settings['rs_step_size'] * grad
         self.total_steps += runner.n_steps
+        self.train_settings['rs_eps'] *= 0.9
+        self.train_settings['rs_step_size'] *= 0.9
 
     def evaluate(self, **kwargs):
         return self.evaluator.evaluate(coef_to_fn(self.coef_), **kwargs)

### 2019-02-01 source/inverse_dynamics.py
FIX BUG: inverse_dynamics was unnecessarily throwing away training points
The change introduced in 58e44d2 to use *all* rollout footsteps for
training (even the ones that were already good without expert
annotation) was accidentally reverted in eefcbc6 (which was just trying
to speedup computation by skipping the expert annotation for those
footsteps).

Unfortunately that means I will probably need to rerun the CIM
experiments :( or I could revert this bugfix until after RSS...not sure.

@@ -100,22 +100,17 @@ class LearnInverseDynamics:
         for i, (state, action, reward) in enumerate(experience):
             print("Finding expert label for state {}/{}".format(i, total))
             if reward < 1-self.train_settings['tol']:
-                self.label_state(state, action)
+                action = self.learn_action(state, action)
+            if action is None:
+                # Random search couldn't find a good enough action; don't use this for training.
+                self.total_failed_annotations += 1
+            else:
+                # The action is good enough; use this (state, action) pair for training.
+                self.train_features.append(state)
+                # We don't need to mirror the action, because Simbicon3D already handled that
+                # during random search.
+                self.train_responses.append(action)
         self.env.clear_skeletons()
-
-    def label_state(self, features, action):
-        mean_action = self.learn_action(features, action)
-        if mean_action is None:
-            # Random search couldn't find a good enough action; don't use this for training.
-            self.total_failed_annotations += 1
-            return
-        # The train set is a set of (features, action) pairs
-        # where taking `action` when the environment features are `features`
-        # will hit the target with the swing foot (within 5 cm).
-        self.train_features.append(features)
-        # We don't need to mirror the action, because Simbicon3D already handled that
-        # during random search.
-        self.train_responses.append(mean_action)
 
     def learn_action(self, features, action):
         start_state, target = reconstruct_state(features, self.env.consts())

### 2019-02-01 source/experiment.py
Move results-plotting code to separate file
In preparation for generating the figures for the actual paper.

@@ -4,26 +4,19 @@ import numpy as np
 import pickle
 import os.path
 import os
-import matplotlib.pyplot as plt
-import seaborn as sns
 from collections import defaultdict
 from inverse_dynamics import LearnInverseDynamics
 from rs_baseline import RandomSearchBaseline
 import curriculum as cur
+import plotter
 
 DIR_FMT = 'data/{}/'
 
 N_EVAL_TRAJECTORIES = 8
 KEYS_TO_SAVE = ['total_reward', 'max_error', 'n_steps', 'seed']
-KEYS_TO_PLOT = ['total_reward']
 
 class Experiment:
     def __init__(self, env, learner_class, name, final_eval_settings):
-        self.settings = {
-                "total_reward": "blue",
-                "max_error": "red",
-                "n_steps": "black",
-                }
         self.name = name
         self.learn = learner_class(env, self.name)
         self.n_eval_trajectories = N_EVAL_TRAJECTORIES
@@ -39,6 +32,10 @@ class Experiment:
         fname = self.save_filename('results.pkl')
         with open(fname, 'wb') as f:
             pickle.dump(self.results, f)
+        self.plot_results()
+
+    def plot_results(self):
+        plotter.plot_results(self.results, self.learn.history, self.save_filename(''))
 
     def load(self, final_eval_settings):
         fname = self.save_filename('results.pkl')
@@ -76,7 +73,6 @@ class Experiment:
                     results[k].append(result[k])
             for k,v in results.items():
                 self.results[settings_name][k].append(v)
-            self.plot_results(settings_name)
         self.save()
         #self.learn.evaluate() # For human consumption
 
@@ -86,34 +82,6 @@ class Experiment:
             self.learn.evaluator.set_eval_settings(eval_settings)
             self.learn.training_iter()
             self.run_evaluations()
-
-    def plot_results(self, settings_name):
-        lines = []
-        labels = []
-        for k in KEYS_TO_PLOT:
-            data = np.array(self.results[settings_name][k])
-            # Index `1` gives total_steps for both RandomSearchBaseline and LearnInverseDynamics.
-            # TODO make this interface more robust/obvious.
-            x = map(lambda h: h[1], self.learn.history)
-            x = [0] + list(x)
-            color = self.settings[k]
[…]

### 2019-02-01 source/plotter.py

@@ -1,0 +1,40 @@
+
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+KEYS_TO_PLOT = ['total_reward']
+SETTINGS = {
+        "total_reward": "red",
+        "max_error": "red",
+        "n_steps": "black",
+        }
+
+def plot_results(results, history, save_dir):
+    for settings_name in results.keys():
+        lines = []
+        labels = []
+        for k in KEYS_TO_PLOT:
+            data = np.array(results[settings_name][k])
+            # Index `1` gives total_steps for both RandomSearchBaseline and LearnInverseDynamics.
+            # TODO make this interface more robust/obvious.
+            x = map(lambda h: h[1], history)
+            x = [0] + list(x)
+            color = SETTINGS[k]
+            mean = np.mean(data, 1)
+            line, = plt.plot(x, mean, color=color)
+            #plt.fill_between(x, mean-std, mean+std, color=color, alpha=0.2)
+            plt.fill_between(x, np.min(data, 1), np.max(data, 1), color=color, alpha=0.2)
+            labels.append(k)
+            lines.append(line)
+
+        plt.title(settings_name)
+        plt.xlabel("Total number of simulated footsteps taken")
+        plt.ylabel("Total reward")
+        plt.legend(lines, labels)
+
+        sns.set_style('white')
+        sns.despine()
+
+        plt.savefig(save_dir + '{}.png'.format(settings_name))
+        plt.clf()

### 2019-02-01 source/experiment.py
Code to generate baseline-comparison figures for paper

@@ -8,7 +8,6 @@ from collections import defaultdict
 from inverse_dynamics import LearnInverseDynamics
 from rs_baseline import RandomSearchBaseline
 import curriculum as cur
-import plotter
 
 DIR_FMT = 'data/{}/'
 
@@ -32,10 +31,6 @@ class Experiment:
         fname = self.save_filename('results.pkl')
         with open(fname, 'wb') as f:
             pickle.dump(self.results, f)
-        self.plot_results()
-
-    def plot_results(self):
-        plotter.plot_results(self.results, self.learn.history, self.save_filename(''))
 
     def load(self, final_eval_settings):
         fname = self.save_filename('results.pkl')
@@ -129,4 +124,4 @@ def ex_2D_rs(uq_id):
 
 if __name__ == '__main__':
     UQ_ID = sys.argv[1]
-    ex_2D_cim_easy(UQ_ID)
+    ex_2D_cim(UQ_ID)

### 2019-02-01 source/inverse_dynamics.py

@@ -48,7 +48,8 @@ class LearnInverseDynamics:
         fname = TRAIN_FMT.format(self.name)
         with open(fname, 'rb') as f:
             self.history, self.train_features, self.train_responses = pickle.load(f)
-        self.revert_to_iteration(len(self.history), self.name)
+        if self.env is not None:
+            self.revert_to_iteration(len(self.history), self.name)
 
     def revert_to_iteration(self, iteration, new_name):
         # Resets so the next iteration index will be `iteration`

### 2019-02-01 source/plotter.py

@@ -1,40 +1,80 @@
-
+from IPython import embed
+from experiment import Experiment
+from inverse_dynamics import LearnInverseDynamics
+from rs_baseline import RandomSearchBaseline
 import numpy as np
 import matplotlib.pyplot as plt
 import seaborn as sns
 
 KEYS_TO_PLOT = ['total_reward']
 SETTINGS = {
-        "total_reward": "red",
-        "max_error": "red",
-        "n_steps": "black",
+        "cim_final_easy": ["green", "CIM", LearnInverseDynamics, [5,6,7,8]],
+        "cim_final": ["green", "CIM", LearnInverseDynamics, [1,2,3,4]],
+        "rs_final": ["purple", "ARS baseline", RandomSearchBaseline, [13,14,15,16]],
+        "nocur_final": ["blue", "No curriculum", LearnInverseDynamics, [9,10,11,12]],
         }
 
-def plot_results(results, history, save_dir):
-    for settings_name in results.keys():
-        lines = []
-        labels = []
-        for k in KEYS_TO_PLOT:
-            data = np.array(results[settings_name][k])
-            # Index `1` gives total_steps for both RandomSearchBaseline and LearnInverseDynamics.
-            # TODO make this interface more robust/obvious.
-            x = map(lambda h: h[1], history)
-            x = [0] + list(x)
-            color = SETTINGS[k]
-            mean = np.mean(data, 1)
-            line, = plt.plot(x, mean, color=color)
-            #plt.fill_between(x, mean-std, mean+std, color=color, alpha=0.2)
-            plt.fill_between(x, np.min(data, 1), np.max(data, 1), color=color, alpha=0.2)
-            labels.append(k)
-            lines.append(line)
+def fill_plot(x, data, color):
+    line, = plt.plot(x, np.percentile(data, 50, 1), color=color)
+    #plt.fill_between(x, mean-std, mean+std, color=color, alpha=0.2)
+    plt.fill_between(x, np.percentile(data, 5, 1), np.percentile(data, 95, 1),
+            color=color, alpha=0.2)
+    return line
 
-        plt.title(settings_name)
-        plt.xlabel("Total number of simulated footsteps taken")
-        plt.ylabel("Total reward")
-        plt.legend(lines, labels)
+def steps_from_history(history):
[…]

### 2019-02-01 source/rs_baseline.py

@@ -62,7 +62,8 @@ class RandomSearchBaseline:
         fname = TRAIN_FMT.format(self.name)
         with open(fname, 'rb') as f:
             self.history = pickle.load(f)
-        self.revert_to_iteration(len(self.history), self.name)
+        if self.env is not None:
+            self.revert_to_iteration(len(self.history), self.name)
 
     def revert_to_iteration(self, iteration, new_name):
         # Resets so the next iteration index will be `iteration`

### 2019-02-01 paper/rss/main.tex
Add baseline plots to paper

@@
 These approaches tend to generate \emph{hierarchical} plans in order to reduce computational requirements.
-First, they choose footstep locations using search, like A*.\nhatch{citation??}
+First, they choose footstep locations using search, like A* \citep{huang2013step}.
 Based on the footstep plan, they generate a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum \citep{mordatch2010robust} or contact wrench cone \citep{dai2016planning}.
@@
 
-\nhatch{Honda and Boston Dynamics do a lot with dynamic gaits, but I don't really understand how it works. To what extent can they plan step locations and handle rough terrain?}
+%\nhatch{Honda and Boston Dynamics do a lot with dynamic gaits, but I don't really understand how it works. To what extent can they plan step locations and handle rough terrain?}
 
@@
 \begin{table*}
-  \caption{Parameters of the trajectory generator.
-  Note that \texttt{stance\_hip\_roll} is used only in 3D, and \texttt{swing\_knee} is used only in 2D (since in 3D we don't attempt environments with stairs).}
+  \normalsize
+  \caption{Parameters of the trajectory generator.}
   \label{table:params}
@@
     \texttt{swing\_hip} & Angle for swing hip during UP state (saggital plane). \\
-    \texttt{swing\_knee} & Angle for swing knee during UP state (saggital plane). \\
-    \texttt{stance\_hip\_roll} & Angle for stance hip during all three states (coronal plane). \\
+    \texttt{swing\_knee} & 2D only: Angle for swing knee during UP state (saggital plane). \\
+    \texttt{stance\_hip\_roll} & 3D only: Angle for stance hip during all three states (coronal plane). \\
   \end{tabular}
@@
 For general RL problems, this would be difficult because of the infamous \emph{credit assignment problem}:
-sometimes, an action during one period of the episode can leave the robot in a bad position to achieve the desired result for the next period.
+sometimes, an action in an early part of an episode can leave the robot in a bad position to achieve the desired result in later parts.
 %Such an action is in fact a bad action, even if the robot does achieve the desired result during the earlier period.
@@
 Hence, it makes sense to scale the difficulty of the expert optimization problem roughly proportionally to the current capabilities of the HLP.
-This is similar to other slow-expert/fast-learner approaches such as ExIt \citep{anthony2017thinking} and AlphaGo Zero \citep{silver2017mastering}.
+This is similar to other slow-expert/fast-learner algorithms such as ExIt \citep{anthony2017thinking} and AlphaGo Zero \citep{silver2017mastering}.
 
@@
 
-For each approach (CIM and baseline) we run the experiment from scratch with four different random seeds.
-Results are shown in Fig.
-TODO.
+Based on a manual hyperparameter search, we decided that the baseline would use $Q=8$ random directions and a step size of $0.01$ decaying by $0.9$ at each iteration.
+Furthermore, we decided that all $2Q$ rollouts during each iteration of ARS should use the same random seed (i.e. the same placement of stepping stones).
+This reduces the variance of the gradient estimate in the sense that for this particular seed, the gradient estimate is more likely to be a good descent direction.
+The random seed changes between iterations so that the learning algorithm gets to see the full distribution of testing environments.
+
+For each approach (CIM and baseline) we run the experiment from scratch with four different starting random seeds.
+Results comparing its performance to CIM on 2D-Easy are shown in Fig. \ref{fig:ars baseline}.
 The X axis shows the total number of simulated footsteps taken during training, and the Y axis shows total reward.
-We plot the mean and standard deviation over 32 testing episodes (eight episodes for each of the four experiments).
+We plot the median and 5th/95th percentiles over 32 testing episodes (eight episodes for each of the four starting random seeds).
+This ARS baseline learned very slowly even for the 2D-Easy environment, so we did not attempt to evaluate it on 2D-hard.
+
+\begin{figure}
+  \includegraphics[width=0.48\textwidth]{../figures/ars_baseline.pdf}
+  \caption{\label{fig:ars baseline} Training curves on 2D-Easy environment.}
+\end{figure}
 
@@
 Instead, the baseline begins training in the final, most difficult testing environment.
-Results are shown in Fig. TODO.
+Results are shown in Fig. \ref{fig:nocur baseline}.
+
+\begin{figure}
+  \includegraphics[width=0.48\textwidth]{../figures/nocur_baseline.pdf}
+  \caption{\label{fig:nocur baseline} Training curves on 2D-Hard environment.}
+\end{figure}
 
@@
 In future work, we plan to investigate how to transfer policies trained with CIM to real-world bipedal robots.
+
+\nhatch{future work: adaptive curriculum (adjusts to learner's current ability)}
 

### 2019-02-01 source/plotter.py

@@ -6,6 +6,9 @@ import numpy as np
 import matplotlib.pyplot as plt
 import seaborn as sns
 
+import matplotlib
+matplotlib.rcParams.update({'font.size': 14})
+
 KEYS_TO_PLOT = ['total_reward']
 SETTINGS = {
         "cim_final_easy": ["green", "CIM", LearnInverseDynamics, [5,6,7,8]],
@@ -64,7 +67,6 @@ def gen_figures():
         lines.append(multiseed_plot(exp))
         labels.append(SETTINGS[exp][1])
     plt.legend(lines, labels)
-    plt.title('Training curves on 2D-Easy environment')
     save_plot('../paper/figures/ars_baseline.pdf')
 
     lines = []
@@ -73,7 +75,6 @@ def gen_figures():
         lines.append(multiseed_plot(exp))
         labels.append(SETTINGS[exp][1])
     plt.legend(lines, labels, loc='lower right')
-    plt.title('Training curves on 2D-Hard environment')
     save_plot('../paper/figures/nocur_baseline.pdf')
 
 if __name__ == '__main__':

### 2019-02-02 paper/rss/main.tex
Updates to ARS ablations section

@@
 That is, at each iteration we update the HLP directly by estimating the gradient as a weighted combination of $Q$ random directions,
-with weights determined by the change in total reward as defined in section \ref{sec:expert}.
+with weights determined by the change in total shaped reward as defined in section \ref{sec:expert}.
 Note that this baseline is very similar to PMTG \citep{iscen2018pmtg}.
@@
 Based on a manual hyperparameter search, we decided that the baseline would use $Q=8$ random directions and a step size of $0.01$ decaying by $0.9$ at each iteration.
-Furthermore, we decided that all $2Q$ rollouts during each iteration of ARS should use the same random seed (i.e. the same placement of stepping stones).
+Furthermore, we decided that all $2Q$ rollouts during each iteration of ARS would use the same random seed (i.e. the same placement of stepping stones).
 This reduces the variance of the gradient estimate in the sense that for this particular seed, the gradient estimate is more likely to be a good descent direction.
 The random seed changes between iterations so that the learning algorithm gets to see the full distribution of testing environments.
+Results were comparable or worse for the other hyperparameter settings that we tried.
 
 For each approach (CIM and baseline) we run the experiment from scratch with four different starting random seeds.
-Results comparing its performance to CIM on 2D-Easy are shown in Fig. \ref{fig:ars baseline}.
+Results on the 2D-Easy environment are shown in Fig. \ref{fig:ars baseline}.
 The X axis shows the total number of simulated footsteps taken during training, and the Y axis shows total reward.
 We plot the median and 5th/95th percentiles over 32 testing episodes (eight episodes for each of the four starting random seeds).
-This ARS baseline learned very slowly even for the 2D-Easy environment, so we did not attempt to evaluate it on 2D-hard.
 
@@
 \end{figure}
+
+Taking advantage of simplifications to the credit assignment problem seems to be very helpful for this task.
+Since the ARS baseline learns very slowly even on the 2D-Easy environment, we did not attempt to evaluate it on 2D-Hard.
+In future work, to further validate the importance of imitation learning for periodic tasks, it would be interesting to compare CIM against a reinforcement learning algorithm such as PPO \citep{schulman2017proximal}.
 

### 2019-02-02 paper/rss/main.tex
Updates to no-curriculum-baseline section

@@
 In this section, we evaluate CIM against a baseline that does not use curriculum learning.
-Instead, the baseline begins training in the final, most difficult testing environment.
-Results are shown in Fig. \ref{fig:nocur baseline}.
+Instead, the baseline begins training with a quadratic model in the final, most difficult testing environment.
+
+Due to the sample complexity of the quadratic model, making this a fair comparison requires changing the training schedule used for imitation learning.
+For the CIM experiments, we retrained the model every four rollouts,
+but with a quadratic model this leads to disastrous overfitting and actually makes the expert \emph{worse}.
+We tried several alternative imitation learning schedules.
+Best results for the baseline were obtained by collecting an initial dataset of 32 rollouts and thereafter retraining every 16 rollouts.
+Hyperparameters for the baseline were otherwise identical to the ones used for CIM.
+
+Results on the 2D-Hard environment are shown in Fig. \ref{fig:nocur baseline}.
+Like in Fig. \ref{fig:rs baseline},
+we plot the median reward and 5th/95th percentiles over 32 testing episodes.
+
+Even without a curriculum, the algorithm is able to find a policy that \emph{usually} achieves high reward.
+However, CIM reaches the same median performance twice as quickly.
+The story is similar for fifth-percentile performance.
+\nhatch{TODO finish this once rerunning experiments finish}
 

### 2019-02-02 paper/rss/main.tex
Add abstract

@@
 \begin{abstract}
-The abstract goes here.
+  Biped locomotion with precise foot placement has previously been achieved only for static gaits, using detailed dynamics models and hierarchical optimization-based planning.
+  Meanwhile, learning-based approaches to locomotion, such as deep reinforcement learning, are still too sample-inefficient to achieve the precise foot placement necessary to traverse rough terrain.
+  In this paper, we improve sample efficiency for learning-based control techniques.
+  Our strategy applies to any periodic robotic control task that admits significant prior knowledge.
+  First, we engineer a versatile low-level policy (a.k.a. trajectory generator) with interpretable high-level parameters.
+  Second, we use reward shaping, imitation learning, and curriculum learning to train a high-level policy for the control task.
+  We evaluate this technique with a simulated biped locomotion task over rough terrain.
+  To the best of our knowledge, our experiments are the first to achieve such precise foot placement
+  using a fast, non-static gait that requires no optimization during policy execution.
 \end{abstract}
@@
 Results on the 2D-Hard environment are shown in Fig. \ref{fig:nocur baseline}.
-Like in Fig. \ref{fig:rs baseline},
+Like in Fig. \ref{fig:ars baseline},
 we plot the median reward and 5th/95th percentiles over 32 testing episodes.

### 2019-02-02 paper/rss/main.tex
Forgot to commit the figures earlier

@@
 The X axis shows the total number of simulated footsteps taken during training, and the Y axis shows total reward.
-We plot the median and 5th/95th percentiles over 32 testing episodes (eight episodes for each of the four starting random seeds).
+We plot the median and 10th/90th percentiles over 32 testing episodes (eight episodes for each of the four starting random seeds).
 
@@
 Like in Fig. \ref{fig:ars baseline},
-we plot the median reward and 5th/95th percentiles over 32 testing episodes.
+we plot the median reward and 10th/90th percentiles over 32 testing episodes.
 

### 2019-02-02 source/plotter.py

@@ -12,7 +12,7 @@ matplotlib.rcParams.update({'font.size': 14})
 KEYS_TO_PLOT = ['total_reward']
 SETTINGS = {
         "cim_final_easy": ["green", "CIM", LearnInverseDynamics, [5,6,7,8]],
-        "cim_final": ["green", "CIM", LearnInverseDynamics, [1,2,3,4]],
+        "cim_final": ["green", "CIM", LearnInverseDynamics, [17,18,19,20]],
         "rs_final": ["purple", "ARS baseline", RandomSearchBaseline, [13,14,15,16]],
         "nocur_final": ["blue", "No curriculum", LearnInverseDynamics, [9,10,11,12]],
         }
@@ -20,7 +20,7 @@ SETTINGS = {
 def fill_plot(x, data, color):
     line, = plt.plot(x, np.percentile(data, 50, 1), color=color)
     #plt.fill_between(x, mean-std, mean+std, color=color, alpha=0.2)
-    plt.fill_between(x, np.percentile(data, 5, 1), np.percentile(data, 95, 1),
+    plt.fill_between(x, np.percentile(data, 10, 1), np.percentile(data, 90, 1),
             color=color, alpha=0.2)
     return line
 

### 2019-02-02 paper/rss/main.tex
edits to introduction
Fix some TODOs in the paper

@@
 We make several novel modifications to SIMBICON in order to control its footstep placement more precisely:\footnote{
-For more details, see Appendix \ref{app:simbicon}.}
+For more details, see Appendix \ref{app:simbicon} in the supplemental material.}
 \begin{enumerate}
@@
 
-Even without a curriculum, the algorithm is able to find a policy that \emph{usually} achieves high reward.
-However, CIM reaches the same median performance twice as quickly.
-The story is similar for fifth-percentile performance.
-\nhatch{TODO finish this once rerunning experiments finish}
+Even without a curriculum, the algorithm is able to find a policy whose median performance completes all sixteen steps of an episode.
+However, CIM reaches that same median performance three times as quickly.
+Similarly, CIM's 10th-percentile performance is almost perfect with less than 6000 training footsteps, while without a curriculum it takes twice as long to reach that level.
 
@@
 In future work, we plan to investigate how to transfer policies trained with CIM to real-world bipedal robots.
-
-\nhatch{future work: adaptive curriculum (adjusts to learner's current ability)}
+We also plan to investigate whether it helps to have an \emph{adaptive} curriculum that adjusts to the learner's current ability.
 

### 2019-02-02 paper/rss/main.tex
edits

@@
 
-\title{Curriculum and Imitation Learning for Biped Locomotion on Rough Terrain}
+\title{Curriculum and Imitation Learning for Bipedal Locomotion on Rough Terrain}
 
@@
 \begin{abstract}
-  Biped locomotion with precise foot placement has previously been achieved only for static gaits, using detailed dynamics models and hierarchical optimization-based planning.
+  bipedal locomotion with precise foot placement has previously been achieved only for static gaits, using detailed dynamics models and hierarchical optimization-based planning.
   Meanwhile, learning-based approaches to locomotion, such as deep reinforcement learning, are still too sample-inefficient to achieve the precise foot placement necessary to traverse rough terrain.
@@
   First, we engineer a versatile low-level policy (a.k.a. trajectory generator) with interpretable high-level parameters.
-  Second, we use reward shaping, imitation learning, and curriculum learning to train a high-level policy for the control task.
-  We evaluate this technique with a simulated biped locomotion task over rough terrain.
+  Second, we use a combination of reward shaping, imitation learning, and curriculum learning to train a high-level policy for the control task.
+  We evaluate this technique with a simulated bipedal locomotion task over rough terrain.
   To the best of our knowledge, our experiments are the first to achieve such precise foot placement
@@
 Locomotion is a difficult, underactuated control problem.
-\emph{Biped} locomotion is more difficult because the support polygon is small and the center of mass is relatively high.
-Biped locomotion on \emph{rough terrain} is more difficult still, because frequent large obstacles make even small mistakes catastrophic.
+\emph{Bipedal} locomotion is more difficult because the support polygon is small and the center of mass is relatively high.
+bipedal locomotion on \emph{rough terrain} is more difficult still, because frequent large obstacles make even small mistakes catastrophic.
 
@@
 Deep RL is a popular technique to learn highly dynamic motions for humanoids \citep{peng2018deepmimic, heess2017emergence}.
-Using sim-to-real techniques, \citet{tan2018sim} have even applied deep RL techniques to physical robots, albeit for quadruped locomotion rather than biped.
+Using sim-to-real techniques, \citet{tan2018sim} have even applied deep RL techniques to physical robots, albeit for quadrupedal locomotion rather than bipedal.
 The policies learned by deep RL are both computationally and physically fast.
@@
 In this manuscript, we propose several additional ways to simplify robotic learning tasks by taking advantage of prior knowledge.
-Our experiments then demonstrate that these simplifications, taken together, make it feasible to use machine learning for simulated biped locomotion over rough terrain.
+Our experiments then demonstrate that these simplifications, taken together, make it feasible to use machine learning for simulated bipedal locomotion over rough terrain.
 In contrast to prior work in footstep planning, our approach learns a dynamic policy that requires no optimization at execution time.
@@
 \begin{itemize}
-  \item a trajectory generator for controllable biped locomotion, based on SIMBICON \citep{yin2007simbicon} but with several modifications to enhance controllability;
+  \item a trajectory generator for controllable bipedal locomotion, based on SIMBICON \citep{yin2007simbicon} but with several modifications to enhance controllability;
   \item a policy architecture similar to PMTG but with a simpler HLP interface;
   \item an efficient learning algorithm for this architecture, which takes advantage of imitation and curriculum learning; and
-  \item simulated experiments evaluating the necessity and sufficiency of this approach for dynamic biped locomotion over rough terrain.
+  \item simulated experiments evaluating the necessity and sufficiency of this approach for dynamic bipedal locomotion over rough terrain.
 \end{itemize}
@@
 
-Most of the prior work on deep RL for biped locomotion on rough terrain does not explicitly plan individual footstep locations.
+Most of the prior work on deep RL for bipedal locomotion on rough terrain does not explicitly plan individual footstep locations.
 Examples of this are \cite{peng2018deepmimic, heess2017emergence, peng2016terrain}. The resulting policies travel forward quickly, avoid certain obstacles, and are robust to small missteps.
@@
 In subsequent subsections, it will be helpful to have in mind a concrete example of this kind of TG.
-For biped locomotion, our TG is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
+For bipedal locomotion, our TG is a modification of the walking gait from SIMBICON \citep{yin2007simbicon}.
 SIMBICON uses a finite state machine (FSM) with two states: an UP state that ends after 0.3 seconds, and a DOWN state that ends on swing foot contact.
@@
 %The overall thrust of these differences from PMTG is to simplify the high-level environment dynamics as much as possible.
-%This makes CIM suitable for highly challenging environments, such as biped locomotion over rough terrain, where pure RL or random search algorithms are unlikely to succeed.
+%This makes CIM suitable for highly challenging environments, such as bipedal locomotion over rough terrain, where pure RL or random search algorithms are unlikely to succeed.
 
@@
 
-We evaluate CIM on the problem of dynamic biped locomotion over rough terrain.
+We evaluate CIM on the problem of dynamic bipedal locomotion over rough terrain.
 Using the DART simulator \citep{lee2018dart}, we implement two RL problems as illustrated in Fig. \ref{fig:example episodes}.
@@
 The only difference is that its policy architecture still uses the temporal abstraction explained in section \ref{sec:tg}.\footnote{
-As we have not implemented a TG for biped locomotion that can be continuously modulated, this seems like the fairest comparison we could make.}
+As we have not implemented a TG for bipedal locomotion that can be continuously modulated, this seems like the fairest comparison we could make.}
 
@@
 
-In particular, CIM is an intriguing direction for biped locomotion research.
+In particular, CIM is an intriguing direction for bipedal locomotion research.
 To the best of our knowledge, our experiments are the first to achieve such precise foot placement
@@
 
-\section{Detailed description of trajectory generator for biped locomotion}
+\section{Detailed description of trajectory generator for bipedal locomotion}
 

### 2019-02-02 paper/rss/main.tex
More precisely define the reward function

--- 2019-02-02 paper/rss/main.tex
@@
 \usepackage{graphicx}
+\usepackage{amsmath}
 
@@
 \newcommand{\nhatch}[1]{{\leavevmode\color{magenta} Nathan: #1}}
+\newcommand{\norm}[1]{\lVert #1 \rVert}
 
@@
 
-For example, in the case of the locomotion experiments below, our expert is defined by the following optimization problem.
-Given a starting state and a target footstep location, we define the \emph{achieved step location} for an action as the location of the swing heel at the moment that the step ends, which is the moment that any part of the swing foot makes contact with the ground.\footnote{The step also ends if the robot crashes; i.e.
+For example, in the case of the locomotion experiments below, we define the reward function as follows.
+Given a starting state $\mathbf{s}$ and a target footstep location $\mathbf{t}$, we define the final heel location $\mathbf{f}_{heel}$ for an action $\mathbf{a}$ as the location of the swing heel at the moment that the step ends, which is the moment that any part of the swing foot makes contact with the ground.\footnote{The step also ends if the robot crashes; i.e.
 falls off of the edge of a stepping stone, falls over, etc.}
-The objective is the Euclidean distance between the target and achieved step locations,
-plus a hinge loss penalty if the CoM ends up too far \emph{past} the target (projected along the direction of travel).
-\nhatch{TODO: describe this reward more carefully. In particular, it is one minus the penalties described above.}
-At each iteration, we estimate gradients for this objective as a weighted average of $Q$ random directions according to ARS \citep{mania2018simple}.
-We stop iterating once the accuracy reaches two centimeters, or once a certain maximum number of iterations (in our case, five) is reached.
-The hinge loss penalty protects the robot's balance, ensuring that if the expert finds a successful action, it will actually be a \emph{good} action to use to train the HLP.
+We define the final CoM location $\mathbf{f}_{CoM}$ as the location of the CoM at the same moment.
+We define ${p}_{heel}$ and ${p}_{CoM}$ by projecting $\mathbf{f}_{heel}$ and $\mathbf{f}_{CoM}$ respectively along the direction of travel.
+Let $\delta = {p}_{heel} - {p}_{CoM}$, the (signed) distance in meters between these two projected points along the direction of travel.
+Then the reward for this action is
+\[
+r(\mathbf{s},\mathbf{a}) := 1 - \norm{\mathbf{f}_{heel} - \mathbf{t}}_2 - \max\{|\delta-0.1|-0.1, 0\}
+\]
+The final term is the \emph{balance penalty}, ensuring that for high-reward actions, the CoM must be approximately ten centimeters behind the heel-strike location.
+
+Given a state $\mathbf{s}$, the expert attempts to find a good action $\mathbf{a}$ using random search, specifically ARS \citep{mania2018simple}.
+At each iteration, we estimate gradients for the objective $r(\mathbf s, \mathbf a)$ as a weighted average of $Q$ random directions, then take a step in that direction.
+We stop iterating once the accuracy reaches two centimeters ($r \geq 0.98$), or once a certain maximum number of iterations (in our case, five) is reached.
 

### 2019-02-02 paper/rss/main.tex
edits to introduction

--- 2019-02-02 paper/rss/main.tex
@@
 They are physically slow because accurate modeling requires relatively static walking, and also because the optimization procedures are computationally slow.
-For the DRC, reliability was more important than speed, but eventually we want to have robots that can traverse rough terrain using faster, dynamic gaits.
+For the DRC, reliability was more important than speed, but eventually the goal is to build robots that can traverse rough terrain using faster, dynamic gaits.
 
@@
 Deep RL is a popular technique to learn highly dynamic motions for humanoids \citep{peng2018deepmimic, heess2017emergence}.
-Using sim-to-real techniques, \citet{tan2018sim} have even applied deep RL techniques to physical robots, albeit for quadrupedal locomotion rather than bipedal.
+Using sim-to-real techniques, \citet{tan2018sim} have applied deep RL techniques to physical robots, albeit for quadrupedal locomotion rather than bipedal locomotion.
 The policies learned by deep RL are both computationally and physically fast.

### 2019-02-02 paper/rss/main.tex
edits to section II

@@
 These approaches tend to generate \emph{hierarchical} plans in order to reduce computational requirements.
-First, they choose footstep locations using search, like A* \citep{huang2013step}.
-Based on the footstep plan, they generate a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum \citep{mordatch2010robust} or contact wrench cone \citep{dai2016planning}.
-Then they use inverse kinematics to plan trajectories for end-effectors \citep{zucker2010optimization}.
-Finally, these trajectories are optimized using the full dynamics model to account for obstacles and other details \citep{ratliff2009chomp}.
+At the highest level, the planner chooses footstep locations using, e.g. A* search~\citep{huang2013step}.
+Based on the footstep plan, a lower-level planner generates a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum~\citep{mordatch2010robust} or contact wrench cone~\citep{dai2016planning}.
+Next, inverse kinematics are used to plan trajectories for end-effectors \citep{zucker2010optimization}.
+Finally, these trajectories are optimized using the full dynamics model to account for obstacles and other details~\citep{ratliff2009chomp}.
 

### 2019-02-02 paper/rss/main.tex
edits to section II

@@
 
-These approaches tend to be reliable but slow; e.g. four seconds per step in \citet{feng2015optimization}.
+Hierarchical planning approaches tend to be reliable but slow; e.g. four seconds per step in \citet{feng2015optimization}.
 This is because static walking is easier to model and optimization takes a lot of computation time.
 They also require significant domain knowledge and manual tuning in order to engineer good models and handle situations when the real world deviates from ideal physics.
-For now, they are the best known approaches to robotic locomotion, given the supreme importance of safety and reliability for large physical systems and high-stakes competitions such as the DARPA Robotics and Learning Locomotion Challenges.
+Hierarchical planners are currently the best known approaches to robotic locomotion, given the supreme importance of safety and reliability for large physical systems and high-stakes competitions such as the DARPA Robotics and Learning Locomotion Challenges.
 However, we are interested in exploring alternative possibilities that can exploit the dynamics of the system to achieve faster and more energy-efficient gaits.
@@
 There are a few prior works that use learning-based approaches to do step planning for simulated systems.
-\cite{peng2017deeploco} use a hierarchical reinforcement learning algorithm where the upper-level controller chooses footstep locations and the lower-level controller tries to execute those plans.
-However, it produces target step locations that are unrealistic and rarely actually achieved.
+\citet{peng2017deeploco} use a hierarchical reinforcement learning algorithm where the upper-level controller chooses footstep locations and the lower-level controller tries to execute those plans.
+However, this strategy produces target step locations that are unrealistic and rarely actually achieved.
 The path that the agent follows is perfectly flat; on rough terrain, such inaccurate step planning would probably fail.
-\cite{karpathy2012curriculum} use curriculum learning for a simple planar biped to collect a set of (state, action, resulting step distance) tuples.
+\citet{karpathy2012curriculum} use curriculum learning for a simple planar biped to collect a set of (state, action, resulting step distance) tuples.
 The final policy is based on nearest-neighbor search among this set of collected experience.

### 2019-02-02 paper/rss/main.tex
edits to section II

@@
 \usepackage{amsmath}
+\newcommand{\boots}[1]{{\textcolor{red} {Byron: {#1}}}}
 
@@
 
-Most of the prior work on deep RL for bipedal locomotion on rough terrain does not explicitly plan individual footstep locations.
-Examples of this are \cite{peng2018deepmimic, heess2017emergence, peng2016terrain}. The resulting policies travel forward quickly, avoid certain obstacles, and are robust to small missteps.
+Most prior work on deep RL for bipedal locomotion on rough terrain does not explicitly plan individual footstep locations.
+Examples of this \boots{examples of what?} are \cite{peng2018deepmimic, heess2017emergence, peng2016terrain}.
+The resulting policies travel forward quickly, avoid certain obstacles, and are robust to small missteps.
 However, they have not been shown to work in situations with dense or large obstacles where small missteps can be catastrophic.
@@
 Perhaps the most closely related prior work is \citet{iscen2018pmtg}.
-As described in section \ref{sec:intro}, PMTGs use a trajectory generator to simplify the environment when learning challenging tasks, such as controllable robot locomotion.
-This allows the HLP to use a very simple hypothesis class (e.g. linear) and a very simple learning algorithm (e.g. ARS \citep{mania2018simple}).
+As described in Section~\ref{sec:intro}, PMTGs use a trajectory generator to simplify the environment when learning challenging tasks, such as controllable robot locomotion.
+This allows the HLP to use a very simple hypothesis class (e.g. linear) and a very simple learning algorithm (e.g. \boots{what is ARS?} ARS \citep{mania2018simple}).
 However, PMTG has been applied only to control walking speed for a quadruped on flat ground.
-The policy architecture in our paper is similar to PMTG in that we learn a high-level policy that modulates a TG.
-However, one important difference is that in our paper, the HLP and the TG run at different frequencies.
+The policy architecture in our work is similar to PMTG in that we learn a high-level policy that modulates a TG.
+However, one important difference is that in our work, the HLP and the TG run at different frequencies. \boots{which means what?}
 We will give a detailed description of these differences in the next section.

### 2019-02-02 paper/rss/main.tex
Small edits to appendix

--- 2019-02-02 paper/rss/main.tex
@@
 \begin{enumerate}
-  \item Set the target swing leg joint angles in the DOWN state using inverse kinematics (IK).
-    Again, adjust the target swing heel location for the purposes of IK from the true target location based on the CoM speed and the same gain parameter as above.
 
@@
     This does not have much effect on the step distance, but it helps make the dynamics of each impact more predictable, which is important for accurate function approximation.
-    \nhatch{Again, all of this will have to be validated with ablations\dots.}
 
@@
 
-Note that there are also many additional parameters of SIMBICON that we could have exposed to the high-level policy; e.g. the stance ankle angle of the UP state.
+Note that there are also many additional parameters of SIMBICON that we could have exposed to the high-level policy; e.g.
+the stance ankle angle of the UP state, or the balance gain parameters used by SIMBICON.
 Of course, we prefer to keep the problem dimension as low as possible, and we found that the short list of parameters given in Table \ref{table:params} was enough for our experiments.

### 2019-02-02 paper/rss/main.tex
Address comments

@@
 
-Most prior work on deep RL for bipedal locomotion on rough terrain does not explicitly plan individual footstep locations.
-Examples of this \boots{examples of what?} are \cite{peng2018deepmimic, heess2017emergence, peng2016terrain}.
+Most prior work on deep RL for bipedal locomotion on rough terrain does not explicitly plan individual footstep locations \citep{peng2018deepmimic, heess2017emergence, peng2016terrain}.
 The resulting policies travel forward quickly, avoid certain obstacles, and are robust to small missteps.
@@
 As described in Section~\ref{sec:intro}, PMTGs use a trajectory generator to simplify the environment when learning challenging tasks, such as controllable robot locomotion.
-This allows the HLP to use a very simple hypothesis class (e.g. linear) and a very simple learning algorithm (e.g. \boots{what is ARS?} ARS \citep{mania2018simple}).
+This allows the HLP to use a very simple hypothesis class (e.g. linear) and a very simple learning algorithm (e.g. Augmented Random Search \citep{mania2018simple}).
 However, PMTG has been applied only to control walking speed for a quadruped on flat ground.
 The policy architecture in our work is similar to PMTG in that we learn a high-level policy that modulates a TG.
-However, one important difference is that in our work, the HLP and the TG run at different frequencies. \boots{which means what?}
-We will give a detailed description of these differences in the next section.
+However, one important difference is that in our work, the HLP and the TG run at different frequencies.
+In the next section, we give a detailed description of this and other differences.
 
@@
 
-Given a state $\mathbf{s}$, the expert attempts to find a good action $\mathbf{a}$ using random search, specifically ARS \citep{mania2018simple}.
+Given a state $\mathbf{s}$, the expert attempts to find a good action $\mathbf{a}$ using random search, specifically Augmented Random Search (ARS) \citep{mania2018simple}.
 At each iteration, we estimate gradients for the objective $r(\mathbf s, \mathbf a)$ as a weighted average of $Q$ random directions, then take a step in that direction.

### 2019-02-02 paper/rss/main.tex
Add doppelganger for 2D environment
This is mostly to make it easier to hack together a diagram of the FSM.

Also tweak training settings for 3D; trying to run official 3D experiments
now.

@@
 To the best of our knowledge, our approach is the first to achieve such precise foot placement
-using a fast, dynamic gait that requires no optimization during policy execution.
+using a fast, non-static gait that requires no optimization during policy execution.
 
@@
 To the best of our knowledge, our experiments are the first to achieve such precise foot placement
-using a fast, dynamic gait that requires no optimization during policy execution.
+using a fast, non-static gait that requires no optimization during policy execution.
 However, one significant limitation of our work so far is that all experiments were performed in simulation.

### 2019-02-02 source/consts_2D.py

@@ -2,6 +2,8 @@ import numpy as np
 import pydart2.utils.transformations as libtransform
 from utils import build_mask
 
+skel_file = 'skel/walker2d.skel'
+
 perm = [0,1,2,3,4,5,6,7,8]
 sign_switches = []
 

### 2019-02-02 source/curriculum.py

@@ -33,7 +33,6 @@ TRAIN_SETTINGS_2D = {
     'model_class': 'linear',
     }
 
-# TODO do this for 3D as well
 cp_orig = TRAIN_SETTINGS_2D['controllable_params']
 col = cp_orig.reshape((-1,1))
 row = np.concatenate((o2d, [True])).reshape((1,-1))
@@ -91,7 +90,7 @@ o3d = np.concatenate([oq_3d, oq_3d,
                       ot_3d, ot_3d, ot_3d, ot_3d])
 
 TRAIN_SETTINGS_3D = {
-    'n_trajectories': 3,
+    'n_trajectories': 4,
     'n_dirs': 8,
     'tol': 0.05,
     'max_intolerable_steps': 3,
@@ -107,8 +106,8 @@ TRAIN_SETTINGS_3D = {
     'model_class': 'linear',
     }
 
-TRAIN_SETTINGS_3D_PRECISE = {**TRAIN_SETTINGS_3D,
-    'tol': 0.02,
+TRAIN_SETTINGS_3D_PLUS = {**TRAIN_SETTINGS_3D,
+    'model_class': 'quadratic',
     }
 
 SETTINGS_3D_EASY = {
@@ -129,13 +128,13 @@ SETTINGS_3D_MEDIUM = {**SETTINGS_3D_EASY,
     }
 
 SETTINGS_3D_HARD = {**SETTINGS_3D_MEDIUM,
-    'dist_spread': 0.5,
-    'z_spread': 0.2,
-    }
-
-SETTINGS_3D_HARDER = {**SETTINGS_3D_HARD,
     'use_stepping_stones': True,
     'ground_length': 0.3,
     'ground_width': 0.2,
+    }
+
+SETTINGS_3D_HARDER = {**SETTINGS_3D_HARD,
+    'dist_spread': 0.5,
+    'z_spread': 0.2,
     }
 

### 2019-02-02 source/experiment.py

@@ -82,14 +82,11 @@ def ex_3D(uq_id):
     from simple_3D_env import Simple3DEnv
     from simbicon_3D import Simbicon3D
     env = Simple3DEnv(Simbicon3D)
-    ex = Experiment(env, "new_experiment"+uq_id, ['SETTINGS_3D_HARDER'])
-    ex.run_iters(2, cur.SETTINGS_3D_EASY, cur.TRAIN_SETTINGS_3D)
-    ex.run_iters(6, cur.SETTINGS_3D_MEDIUM, cur.TRAIN_SETTINGS_3D)
-    ex.run_iters(6, cur.SETTINGS_3D_MEDIUM, cur.TRAIN_SETTINGS_3D_PRECISE)
+    ex = Experiment(env, LearnInverseDynamics, "3D_test_"+uq_id, ['SETTINGS_3D_MEDIUM', 'SETTINGS_3D_HARD'])
+    ex.run_iters(5, cur.SETTINGS_3D_EASY, cur.TRAIN_SETTINGS_3D)
+    ex.run_iters(15, cur.SETTINGS_3D_MEDIUM, cur.TRAIN_SETTINGS_3D)
     ex.run_iters(6, cur.SETTINGS_3D_HARD, cur.TRAIN_SETTINGS_3D)
-    ex.run_iters(6, cur.SETTINGS_3D_HARD, cur.TRAIN_SETTINGS_3D_PRECISE)
-    ex.run_iters(6, cur.SETTINGS_3D_HARDER, cur.TRAIN_SETTINGS_3D_PRECISE)
-    embed()
+    ex.run_iters(12, cur.SETTINGS_3D_HARD, cur.TRAIN_SETTINGS_3D_PLUS)
 
 def ex_2D_cim(uq_id):
     from stepping_stones_env import SteppingStonesEnv
@@ -98,14 +95,12 @@ def ex_2D_cim(uq_id):
     ex.run_iters(3, cur.SETTINGS_2D_EASY, cur.TRAIN_SETTINGS_2D)
     ex.run_iters(18, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D)
     ex.run_iters(12, cur.SETTINGS_2D_HARD, cur.TRAIN_SETTINGS_2D_PLUS)
-    embed()
 
 def ex_2D_cim_easy(uq_id):
     from stepping_stones_env import SteppingStonesEnv
     env = SteppingStonesEnv()
[…]

### 2019-02-02 source/plotter.py

@@ -9,12 +9,12 @@ import seaborn as sns
 import matplotlib
 matplotlib.rcParams.update({'font.size': 14})
 
-KEYS_TO_PLOT = ['total_reward']
 SETTINGS = {
-        "cim_final_easy": ["green", "CIM", LearnInverseDynamics, [5,6,7,8]],
-        "cim_final": ["green", "CIM", LearnInverseDynamics, [17,18,19,20]],
-        "rs_final": ["purple", "ARS baseline", RandomSearchBaseline, [13,14,15,16]],
-        "nocur_final": ["blue", "No curriculum", LearnInverseDynamics, [9,10,11,12]],
+        "cim_final_easy": ["green", "CIM", LearnInverseDynamics, [5,6,7,8], 'SETTINGS_2D_EASY'],
+        "cim_final": ["green", "CIM", LearnInverseDynamics, [17,18,19,20], 'SETTINGS_2D_HARD'],
+        "cim_3D": ["green", "CIM", LearnInverseDynamics, [1], 'SETTINGS_3D_MEDIUM'],
+        "rs_final": ["purple", "ARS baseline", RandomSearchBaseline, [13,14,15,16], 'SETTINGS_2D_EASY'],
+        "nocur_final": ["blue", "No curriculum", LearnInverseDynamics, [9,10,11,12], 'SETTINGS_2D_HARD'],
         }
 
 def fill_plot(x, data, color):
@@ -33,9 +33,9 @@ def steps_from_history(history):
 
 def load_exp(exp_name, seed):
     name = "{}_{}".format(exp_name, seed)
-    settings_name = 'SETTINGS_2D_EASY' if exp_name in ['cim_final_easy', 'rs_final'] else 'SETTINGS_2D_HARD'
-    ex = Experiment(None, SETTINGS[exp_name][2], name, [settings_name])
-    return steps_from_history(ex.learn.history), ex.results[settings_name]['total_reward']
+    eval_settings = SETTINGS[exp_name][4]
+    ex = Experiment(None, SETTINGS[exp_name][2], name, [eval_settings])
+    return steps_from_history(ex.learn.history), ex.results[eval_settings]['total_reward']
 
 def multiseed_plot(exp_name):
     xx = []
@@ -77,5 +77,10 @@ def gen_figures():
     plt.legend(lines, labels, loc='lower right')
     save_plot('../paper/figures/nocur_baseline.pdf')
 
+def test():
[…]

### 2019-02-02 source/simbicon.py

@@ -275,6 +275,8 @@ class Simbicon(PDController):
         q, dq = self.env.get_x()
         params = self.params
         target_q = self.compute_target_q(q, dq)
+        self.update_doppelganger(target_q)
+
         # We briefly increase Kd (mechanical impedance?) for the stance knee
         # in order to prevent the robot from stepping so hard that it bounces.
         fix_Kd = self.direction == UP and self.time() - self.step_started < 0.1
@@ -295,18 +297,32 @@ class Simbicon(PDController):
         torques = self.env.from_features(control)
         return torques
 
-def test(env, length, seed=None, runway_length=15, runway_x=0):
+    def update_doppelganger(self, tq):
+        dop = self.env.doppelganger
+        if dop is None:
+            return
+        c = self.env.consts()
+        tq = tq.copy()
+        dop.q = self.env.from_features(tq)
+        ik = InverseKinematics(self.env.doppelganger, self.env)
+        offset = c.THIGH_BODYNODE_OFFSET
+        dop_bodynode = ik.get_bodynode(self.stance_idx, offset)
+        robot_bodynode = self.ik.get_bodynode(self.stance_idx, offset)
+        tq[:c.BRICK_DOF] = ik.get_dofs(robot_bodynode.transform(), dop_bodynode)
+        self.env.doppelganger.q = self.env.from_features(tq)
+        self.env.doppelganger.dq = np.zeros(c.Q_DIM)
+
+def test(env, length, seed=None, runway_length=15, runway_x=0, render=1):
     env.clear_skeletons() # Necessary in order to change the runway length
     env.sdf_loader.ground_length = runway_length
-    start_state = env.reset(seed=seed, random=0.005, render=1)
+    start_state = env.reset(seed=seed, random=0.005, render=render)
     env.sdf_loader.put_grounds([[runway_x,0,0]])
+    action = np.zeros(sp.N_PARAMS)
     for i in range(8):
-        # TODO: for very small target steps (e.g. 10 cm), the velocity is so small that
-        # the robot can get stuck in the UP state, balancing on one leg.
-        # TODO: for long steps, (e.g. 80 cm) the robot hits the target with its toe rather
-        # than its heel. This makes difficult training environments for random optimization.
+        if i > 2:
+            action[sp.UP_IDX+sp.SWING_ANKLE_RELATIVE] = -0.3
[…]

### 2019-02-02 source/simbicon_3D.py

@@ -16,7 +16,7 @@ class Simbicon3D(Simbicon):
     def base_gait(self):
         gait = [0.14, 0.5, 0.2, -0.1, 0.2,
                 0.4, -1.1,   0.0, -0.05,
-                -0.0, -0.00, 0.05, -0.1,
+                -0.0, -0.00, 0.1, -0.1,
                 0.5, 0.2, 0.0, 0.0, 0.0, 0.0]
         return np.array(gait)
 
@@ -62,23 +62,7 @@ class Simbicon3D(Simbicon):
 
         tq[TORSO_ROLL] = -q[ROOT_ROLL]
 
-        self.update_doppelganger(tq)
         return tq
-
-    def update_doppelganger(self, tq):
-        dop = self.env.doppelganger
-        if dop is None:
-            return
-        c = self.env.consts()
-        tq = tq.copy()
-        dop.q = self.env.from_features(tq)
-        ik = InverseKinematics(self.env.doppelganger, self.env)
-        offset = c.THIGH_BODYNODE_OFFSET
-        dop_bodynode = ik.get_bodynode(self.stance_idx, offset)
-        robot_bodynode = self.ik.get_bodynode(self.stance_idx, offset)
-        tq[:6] = ik.get_dofs(robot_bodynode.transform(), dop_bodynode)
-        self.env.doppelganger.q = self.env.from_features(tq)
-        self.env.doppelganger.dq = np.zeros(c.Q_DIM)
 
 def next_target(start, heading, length, env):
     c = env.controller

### 2019-02-02 source/simple_3D_env.py

@@ -8,7 +8,6 @@ from OpenGL.GL import GLfloat
 from stepping_stones_env import SteppingStonesEnv
 import consts_armless as consts
 
-SIMULATION_RATE = 1.0 / 2000.0
 THETA = -np.pi/12
 PHI = np.pi / 1.5
 ZOOM = 5.0
@@ -66,20 +65,6 @@ class Simple3DEnv(SteppingStonesEnv):
         if self.world.frame % framerate == 0:
             #print(self.world.time())
             self._render()
-
-    def load_world(self):
-        skel_file = consts.skel_file
-        world = pydart.World(SIMULATION_RATE, skel_file)
-        skel = world.skeletons[1]
-        self.doppelganger = None
-        if len(world.skeletons) == 3:
-            self.doppelganger = world.skeletons[2]
-            assert(self.doppelganger.name == "doppelganger")
-        skel.set_self_collision_check(True)
-        # The kima_human_box model actually has preset damping coeffs.
-        #for dof in skel.dofs[6:]:
-        #    dof.set_damping_coefficient(0.2)
-        return world
 
     def consts(self):
         return consts

### 2019-02-02 source/stepping_stones_env.py

@@ -33,6 +33,8 @@ class SteppingStonesEnv:
         self.viewer = None
         self.sdf_loader = SDFLoader(self.consts().DEFAULT_GROUND_WIDTH)
         pydart.init(verbose=False)
+        c = self.consts()
+        self.is_3D = (c.BRICK_DOF == 6)
         self.clear_skeletons()
 
         self.video_recorder = None
@@ -40,9 +42,6 @@ class SteppingStonesEnv:
         self.metadata = {'render.modes': ['rgb_array', 'human']}
         #self.reward_range = range(10)
         #self.spec = None
-
-        c = self.consts()
-        self.is_3D = (c.BRICK_DOF == 6)
 
     def consts(self):
         return consts_2D
@@ -236,13 +235,22 @@ class SteppingStonesEnv:
 
     def update_viewer(self, com):
         track_point = self.track_point or com
+        track_point = track_point + np.array([-2, 0, 0])
+        #tb.trans[1] = -0.5
         self.viewer.scene.tb.trans[0] = -track_point[0]
 
     def gui(self):
         pydart.gui.viewer.launch(self.world)
 
     def load_world(self):
-        skel = "skel/walker2d.skel"
-        world = pydart.World(SIMULATION_RATE, skel)
+        skel_file = self.consts().skel_file
+        world = pydart.World(SIMULATION_RATE, skel_file)
+        self.doppelganger = None
+        if len(world.skeletons) == 3:
+            self.doppelganger = world.skeletons[2]
+            assert(self.doppelganger.name == "doppelganger")
+        if self.is_3D:
+            skel = world.skeletons[1]
+            skel.set_self_collision_check(True)
         return world
 

### 2019-02-02 source/evaluator.py
Make it easier to take example screenshots of environments
Fiddle with the Experiment.visualize() method until it looks right.

Also remove the old code used to get Simbicon doppelganger screenshots.

@@ -1,5 +1,6 @@
 import numpy as np
 import utils
+from sdf_loader import GREEN
 
 class Evaluator:
     def __init__(self, env):
@@ -11,7 +12,7 @@ class Evaluator:
             self.env.sdf_loader.ground_width = self.eval_settings['ground_width']
             self.env.sdf_loader.ground_length = self.eval_settings['ground_length']
         else:
-            self.env.sdf_loader.ground_length = 10.0
+            self.env.sdf_loader.ground_length = 7.5
             self.env.sdf_loader.ground_width = self.env.consts().DEFAULT_GROUND_WIDTH
         self.env.clear_skeletons()
 
@@ -27,6 +28,9 @@ class Evaluator:
             self.env.sdf_loader.put_grounds(targets)
         else:
             self.env.sdf_loader.put_grounds(targets[:1])
+            if render is not None:
+                for i,t in enumerate(targets[2:]):
+                    self.env.sdf_loader.put_dot(t, str(i), color=GREEN)
         total_offset = 0
         num_successful_steps = 0
         num_intolerable_steps = 0

### 2019-02-02 source/experiment.py

@@ -78,11 +78,23 @@ class Experiment:
             self.learn.training_iter()
             self.run_evaluations()
 
+    def visualize(self):
+        import stepping_stones_env
+        # Ensure we take less than one step
+        stepping_stones_env.EPISODE_TIME_LIMIT = 0.3
+        self.learn.evaluator.set_eval_settings(cur.SETTINGS_3D_HARD)
+        self.learn.env.track_point = [3.0, 0.0, 0.0] # Use in 3D
+        #self.learn.env.track_point = [3.7, 0.0, 0.0] # Use in 2D
+        #self.learn.env.zoom = 6.0 # Use in 2D
+        self.learn.env.theta = -np.pi/4
+        self.learn.env.phi = 0.0
+        self.learn.evaluate(render=1)
+
 def ex_3D(uq_id):
     from simple_3D_env import Simple3DEnv
     from simbicon_3D import Simbicon3D
     env = Simple3DEnv(Simbicon3D)
-    ex = Experiment(env, LearnInverseDynamics, "3D_test_"+uq_id, ['SETTINGS_3D_MEDIUM', 'SETTINGS_3D_HARD'])
+    ex = Experiment(env, LearnInverseDynamics, "cim_3D_"+uq_id, ['SETTINGS_3D_MEDIUM', 'SETTINGS_3D_HARD'])
     ex.run_iters(5, cur.SETTINGS_3D_EASY, cur.TRAIN_SETTINGS_3D)
     ex.run_iters(15, cur.SETTINGS_3D_MEDIUM, cur.TRAIN_SETTINGS_3D)
     ex.run_iters(6, cur.SETTINGS_3D_HARD, cur.TRAIN_SETTINGS_3D)

### 2019-02-02 source/simbicon.py

@@ -312,15 +312,16 @@ class Simbicon(PDController):
         self.env.doppelganger.q = self.env.from_features(tq)
         self.env.doppelganger.dq = np.zeros(c.Q_DIM)
 
-def test(env, length, seed=None, runway_length=15, runway_x=0, render=1):
+def test(env, length, n=8, seed=None, runway_length=15, runway_x=0, render=1, video_save_dir=None):
     env.clear_skeletons() # Necessary in order to change the runway length
     env.sdf_loader.ground_length = runway_length
-    start_state = env.reset(seed=seed, random=0.005, render=render)
+    start_state = env.reset(seed=seed, random=0.005,
+            render=render, video_save_dir=video_save_dir)
     env.sdf_loader.put_grounds([[runway_x,0,0]])
     action = np.zeros(sp.N_PARAMS)
-    for i in range(8):
-        if i > 2:
-            action[sp.UP_IDX+sp.SWING_ANKLE_RELATIVE] = -0.3
+    for i in range(n):
+        #if i > 2:
+        #    action[sp.UP_IDX+sp.SWING_ANKLE_RELATIVE] = -0.3
         t = length*(0.5 + i)# + np.random.uniform(low=-0.2, high=0.2)
         _, terminated = env.simulate([t,0,0], action=action)
         if terminated:
@@ -344,6 +345,7 @@ def reproduce_bug(env):
 if __name__ == '__main__':
     from stepping_stones_env import SteppingStonesEnv
     env = SteppingStonesEnv()
+    #test(env, 0.5, n=4, render=2, video_save_dir='monitoring')
     test(env, 0.5)
     #reproduce_bug(env)
     embed()

### 2019-02-02 source/stepping_stones_env.py

@@ -84,8 +84,11 @@ class SteppingStonesEnv:
 
     def init_camera(self):
         tb = Trackball(theta=-45.0, phi = 0.0)
-        tb.trans[2] = -5
         self.track_point = None
+        self.zoom = 5.0
+        def zoom_to(dx, dy):
+            self.zoom -= dy/20
+        tb.zoom_to = zoom_to
         return tb
 
     def reset(self, state=None, video_save_dir=None, render=None, random=0.0, seed=None):
@@ -137,7 +140,7 @@ class SteppingStonesEnv:
             return None, False, None
 
     # Run one footstep of simulation, returning the final state
-    def simulate(self, target, target_heading=None, action=None, put_dots=True):
+    def simulate(self, target, target_heading=None, action=None, put_dots=False):
         self.controller.set_gait_raw(raw_gait=action, target_heading=target_heading, target=target)
         steps_per_render = None
         if self.render_rate:
@@ -235,9 +238,10 @@ class SteppingStonesEnv:
 
     def update_viewer(self, com):
         track_point = self.track_point or com
-        track_point = track_point + np.array([-2, 0, 0])
+        #track_point = track_point + np.array([-2, 0, 0])
         #tb.trans[1] = -0.5
         self.viewer.scene.tb.trans[0] = -track_point[0]
+        self.viewer.scene.tb.trans[2] = -self.zoom
 
     def gui(self):
         pydart.gui.viewer.launch(self.world)

### 2019-02-02 paper/rss/main.tex
Add more figures to paper: curriculum and FSM visualizations

@@
 
+\begin{figure}
+  \includegraphics[width=0.15\textwidth]{../figures/TOE_OFF.png}
+  \includegraphics[width=0.12\textwidth]{../figures/UP.png}
+  \includegraphics[width=0.175\textwidth]{../figures/DOWN.png}
+  \caption{\label{fig:fsm} Visualization of target joint angles for the locomotion TG.
+  State TOE-OFF (left) transitions to state UP (middle) after 0.2s, which transitions to state DOWN once the robot is close to the target location.
+  Finally, state DOWN transitions to state UP upon heel-strike.}
+\end{figure}
+
 In subsequent subsections, it will be helpful to have in mind a concrete example of this kind of TG.
@@
 
-See Table \ref{table:params} for a summary of the parameters of the TG.
+See Fig. \ref{fig:fsm} for a visualization of the TG's states, and
+see Table \ref{table:params} for a summary of the parameters of the TG.
 These parameters are \emph{residual} in the sense that setting all of them to zero will produce the default behavior of the TG.
@@
 This is similar to other slow-expert/fast-learner algorithms such as ExIt \citep{anthony2017thinking} and AlphaGo Zero \citep{silver2017mastering}.
+
+\begin{figure}
+  \includegraphics[width=0.48\textwidth]{../figures/2D_EASY.png}
+  \includegraphics[width=0.48\textwidth]{../figures/2D_HARD.png}
+  \includegraphics[width=0.48\textwidth]{../figures/3D_EASY.png}
+  \includegraphics[width=0.48\textwidth]{../figures/3D_MEDIUM.png}
+  \includegraphics[width=0.48\textwidth]{../figures/3D_HARD.png}
+  \caption{\label{fig:curriculum} Example episodes from the various environments used in the locomotion curriculum.
+  Top to bottom: 2D-Easy, 2D-Hard, 3D-Easy, 3D-Medium, and 3D-Hard.}
+\end{figure}
 

### 2019-02-08 paper/rss/main.tex
Small typo fix

@@
 \begin{abstract}
-  bipedal locomotion with precise foot placement has previously been achieved only for static gaits, using detailed dynamics models and hierarchical optimization-based planning.
+  Bipedal locomotion with precise foot placement has previously been achieved only for static gaits, using detailed dynamics models and hierarchical optimization-based planning.
   Meanwhile, learning-based approaches to locomotion, such as deep reinforcement learning, are still too sample-inefficient to achieve the precise foot placement necessary to traverse rough terrain.

### 2019-02-08 source/plotter.py

@@ -12,7 +12,8 @@ matplotlib.rcParams.update({'font.size': 14})
 SETTINGS = {
         "cim_final_easy": ["green", "CIM", LearnInverseDynamics, [5,6,7,8], 'SETTINGS_2D_EASY'],
         "cim_final": ["green", "CIM", LearnInverseDynamics, [17,18,19,20], 'SETTINGS_2D_HARD'],
-        "cim_3D": ["green", "CIM", LearnInverseDynamics, [1], 'SETTINGS_3D_MEDIUM'],
+        "cim_3D": ["green", "CIM", LearnInverseDynamics, [1], 'SETTINGS_3D_HARD'],
+        "3D_test": ["green", "CIM", LearnInverseDynamics, [1], 'SETTINGS_3D_HARD'],
         "rs_final": ["purple", "ARS baseline", RandomSearchBaseline, [13,14,15,16], 'SETTINGS_2D_EASY'],
         "nocur_final": ["blue", "No curriculum", LearnInverseDynamics, [9,10,11,12], 'SETTINGS_2D_HARD'],
         }
@@ -78,8 +79,8 @@ def gen_figures():
     save_plot('../paper/figures/nocur_baseline.pdf')
 
 def test():
-    multiseed_plot('cim_3D')
-    save_plot('test.png')
+    multiseed_plot('3D_test')
+    save_plot('hard.png')
 
 if __name__ == '__main__':
     #gen_figures()

### 2019-02-08 source/plotter.py
Plot results for individual seeds for debugging/inspection

@@ -38,10 +38,12 @@ def load_exp(exp_name, seed):
     ex = Experiment(None, SETTINGS[exp_name][2], name, [eval_settings])
     return steps_from_history(ex.learn.history), ex.results[eval_settings]['total_reward']
 
-def multiseed_plot(exp_name):
+def multiseed_plot(exp_name, seeds=None):
     xx = []
     yy = []
-    for seed in SETTINGS[exp_name][3]:
+    if seeds is None:
+        seeds = SETTINGS[exp_name][3]
+    for seed in seeds:
         x,y = load_exp(exp_name, seed)
         xx.append(x)
         yy.append(y)
@@ -78,10 +80,18 @@ def gen_figures():
     plt.legend(lines, labels, loc='lower right')
     save_plot('../paper/figures/nocur_baseline.pdf')
 
-def test():
-    multiseed_plot('3D_test')
-    save_plot('hard.png')
+def plot_single_seed(exp_name, seed):
+    multiseed_plot(exp_name, [seed])
+    eval_settings = SETTINGS[exp_name][4]
+    save_plot('data/{}_{}/{}.pdf'.format(exp_name, seed, eval_settings))
+
+def all_single_seeds():
+    for exp_name in SETTINGS.keys():
+        seeds = SETTINGS[exp_name][3]
+        for seed in seeds:
+            plot_single_seed(exp_name, seed)
 
 if __name__ == '__main__':
     #gen_figures()
-    test()
+    #plot_single_seed('3D_test', 1)
+    all_single_seeds()

### 2019-03-10 paper/rss/main.tex
Small paper edits

@@
 These approaches tend to generate \emph{hierarchical} plans in order to reduce computational requirements.
-At the highest level, the planner chooses footstep locations using, e.g. A* search~\citep{huang2013step}.
+At the highest level, the planner chooses footstep locations using e.g. A* search~\citep{huang2013step}.
 Based on the footstep plan, a lower-level planner generates a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum~\citep{mordatch2010robust} or contact wrench cone~\citep{dai2016planning}.
@@
 
-Hierarchical planning approaches tend to be reliable but slow; e.g. four seconds per step in \citet{feng2015optimization}.
+Hierarchical optimization approaches tend to be reliable but slow; e.g. four seconds per step in \citet{feng2015optimization}.
 This is because static walking is easier to model and optimization takes a lot of computation time.
-They also require significant domain knowledge and manual tuning in order to engineer good models and handle situations when the real world deviates from ideal physics.
+These approaches also require significant domain knowledge and manual tuning in order to engineer good models and handle situations when the real world deviates from ideal physics.
 Hierarchical planners are currently the best known approaches to robotic locomotion, given the supreme importance of safety and reliability for large physical systems and high-stakes competitions such as the DARPA Robotics and Learning Locomotion Challenges.
@@
 However, one important difference is that in our work, the HLP and the TG run at different frequencies.
-In the next section, we give a detailed description of this and other differences.
+In the next section, we give a detailed description of the differences.
 
@@
 with weights determined by the change in total shaped reward as defined in section \ref{sec:expert}.
-Note that this baseline is very similar to PMTG \citep{iscen2018pmtg}.
+Note that this baseline is very similar to PMTG (specifically, the experiments with linear HLPs in \citet{iscen2018pmtg}).
 The only difference is that its policy architecture still uses the temporal abstraction explained in section \ref{sec:tg}.\footnote{

### 2019-05-30 paper/rss/main.tex
Reread paper, mark some TODOs

@@
   Meanwhile, learning-based approaches to locomotion, such as deep reinforcement learning, are still too sample-inefficient to achieve the precise foot placement necessary to traverse rough terrain.
-  In this paper, we improve sample efficiency for learning-based control techniques.
-  Our strategy applies to any periodic robotic control task that admits significant prior knowledge.
+  In this paper, we propose a sample-efficient learning-based control technique.
+  Our strategy applies to any periodic \nhatch{why is periodicity necessary?} robotic control task that admits significant prior knowledge.
   First, we engineer a versatile low-level policy (a.k.a. trajectory generator) with interpretable high-level parameters.
@@
   To the best of our knowledge, our experiments are the first to achieve such precise foot placement
-  using a fast, non-static gait that requires no optimization during policy execution.
+  using a model-free, non-static gait that requires no optimization during policy execution.
 \end{abstract}
@@
 \emph{Bipedal} locomotion is more difficult because the support polygon is small and the center of mass is relatively high.
-bipedal locomotion on \emph{rough terrain} is more difficult still, because frequent large obstacles make even small mistakes catastrophic.
+Bipedal locomotion on \emph{rough terrain} is more difficult still, because frequent large obstacles make even small mistakes catastrophic.
 
-One way to handle these challenges is with detailed models and extensive optimization.
+One way to handle these challenges is with detailed dynamics models.
 For example, the WPI-CMU team in the DARPA Robotics Challenge (DRC) controlled ATLAS using a hierarchical optimization approach that carefully controlled the robot's center of mass (CoM) and center of pressure (CoP) \citep{feng2015optimization}.
@@
 Our experiments then demonstrate that these simplifications, taken together, make it feasible to use machine learning for simulated bipedal locomotion over rough terrain.
-In contrast to prior work in footstep planning, our approach learns a dynamic policy that requires no optimization at execution time.
+In contrast to prior work in footstep planning, our approach learns a dynamic, model-free policy that requires no optimization at execution time.
 In contrast to prior work in learning-based locomotion, our learned policy \emph{controls its footstep locations}, which allows it to handle larger and more frequent obstacles.
@@
 
-Many approaches to locomotion on physical robots use optimization-based planning with detailed models.
+Many approaches to locomotion on physical robots use optimization-based planning with detailed dynamics models.
 These approaches tend to generate \emph{hierarchical} plans in order to reduce computational requirements.
 At the highest level, the planner chooses footstep locations using e.g. A* search~\citep{huang2013step}.
-Based on the footstep plan, a lower-level planner generates a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum~\citep{mordatch2010robust} or contact wrench cone~\citep{dai2016planning}.
+Based on the footstep plan, a lower-level planner optimizes a CoM trajectory with a simplified physics model such as a spring-loaded inverted pendulum~\citep{mordatch2010robust} or contact wrench cone~\citep{dai2016planning}.
 Next, inverse kinematics are used to plan trajectories for end-effectors \citep{zucker2010optimization}.
-Finally, these trajectories are optimized using the full dynamics model to account for obstacles and other details~\citep{ratliff2009chomp}.
+Finally, these trajectories are fine-tuned using the full dynamics model to account for obstacles and other details~\citep{ratliff2009chomp}.
 
@@
 \citet{peng2017deeploco} use a hierarchical reinforcement learning algorithm where the upper-level controller chooses footstep locations and the lower-level controller tries to execute those plans.
-However, this strategy produces target step locations that are unrealistic and rarely actually achieved.
+However, in their experiments this strategy produces target step locations that are unrealistic and rarely actually achieved.
 The path that the agent follows is perfectly flat; on rough terrain, such inaccurate step planning would probably fail.
+\nhatch{TODO: expand upon why it probably still wouldn't work even if trained in rough-terrain environments}
 \citet{karpathy2012curriculum} use curriculum learning for a simple planar biped to collect a set of (state, action, resulting step distance) tuples.
@@
 However, they have not been shown to work in situations with dense or large obstacles where small missteps can be catastrophic.
-Such environments require accurate predictive planning at the level of individual contacts.
+Such environments require accurate predictive planning \nhatch{not planning exactly . . . planning implies multi-time-step optimization I think} at the level of individual contacts.
+
+\nhatch{TODO: add a dedicated discussion of MACE, which has the roughest terrain of all of this related work.
+Why can't it handle my version of rough terrain?
+Maybe also directly discuss Heess et al., why it also does not handle rough terrain}
 
@@
 However, one important difference is that in our work, the HLP and the TG run at different frequencies.
+\nhatch{Biped locomotion rough terrain is also, just, a more difficult problem}
 In the next section, we give a detailed description of the differences.
@@
   \item Instead of a 0.3s delay, start the DOWN state when both the swing heel and the CoM are within a certain \emph{distance} of the target heel-strike location.\footnote{
-    If the swing foot hits the ground early, the step immediately ends.}
+    If the swing foot hits the ground early, the step immediately ends.\nhatch{i.e. switch immediately to the next UP state}}
     This distance is proportional to the CoM speed, where the proportionality constant (gain) is controlled by the parameter \texttt{down\_gain}.
@@
 For example, in the locomotion experiments below, the control input is the location of the next footstep target,
-and we use a linear hypothesis class for the HLP.
+and we use a linear \nhatch{TODO} hypothesis class for the HLP.
 
@@
 Due to the unusual policy architecture of CIM, it is unlikely that preexisting experts exist,
-so CIM designs its own.
+so CIM designs its own \nhatch{this makes no sense\dots\emph{I} designed the expert}.
 For general RL problems, this would be difficult because of the infamous \emph{credit assignment problem}:
@@
 \]
-The final term is the \emph{balance penalty}, ensuring that for high-reward actions, the CoM must be approximately ten centimeters behind the heel-strike location.
+The final term is the \emph{balance penalty}, ensuring that for high-reward actions, the CoM must be approximately ten centimeters behind the heel-strike location.\nhatch{explain why there's no $\lambda$ balancing the weights of these terms}
 
-Given a state $\mathbf{s}$, the expert attempts to find a good action $\mathbf{a}$ using random search, specifically Augmented Random Search (ARS) \citep{mania2018simple}.
+Given a state $\mathbf{s}$, the expert attempts to find a good action $\mathbf{a}$ using random search, specifically Augmented Random Search (ARS) \citep{mania2018simple}. \nhatch{TODO update}
 At each iteration, we estimate gradients for the objective $r(\mathbf s, \mathbf a)$ as a weighted average of $Q$ random directions, then take a step in that direction.
@@
 To the best of our knowledge, our experiments are the first to achieve such precise foot placement
-using a fast, non-static gait that requires no optimization during policy execution.
+using a model-free, non-static gait that requires no optimization during policy execution.
 However, one significant limitation of our work so far is that all experiments were performed in simulation.

### 2019-05-30 source/darwin_env.py
Make Darwin start from a slightly crouching pose
In theory, this might make it easier to stay balanced (especially when
transferring to the real robot).

@@ -2,6 +2,7 @@ from simple_3D_env import Simple3DEnv, test_pd_control, test_no_control, setup_d
 import numpy as np
 import consts_darwin
 from IPython import embed
+from hardware.Sim2Robot import raw_darwin_init_state
 
 class DarwinEnv(Simple3DEnv):
     def consts(self):
@@ -44,7 +45,13 @@ class DarwinEnv(Simple3DEnv):
         import curriculum as cur
         cur.SETTINGS_3D_EASY['z_mean'] /= 2
 
+        skel.q = raw_darwin_init_state
+
         return skel
+
+    def reset_world(self):
+        super().reset_world()
+        self.robot_skeleton.q = raw_darwin_init_state
 
 if __name__ == "__main__":
     from pd_control import PDController

### 2019-05-30 source/hardware/Sim2Robot.py

@@ -40,3 +40,8 @@ def toRobot(positions):
 
 	return robotState[index].astype(int).tolist()
 
+# Taken from testPolicyJumper.py (and modified index 1 to make arms more symmetric)
+#init_pos_stand = [1898,2195,1845,2248,2381,1712,2048,2048,2052,2044,1637,2459,2653,1443,2389,1707,2057,2039,2048,2048]
+init_pos_stand = [2048,2048,2048,2048,2048,2048,2048,2048,2052,2044,1637,2459,2653,1443,2389,1707,2057,2039,2048,2048]
+raw_darwin_init_state = np.concatenate((np.zeros(6), fromRobot(init_pos_stand)))
+raw_darwin_init_state[[2,4]] = [-0.025, 0.2] # Make feet parallel to and close to the ground.

### 2019-05-30 source/stepping_stones_env.py

@@ -81,12 +81,15 @@ class SteppingStonesEnv:
         tb.zoom_to = zoom_to
         return tb
 
+    def reset_world(self):
+        self.world.reset() # darwin_env overrides this
+
     def reset(self, state=None, video_save_dir=None, render=None, random=0.0, seed=None):
         if seed is None:
             seed = np.random.randint(100000)
         np.random.seed(seed)
 
-        self.world.reset()
+        self.reset_world()
         self.set_state(state, random)
         if self.video_recorder:
             self.close_video_recorder()

### 2019-06-13 source/simbicon.py
Remove unnecessary method

@@ -72,7 +72,7 @@ class Simbicon(PDController):
                 # For backwards compatibility (if we add new Simbicon params later)
                 raw_gait = np.concatenate((raw_gait, [0.]*(sp.N_PARAMS - len(raw_gait))))
             params += raw_gait * sp.PARAM_SCALE
-        self.set_gait(Params(params))
+        self.params = params
 
         if not c.OBSERVE_TARGET:
             self.unit_normal = np.array([0., 1., 0.])
@@ -134,8 +134,6 @@ class Simbicon(PDController):
         params[sp.STANCE_ANKLE_RELATIVE] += + step_dist_diff * 0.4
         params[sp.TORSO_WORLD]           += - step_dist_diff * 0.5
 
-    def set_gait(self, params):
-        self.params = params
 
     def time(self):
         return self.env.time()

### 2019-06-13 source/consts_2D.py
Add sp.TORSO_ROLL; Adjust target_heading based on sp.STANCE_YAW
Perhaps adjusting the target heading will lead to slightly better
balance and render sp.STANCE_YAW an even more useful param.
(Unfortunately it was still leading to some twisting last time I tried
it with the full learning algorithm.)

sp.TORSO_ROLL was an attempt to allow lateral adjustments without
directly messing with hip angles required for balance. But based on a
test in step_learner.py, it didn't work well at all. But I'll leave it
in for now, unused.

Also, this commit removes the effect sp.HEADING (now using sp.STANCE_YAW

@@ -101,4 +101,4 @@ def foot_transform_from_angles(_, pitch, __):
 BASE_GAIT = np.array([0.14, 0, 0.2, 0.0, 0.2,
                       0.4, -1.1,   0.0, -0.05,
                       0,    0, 0.05, -0.1,
-                      0,0,0,0,0,0,0]) # None of these last 7 are used in 2D
+                      0,0,0,0,0,0,0,0,0]) # None of these are used in 2D

### 2019-06-13 source/consts_armless.py

@@ -108,4 +108,4 @@ BASE_GAIT = np.array([0.14, 0.5, 0.2, -0.1, 0.2,
                       -0.0, -0.00, 0.1, -0.1,
                       0.5, 0.2, 0.0, 0.0, 0.0, 0.0,
                       0.0, # This one is only used for Darwin
-                      0.0])
+                      0.0, 0.0])

### 2019-06-13 source/consts_darwin.py

@@ -148,7 +148,7 @@ BASE_GAIT = np.array([0.06, 0.5, 0.2, 0.03, -0.01,
                       -0.0, -0.2,  0.3, -0.1,
                       0.5, 0.2, 0.0, 0.0, 0.0, 0.0,
                       0.1,
-                      0.0])
+                      0.0, 0.0])
 
 # Copied limits from
 # http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.822.6324&rep=rep1&type=pdf

### 2019-06-13 source/simbicon.py

@@ -84,20 +84,9 @@ class Simbicon(PDController):
 
         if target_heading is None:
             branch = self.heading()
-            _, self.target_heading = heading_from_vector(d, branch)
-            if self.env.is_3D:
-                # Lateral balance feedback usually tends to put the swing foot too far "outside"
-                # the target. Hack: adjust the target heading to roughly compensate for this.
-                dist = np.linalg.norm(d)
-                adj = -dist*0.2
-                if self.swing_idx == c.LEFT_IDX:
-                    self.target_heading -= adj
-                else:
-                    self.target_heading += adj
+            self.target_heading = branch - self.params[sp.STANCE_YAW]
         else:
             self.target_heading = target_heading
-        if self.env.is_3D:
-            self.target_heading += self.params[sp.HEADING]
         self.target_direction = np.array(
                 [np.cos(self.target_heading), 0, -np.sin(self.target_heading)])
 

### 2019-06-13 source/simbicon_3D.py

@@ -7,14 +7,14 @@ from sdf_loader import RED, GREEN, BLUE
 import utils
 
 from consts_common3D import *
-from simbicon_params import *
+import simbicon_params as sp
 
 ZERO_GAIN = False
 
 class Simbicon3D(Simbicon):
     def set_gait_raw(self, target, target_heading=None, raw_gait=None):
         if raw_gait is not None and self.swing_idx == LEFT_IDX:
-            raw_gait = raw_gait * MIRROR_PARAMS
+            raw_gait = raw_gait * sp.MIRROR_PARAMS
         return super().set_gait_raw(target, target_heading, raw_gait)
 
     def heading(self):
@@ -42,8 +42,8 @@ class Simbicon3D(Simbicon):
         c = self.env.consts()
         params = self.params
 
-        cd = params[POSITION_BALANCE_GAIN_LAT]
-        cv = params[VELOCITY_BALANCE_GAIN_LAT]
+        cd = params[sp.POSITION_BALANCE_GAIN_LAT]
+        cv = params[sp.VELOCITY_BALANCE_GAIN_LAT]
 
         if hasattr(self.env, 'sdf_loader'):
             proj = q[:3]
@@ -59,10 +59,10 @@ class Simbicon3D(Simbicon):
         balance_feedback = -(cd*d[Z] + cv*v[Z])
 
         tq[self.swing_idx+HIP_ROLL] += balance_feedback - q[ROOT_ROLL]
-        tq[self.stance_idx+ANKLE_ROLL] += params[STANCE_ANKLE_ROLL]
-        tq[self.swing_idx+ANKLE_ROLL] += params[SWING_ANKLE_ROLL]
+        tq[self.stance_idx+ANKLE_ROLL] += params[sp.STANCE_ANKLE_ROLL]
+        tq[self.swing_idx+ANKLE_ROLL] += params[sp.SWING_ANKLE_ROLL]
 
-        tq[TORSO_ROLL] = -q[ROOT_ROLL]
+        tq[TORSO_ROLL] = -q[ROOT_ROLL] + params[sp.TORSO_ROLL]
 
         return tq
 
@@ -76,7 +76,7 @@ def next_target(start, heading, length, env):
     target = start + np.dot(rot, offset)
     return target
 
-def test(env, length, r=1, n=8, a=0.0, delta_a=0.0, relative=False, provide_target_heading=True, mirror=False):
[…]

### 2019-06-13 source/simbicon_params.py

@@ -28,9 +28,10 @@ UP_DURATION = 19
 
 # Add experimental new params at the bottom
 STANCE_YAW = 20
+TORSO_ROLL = 21
 
 import numpy as np
-MIRROR_PARAMS = np.array([1,1,1,1,1, 1,1,1,1, 1,1,1,1, 1,1,-1,-1,-1,-1, 1,-1])
-PARAM_SCALE   = np.array([1,1,1,1,3, 1,1,1,2, 2,2,3,2, 1,1, 1, 1, 3, 3, 1, 1])
+MIRROR_PARAMS = np.array([1,1,1,1,1, 1,1,1,1, 1,1,1,1, 1,1,-1,-1,-1,-1, 1,-1,-1])
+PARAM_SCALE   = np.array([1,1,1,1,3, 1,1,1,2, 2,2,3,2, 1,1, 1, 1, 3, 3, 1, 1, 1])
 
 N_PARAMS = len(PARAM_SCALE)

### 2019-06-13 source/test_inverse_dynamics.py
Fix up in test_inverse_dynamics
There was a bug with not adding all the stepping stones, and the test
took too long with the new tolerance settings, so I reduced the number
of steps per episode.

@@ -8,6 +8,9 @@ import curriculum as cur
 TRAIN_SETTINGS_3D_TEST = {**cur.TRAIN_SETTINGS_3D,
     'n_trajectories': 1,
     }
+EVAL_SETTINGS_3D_TEST = {**cur.SETTINGS_3D_EASY,
+    'n_steps': 4,
+    }
 
 def retrieve_index(learn, i=None):
     learn.env.clear_skeletons()
@@ -27,8 +30,7 @@ def demo_train_set(learn, render=1.0, indices=None):
     ii = indices or range(len(learn.train_responses))
     for i in ii:
         start_state, target, response, features = retrieve_index(learn, i)
-        uss = learn.evaluator.eval_settings['use_stepping_stones']
-        runner = Runner(learn.env, start_state, target, use_stepping_stones=uss)
+        runner = Runner(learn.env, start_state, target)
         runner.reset(render=render)
         reward = runner.run(response)
         print("Index {} Score: {}".format(i, reward)) # Should be pretty close to 1
@@ -56,7 +58,7 @@ def test_regression_bias(learn, i=None):
 
 def test_mirroring(learn, i=3):
     start_state, target, response, features = retrieve_index(learn, i)
-    runner = Runner(learn.env, start_state, target, use_stepping_stones=False)
+    runner = Runner(learn.env, start_state, target)
     trained_response = learn.act(features)
 
     # Note this will only give accurate results if `learn` has been trained
@@ -64,7 +66,7 @@ def test_mirroring(learn, i=3):
     mirrored_start = start_state.copy()
     mirrored_start.mirror()
     mirrored_target = target*[1,1,-1]
-    mirrored_runner = Runner(learn.env, mirrored_start, mirrored_target, use_stepping_stones=False)
+    mirrored_runner = Runner(learn.env, mirrored_start, mirrored_target)
     mirrored_features = mirrored_start.extract_features(mirrored_target)
     mirrored_response = learn.act(mirrored_features)
 
@@ -96,7 +98,7 @@ if __name__ == '__main__':
     learn = LearnInverseDynamics(env, name)
     learn.set_train_settings(TRAIN_SETTINGS_3D_TEST)
     #learn.load_train_set()
-    learn.evaluator.set_eval_settings(cur.SETTINGS_3D_EASY)
+    learn.evaluator.set_eval_settings(EVAL_SETTINGS_3D_TEST)
     #print('Score:', learn.evaluate()['total_score'])
     for i in range(2):
         learn.training_iter()

### 2019-06-13 source/evaluator.py
Remove ability to specify target heading
This opens the possibility of standardizing observations by yaw again
(in state.py), which might help with generalization.

@@ -42,7 +42,7 @@ class Evaluator:
             features = state.extract_features(target)
             action = policy(features)
             raw_pose_start = self.env.robot_skeleton.x
-            end_state, terminated = self.env.simulate(target, target_heading=0.0, action=action)
+            end_state, terminated = self.env.simulate(target, action=action)
             error = np.linalg.norm(end_state.stance_heel_location() - target)
             reward = utils.reward(self.env.controller, end_state)
             total_reward += reward

### 2019-06-13 source/learn_llc.py

@@ -29,7 +29,7 @@ def test(env, length, param_setting, render=None, n=50, terminate_on_slip=True):
     c = env.controller
     penalty = 0.0
     for i in range(n):
-        obs, terminated, n_float = env.simulate([i*length, GL, 0], target_heading=0.0,
+        obs, terminated, n_float = env.simulate([i*length, GL, 0],
                 action=param_setting,
                 put_dots=True, count_float=True)
         if terminated:

### 2019-06-13 source/real_darwin_env.py

@@ -44,7 +44,7 @@ class RealDarwinEnv:
         c = self.consts()
         if self.controller.swing_contact(None, None): # TODO rename: not contact, just timeout
             status_string = self.controller.change_stance(np.zeros(3))
-            self.controller.set_gait_raw(raw_gait=EMBED_B5, target_heading=None, target=None)
+            self.controller.set_gait_raw(None, raw_gait=EMBED_B5)
             print(status_string)
         q_raw, dq_raw = self.robot.read(self.prev_control_time, t)
         # Manually doing this instead of standardized_dofs to avoid the
@@ -75,7 +75,7 @@ class RealDarwinEnv:
 if __name__ == '__main__':
     env = RealDarwinEnv()
     duration = 2 # seconds
-    env.controller.set_gait_raw(raw_gait=EMBED_B5, target_heading=None, target=None)
+    env.controller.set_gait_raw(None, raw_gait=EMBED_B5)
     try:
         while env.time() < duration:
             env.tick()

### 2019-06-13 source/simbicon.py

@@ -64,7 +64,7 @@ class Simbicon(PDController):
     def heading(self):
         return 0.0
 
-    def set_gait_raw(self, target, target_heading=None, raw_gait=None):
+    def set_gait_raw(self, target, raw_gait=None):
         c = self.env.consts()
         params = c.BASE_GAIT.copy()
         if raw_gait is not None:
@@ -82,11 +82,8 @@ class Simbicon(PDController):
         self.target = target
         d = self.target - swing_heel
 
-        if target_heading is None:
-            branch = self.heading()
-            self.target_heading = branch - self.params[sp.STANCE_YAW]
-        else:
-            self.target_heading = target_heading
+        branch = self.heading()
+        self.target_heading = branch - self.params[sp.STANCE_YAW]
         self.target_direction = np.array(
                 [np.cos(self.target_heading), 0, -np.sin(self.target_heading)])
 

### 2019-06-13 source/simbicon_3D.py

@@ -12,10 +12,10 @@ import simbicon_params as sp
 ZERO_GAIN = False
 
 class Simbicon3D(Simbicon):
-    def set_gait_raw(self, target, target_heading=None, raw_gait=None):
+    def set_gait_raw(self, target, raw_gait=None):
         if raw_gait is not None and self.swing_idx == LEFT_IDX:
             raw_gait = raw_gait * sp.MIRROR_PARAMS
-        return super().set_gait_raw(target, target_heading, raw_gait)
+        return super().set_gait_raw(target, raw_gait)
 
     def heading(self):
         return self.env.get_x()[0][ROOT_YAW]

### 2019-06-13 source/step_learner.py

@@ -20,7 +20,7 @@ class Runner:
 
     def run(self, action):
         self.n_runs += 1
-        r, _ = self.env.simulate(self.target, target_heading=0.0, action=action, put_dots=True)
+        r, _ = self.env.simulate(self.target, action=action, put_dots=True)
         if self.env.video_recorder is not None:
             self.env.pause(0.3)
         return utils.reward(self.env.controller, r)
@@ -42,7 +42,7 @@ def collect_start_state(env, targets, video_save_dir, use_stepping_stones=True):
     env.render()
     steps = targets[2:-1] if env.is_3D else targets[1:-1] # TODO refactor so 2D also uses Evaluator
     for t in steps:
-        end_state, _ = env.simulate(t, target_heading=0.0, put_dots=True)
+        end_state, _ = env.simulate(t, put_dots=True)
         env.render()
     env.clear_skeletons()
     return end_state
@@ -77,12 +77,12 @@ def test_3D(video_save_dir):
     opzer = cma_wrapper.CMAWrapper()
     GY = env.consts().GROUND_LEVEL
     learn_last_move(env, opzer, BASIC_3D[:-1], video_save_dir=video_save_dir)
-    env.simulate(BASIC_3D[-1], 0.0)
+    env.simulate(BASIC_3D[-1])
     #env.clear_skeletons()
     # LONG_STEP is a little too hard, but BASIC at least should be learnable
     LONG_STEP_3D = np.array([[0, GY, 0], [0.3, GY, 0.1], [0.6, GY, -0.1], [1.4, GY, 0.1]])
     #learn_last_move(env, opzer, LONG_STEP_3D, video_save_dir=video_save_dir)
-    #env.simulate([1.7, -.9, -.1], 0.0)
+    #env.simulate([1.7, -.9, -.1])
 
 if __name__ == '__main__':
     #test_2D()

### 2019-06-13 source/stepping_stones_env.py

@@ -158,9 +158,9 @@ class SteppingStonesEnv:
             return None, False, stance_contact
 
     # Run one footstep of simulation, returning the final state
-    def simulate(self, target, target_heading=None, action=None, put_dots=False,
+    def simulate(self, target, action=None, put_dots=False,
             count_float=False):
-        self.controller.set_gait_raw(raw_gait=action, target_heading=target_heading, target=target)
+        self.controller.set_gait_raw(target, raw_gait=action)
         steps_per_render = None
         if self.render_rate:
             steps_per_render = int(self.consts().REAL_TIME_STEPS_PER_RENDER / self.render_rate)

### 2019-06-13 source/curriculum.py
Try some more tweaks
- Make swing knee controllable
- Don't penalize extreme headings
- Standardize yaw angles again

This is a follow-up to the previous commit. We'll see how much the
changes it permits actually help the learning algorithm.

@@ -103,6 +103,7 @@ TRAIN_SETTINGS_3D = {
             #sp.POSITION_BALANCE_GAIN_LAT,
             sp.UP_IDX+sp.SWING_HIP_WORLD,
             sp.STANCE_YAW,
+            sp.UP_IDX+sp.SWING_KNEE_RELATIVE,
             ]),
     'observable_features': o3d,
     'model_class': 'linear',

### 2019-06-13 source/state.py

@@ -92,10 +92,10 @@ class State:
         # relative to the direction the robot's actually going). Until I can remove
         # the dependence of Simbicon3D on knowing the target direction, I'm just going
         # to keep that angle as part of the state (by removing these lines of code).
-        #if self.consts.BRICK_DOF == 6:
-        #    angle = copy.raw_state[self.consts.ROOT_YAW]
-        #    rot = copy.rotate(-angle)
-        #    target[:3] = np.dot(rot, target[:3])
+        if self.consts.BRICK_DOF == 6:
+            angle = copy.raw_state[self.consts.ROOT_YAW]
+            rot = copy.rotate(-angle)
+            target[:3] = np.dot(rot, target[:3])
 
         return np.concatenate([copy.raw_state, target])
 

### 2019-06-13 source/test_inverse_dynamics.py

@@ -56,7 +56,7 @@ def test_regression_bias(learn, i=None):
     # This score should also be close to zero in the absence of model bias.
     print("Score:", runner.run(trained_response))
 
-def test_mirroring(learn, i=3):
+def test_mirroring(learn, i=2):
     start_state, target, response, features = retrieve_index(learn, i)
     runner = Runner(learn.env, start_state, target)
     trained_response = learn.act(features)

### 2019-06-13 source/utils.py

@@ -34,10 +34,10 @@ def reward(controller, state):
         score -= (0.03-d)
     if d > 0.2: # COM is far behind the heel
         score -= (d-0.2)
-    if state.consts.BRICK_DOF == 6:
-        slip_angle = np.abs(state.pose()[state.consts.ROOT_YAW])
-        if slip_angle > np.pi/20:
-            score -= (slip_angle-np.pi/20) # Hinge loss
+    #if state.consts.BRICK_DOF == 6:
+    #    slip_angle = np.abs(state.pose()[state.consts.ROOT_YAW])
+    #    if slip_angle > np.pi/20:
+    #        score -= (slip_angle-np.pi/20) # Hinge loss
     return score
 
 def rotmatrix(theta):

### 2019-06-13 paper/CORL2019/main.tex
Add paper draft in CORL format

--- 2019-05-30 paper/rss/main.tex
@@
-\documentclass[conference]{IEEEtran}
-\usepackage{times}
+\documentclass{article}
+\newcommand{\boots}[1]{{\textcolor{red} {Byron: {#1}}}}
+\newcommand{\nhatch}[1]{{\leavevmode\color{magenta} Nathan: #1}}
+\newcommand{\norm}[1]{\lVert #1 \rVert}
 
-% numbers option provides compact numerical references in the text. 
-\usepackage[numbers]{natbib}
-\usepackage{multicol}
-\usepackage[bookmarks=true]{hyperref}
+\usepackage{corl_2019} % initial submission
+%\usepackage[final]{corl_2019} % Uncomment for the camera-ready ``final'' version
+
 \usepackage{xcolor}
@@
 \usepackage{amsmath}
-\newcommand{\boots}[1]{{\textcolor{red} {Byron: {#1}}}}
-
-\pdfinfo{
-   /Author (Homer Simpson)
-   /Title  (Robots: Our new overlords)
-   /CreationDate (D:20101201120000)
-   /Subject (Robots)
-   /Keywords (Robots;Overlords)
-}
-\newcommand{\nhatch}[1]{{\leavevmode\color{magenta} Nathan: #1}}
-\newcommand{\norm}[1]{\lVert #1 \rVert}
-
-\begin{document}
+%\usepackage{times} %Not sure what this does
 
@@
 
-%\author{\authorblockN{Michael Shell}
-%\authorblockA{School of Electrical and\\Computer Engineering\\
-%Georgia Institute of Technology\\
-%Atlanta, Georgia 30332--0250\\
-%Email: mshell@ece.gatech.edu}
-%\and
-%\authorblockN{Homer Simpson}
-%\authorblockA{Twentieth Century Fox\\
-%Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\and
-%\authorblockN{James Kirk\\ and Montgomery Scott}
-%\authorblockA{Starfleet Academy\\
-%San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212\\
-%Fax: (888) 555--1212}}
+% The \author macro works with any number of authors. There are two
+% commands used to separate the names and addresses of multiple
+% authors: \And and \AND.
+%
+% Using \And between authors leaves it to LaTeX to determine where to
+% break the lines. Using \AND forces a line break at that point. So,
+% if LaTeX puts 3 of 4 authors names on the first line, and the last
+% on the second line, try using \AND instead of \And before the third
+% author name.
 
-% avoiding spaces at the end of the author lines is not a problem with
-% conference papers because we don't use \thanks or \IEEEmembership
+% NOTE: authors will be visible only in the camera-ready (ie, when using the option 'final'). 
+% 	For the initial submission the authors will be anonymized.
 
-% for over three affiliations, or if they all won't fit within the width
-% of the page, use this alternative format:
-% 
-%\author{\authorblockN{Michael Shell\authorrefmark{1},
-%Homer Simpson\authorrefmark{2},
-%James Kirk\authorrefmark{3}, 
-%Montgomery Scott\authorrefmark{3} and
-%Eldon Tyrell\authorrefmark{4}}
-%\authorblockA{\authorrefmark{1}School of Electrical and Computer Engineering\\
-%Georgia Institute of Technology,
-%Atlanta, Georgia 30332--0250\\ Email: mshell@ece.gatech.edu}
-%\authorblockA{\authorrefmark{2}Twentieth Century Fox, Springfield, USA\\
-%Email: homer@thesimpsons.com}
-%\authorblockA{\authorrefmark{3}Starfleet Academy, San Francisco, California 96678-2391\\
-%Telephone: (800) 555--1212, Fax: (888) 555--1212}
-%\authorblockA{\authorrefmark{4}Tyrell Inc., 123 Replicant Street, Los Angeles, California 90210--4321}}
+\author{
+  David S.~Hippocampus\\
+  Department of Electrical Engineering and Computer Sciences\\
+  University of California Berkeley 
+  United States\\
+  \texttt{hippo@berkeley.edu} \\
+  %% examples of more authors
+  %% \And
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+  %% \AND
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+  %% \And
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+  %% \And
+  %% Coauthor \\
+  %% Affiliation \\
+  %% Address \\
+  %% \texttt{email} \\
+}
 
+\begin{document}
 \maketitle
+
+%===============================================================================
 
@@
 
-\IEEEpeerreviewmaketitle
+% Two or three meaningful keywords should be added here
+\keywords{CORL, Robots, Learning}
+
+%===============================================================================
 
@@
 
-\bibliographystyle{plainnat}
+% The maximum paper length is 8 pages excluding references and acknowledgements, and 10 pages including references and acknowledgements
+
+\clearpage
+% The acknowledgments are automatically included only in the final version of the paper.
+\acknowledgments{If a paper is accepted, the final camera-ready version will (and probably should) include acknowledgments.
+All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support.}
+
+%===============================================================================
+
+% no \bibliographystyle is required, since the corl style is automatically used.
+
 \bibliography{../bib}
@@
 \label{app:simbicon}
+
+\nhatch{TODO: extract all appendices into a supplementary material .zip file.}
 

### 2019-06-13 source/simbicon.py
Revert to code state used for test_cim_3D_9
This is the best-performing controller so far, but it depends on certain
code features that aren't captured in `history` (for example, it
requires a constant `target_heading` of zero). This commit memorializes
those settings (assuming I remember them correctly). I tested the
controller with these settings and it seems to perform comparably to
what it did during training, which indicates the settings are correct.

@@ -84,6 +84,7 @@ class Simbicon(PDController):
 
         branch = self.heading()
         self.target_heading = branch - self.params[sp.STANCE_YAW]
+        self.target_heading = 0.0
         self.target_direction = np.array(
                 [np.cos(self.target_heading), 0, -np.sin(self.target_heading)])
 

### 2019-06-13 source/state.py

@@ -92,10 +92,10 @@ class State:
         # relative to the direction the robot's actually going). Until I can remove
         # the dependence of Simbicon3D on knowing the target direction, I'm just going
         # to keep that angle as part of the state (by removing these lines of code).
-        if self.consts.BRICK_DOF == 6:
-            angle = copy.raw_state[self.consts.ROOT_YAW]
-            rot = copy.rotate(-angle)
-            target[:3] = np.dot(rot, target[:3])
+        #if self.consts.BRICK_DOF == 6:
+        #    angle = copy.raw_state[self.consts.ROOT_YAW]
+        #    rot = copy.rotate(-angle)
+        #    target[:3] = np.dot(rot, target[:3])
 
         return np.concatenate([copy.raw_state, target])
 

### 2019-06-13 source/utils.py

@@ -34,10 +34,10 @@ def reward(controller, state):
         score -= (0.03-d)
     if d > 0.2: # COM is far behind the heel
         score -= (d-0.2)
-    #if state.consts.BRICK_DOF == 6:
-    #    slip_angle = np.abs(state.pose()[state.consts.ROOT_YAW])
-    #    if slip_angle > np.pi/20:
-    #        score -= (slip_angle-np.pi/20) # Hinge loss
+    if state.consts.BRICK_DOF == 6:
+        slip_angle = np.abs(state.pose()[state.consts.ROOT_YAW])
+        if slip_angle > np.pi/20:
+            score -= (slip_angle-np.pi/20) # Hinge loss
     return score
 
 def rotmatrix(theta):
